---
title: "10114 - Create datasets for analysis using *mrmr*"
author: "thomascsmith"
output: html_notebook
editor_options: 
  chunk_output_type: inline
---

# Create CMR data for Goddard 10114
This was updated from updated code for 10109, in August 2026. For these neighboring sites, the reintroduction dates should be identical and primary periods should be identical; there are small differences in the secondary periods. 

This notebook creates the survey, capture, and translocation files for use in analyzing capture-mark-recapture data using the `mrmr` package. The code retrieves data from the PostgreSQL amphibians database, runs error checks, and creates the relevant files in `mrmr` format. The code is written in distinct chunks to allow review of results from one chunk before running the next. For example, review of the raw data files is necessary to identify potentially problematic records for review and possible removal from the data set.

Before using the code in this file, replace all instances of "xxxx" with the 5-digit site_id of interest, i.e. **10114**. 

Users without credentials to access the database can use the saved CSV files created in the save-raw-data-files code chunk.

Note for credentialed users: Data from the most recent field season may not yet be appended to tables in the amphibians database.
If these recently-collected data are necessary (e.g., for preliminary analyses), complete datasets can be created by copying code chunks in "fulcrum_data_append.Rmd" into this notebook and executing all code chunks.

## Load packages

```{r load-packages}
library(tidyverse)
library(RPostgreSQL)
library(lubridate)
library(assertthat)
# options(readr.show_col_types = FALSE)
```
--- 

## Create current-year survey, capture, and translocation/reintroduction datasets for appending
- Copied the following two code chunks into the "10114_createdata_mrmr.Rmd" script.  
- relies on the manual download/filtering of current year [2026] Fulcrum data (as desribed in the 'fulcrum_data_append.Rmd' file)

```{r create_current_year_data_objects}
visit_10114_current <- read_csv(here::here("data", "raw", "current_year", "survey_10114.csv")) %>% 
  select(record_id, site_id, visit_date, visit_status, visit_comment) %>% 
  rename(comment_visit = visit_comment)

survey_10114_current <- read_csv(here::here("data", "raw", "current_year", "survey_description_10114.csv")) %>% 
  select(record_id, survey_type, survey_description, wind, sky, air_temp, duration, survey_quality, general_survey_comment) %>% 
  filter(survey_type == "cmr") %>% 
  rename(description = survey_description, comment_survey = general_survey_comment)

amphibian_capture_10114_current <- read_csv(here::here("data", "raw", "current_year", "survey_amphibian_capture_10114.csv")) %>% 
  filter(method == "cmr") %>% 
  rename(comment_capture = capture_comment) %>% 
  mutate(pit_tag_ref = as.character(pit_tag_ref)) %>% 
  replace_na(list(species = "ramu", capture_life_stage = "adult")) %>% 
  select(record_id, pit_tag_ref, tag_new, species, capture_life_stage, capture_animal_state, length, comment_capture)

relocate_10114_current <- read_csv(here::here("data", "raw", "current_year", "relocate_10114.csv")) %>% 
  filter(relocate_type == "release") %>% #
  rename(comment_release = release_comments,
         type = relocate_type) %>% 
  select(record_id, release_site_id, type, date, comment_release)

relocate_amphibians_10114_current <- read_csv(here::here("data", "raw", "current_year", "relocate_relocated_individual_10114.csv")) %>% 
  mutate(pit_tag_ref = as.character(pit_tag_ref)) %>% 
  # rename(relocate_amphibians_comments = comments) %>% 
  select(record_id, pit_tag_id=pit_tag_ref, tag_new, life_stage, sex, length, weight, swab_id, comments)
```
```{r join-current-year-tables-together-to-make-appendable-objects}
survey_append_10114 <- visit_10114_current %>% 
  inner_join(survey_10114_current, by = c("record_id")) %>% 
  select(-record_id) |> 
  mutate(visit_date=mdy(visit_date))

capture_append_10114 <- visit_10114_current %>% 
  inner_join(survey_10114_current, by = c("record_id")) %>% 
  inner_join(amphibian_capture_10114_current, by = c("record_id")) %>% 
  select(-record_id) |> 
  mutate(visit_date=mdy(visit_date))

reintroduction_append_10114 <- relocate_10114_current %>% 
  inner_join(relocate_amphibians_10114_current, by = c("record_id")) %>% 
  rename(site_id = release_site_id, 
         capture_life_stage = life_stage) %>% 
  mutate(date = mdy(date)) |> 
  select(-record_id)
```

### Clean up environment

```{r}
# rm(amphibian_capture_10114, relocate_10114, relocate_amphibians_10114, survey_10114, visit_10114)
```


---

## Retrieve data from PostgreSQL database

### Connect to database

- When connecting via SSH, open external connection to database, then replace "Sys.getenv" with "rstudioapi::askForPassword" for host and port, run chunk.   
- Users without credentials can use CSV files created in code chunk "save-raw-data-files" below as a starting point.  

```{r db-connect-set-schemas}
con = dbConnect(dbDriver("PostgreSQL"),
                host = Sys.getenv("host"), 
                port = Sys.getenv("port"), 
                dbname = Sys.getenv("dbname"),
                user = rstudioapi::askForPassword("user name"),
                password = rstudioapi::askForPassword("password")
                # options = "-c search_path=field_surveys" # add this to access the schema for all tables; versus setting schema as each table is accessed and assigned
                )
DBI::dbExecute(con, "SET search_path TO field_surveys, qpcr") # now the schemas are set, in order of query priority
DBI::dbListTables(con) # to test the connection and schemas

```

### Retrieve translocation/reintroduction data from database

- Edit "where" statement to include relevant site_ids and types.  

```{sql retrieve_reintro_data, connection=con, output.var = "reintroduction_10114"}
select 
  visit.site_id, 
  release.release_type AS type,
  visit.visit_date AS date,
  release_frog.pit_tag_ref AS pit_tag_id, 
  release_frog.tag_new, 
  release_frog.sex, 
  release_frog.length, 
  release_frog.weight, 
  release_frog.swab_id, 
  release_frog.bd_exposure
from release_frog
INNER JOIN release ON release_frog.release_id = release.release_id
INNER JOIN visit ON release.visit_id = visit.visit_id
where visit.site_id in (10114, 11886, 11889) 
```

### Retrieve survey data from database

- Edit "where" statement to include relevant site_ids and types.
- If some relevant surveys are classified as "ves", include in "where" statement. 
```{sql retrieve-survey-data, connection=con, output.var = "survey_10114"}
select 
  site_id, 
  visit_date as date, 
  visit_status, 
  survey_type, 
  description, 
  survey_quality,
  wind,
  sky,
  air_temp,
  duration,
  visit_comment,
  survey_comment
from visit
inner join survey on visit.visit_id = survey.visit_id
where site_id in (10114, 11886, 11889) AND 
      survey_type in ('cmr', 'visual') AND
      visit_date > '2020-01-01'
```
```{r}
view(survey_10114)
```

### Retrieve frog capture data from database

- Edit "where" statement to include relevant site_ids and types. 
```{sql retrieve-capture-data, connection=con, output.var = "capture_10114"}
select 
  visit.site_id, 
  visit.visit_date as date, 
  visit.visit_status, 
  survey.survey_type, 
  capture_survey.pit_tag_ref AS pit_tag_id, 
  capture_survey.tag_new, 
  capture_survey.species, 
  capture_survey.capture_life_stage, 
  capture_survey.capture_animal_state, 
  capture_survey.length,
  capture_survey.weight,
  visit.visit_comment,
  survey.survey_comment,
  capture_survey.capture_survey_comment,
  visit.visit_id,
  survey.survey_id,
  capture_survey.capture_survey_id,
  capture_survey.swab_id,
  bd_load.bd_load
from visit
inner join survey on visit.visit_id = survey.visit_id
inner join capture_survey on survey.survey_id = capture_survey.survey_id
left join bd_load on bd_load.sample_id = capture_survey.swab_id
where site_id in (10114, 11886, 11889) and
      survey_type in ('cmr')
```
```{r}
view(capture_10114)
```

```{r db-disconnect3}
dbDisconnect(con)
rm(con)
```

### append current year data

### Review data
- Review retrieved datasets before using in subsequent chunks.  

### Save raw data files
```{r save-raw-data-files, message=FALSE}
reintroduction_10114 %>% write_csv(here::here("data", "raw", "10114_reintroduction_raw.csv")) # if relevant
survey_10114 %>% write_csv(here::here("data", "raw", "10114_survey_raw.csv"))
capture_10114 %>% write_csv(here::here("data", "raw", "10114_capture_raw.csv"))
```

- Read saved files with (for example): 
```{r reread-saved-raw-data}
# reintroduction_10114 <- read_csv(here::here("data", "raw", "10114_reintroduction_raw.csv")) %>% 
#  mutate(pit_tag_id = as.character(pit_tag_id), 
#         type="release") |> 
#  rename(date=release_date)
```
## Create translocation/reintroduction dataset

- Skip if translocations/reintroductions are not relevant.  

### Retain only relevant columns

```{r reintroduction-relevant-columns}
reintroduction_10114 <- reintroduction_10114 %>% 
  select(site_id, date, type, pit_tag_id, tag_new, sex, length, weight, swab_id) 
```
## Append current-year translocation records

-  If translocations are relevant, copy the following code chunk and insert immediately following the "translocation-relevant-columns" code chunk in "xxxxx_createdata_mrmr.Rmd".   

```{r append_currentyear_translocation_records}
reintroduction_10114 <- reintroduction_append_10114 %>% 
  # filter(type == "release") %>% # options are "translocation" or "reintroduction"
  select(site_id, date, type, pit_tag_id, tag_new, sex, length, weight, swab_id) %>% 
  bind_rows(reintroduction_10114) |> 
  unique()
```

### Reintroduction data checks

- Run checks before and after data cleaning (next code chunk) to ensure that all issues have been resolved.  
- Run lines of code one at a time to allow easy evaluation of messages and outputs.  

```{r reintroduction-data-checks, message=FALSE}
# Check for records where pit_tag_ref is NULL
assert_that(noNA(reintroduction_10114$pit_tag_id))

# Check for records containing pit tags shorter than 15 characters
assert_that(!any(nchar(reintroduction_10114$pit_tag_id) != 15))

# Check for records containing pit tags that end with ".1" (known to occur incorrectly throughout the MLRG database) 
assert_that(!any(str_detect(reintroduction_10114$pit_tag_id, "\\.1$")))

# Check for records of subadult frogs
reintroduction_10114 %>% drop_na(length) %>% filter(length < 40)
### produces one frog measured 39 mm in 2023: 900043000216256

# Check for duplicate pit tags on each date
reintroduction_10114 %>% count(date, pit_tag_id) %>% filter(n > 1)

# Are release dates correct?
reintroduction_10114 %>% distinct(date) %>% glimpse()
```

### Clean reintroduction data

- Edit data-cleaning code as necessary.  

```{r clean-reintroduction-data}
reintroduction_10114 <- reintroduction_10114 %>% 
  filter(!is.na(pit_tag_id))
```

### Create final reintroduction dataset formatted for mrmr

```{r final-reintroduction-data}
reintroduction_10114 <- reintroduction_10114 %>% 
  # rename(pit_tag_id = pit_tag_ref) %>% 
  arrange(date, pit_tag_id) %>% 
  select(site_id, type, date, pit_tag_id) |> 
  rename(release_date = date) |> 
  mutate(type = case_when(type == "zoo_reintroduction" ~ "release", .default = type), 
         release_date = case_when( release_date == as.Date("2026-07-17") ~ as.Date("2026-07-18"), .default = release_date))

```

### Save cleaned translocation data

```{r save-cleaned-translocation-data, message=FALSE}
reintroduction_10114 %>% write_csv(here::here("data", "clean", "10114_reintroduction.csv"))
```

## Create survey dataset

### Include only relevant surveys

Multiple numbered sites exist in the 10114 survey area, so we include only one survey for entire site on each survey date. Other sites surveyed and occupied as of August 2026 include:
11886 
11889

#### Primary site surveyed during each secondary period?

- If yes, code chunk will return zero records. Exclude other records as necessary. 
- In filter statement, edit site_id as necessary.

### Append current-year survey records
- Copy the following code chunk and insert immediately before the "check-survey-history" code chunk in "10114_createdata_mrmr.Rmd".  

```{r append_currentyear_survey_records}
survey_10114_build <- survey_append_10114 |> 
  rename(date=visit_date) |> 
  bind_rows(survey_10114) %>% 
  arrange(date, desc(site_id)) %>% 
  select(site_id, date, visit_status, survey_type, description, survey_quality, wind, sky, air_temp, duration, comment_visit, comment_survey)
```

```{r check-survey-history}
survey_10114_build %>% 
  add_count(date) %>% 
  filter(n == 1 & site_id != "10114")
```

#### Include only surveys at primary site

- Run this code chunk as is when only CMR surveys need to be included. If VES surveys also need to be included, modify code accordingly. 

```{r include-relevant-surveys}
survey_10114_build <- survey_10114_build %>% 
  filter(site_id %in% c( 10114 ) ) %>%
  # rename(date = visit_date) %>% 
  arrange(date)|> 
  filter(survey_type=="cmr")
```

### Add translocation/reintroduction dates as surveys

For the July 2026 survey, some CMR work occurred on the same day as the reintroduction, therefore, the reintroduction date is revised to be one day later so that both do not occur concurrently.

```{r add-reintroduction-dates-as-surveys}
survey_10114_build <- reintroduction_10114 %>% 
  rename(survey_type=type) |> 
  distinct(site_id, date, survey_type) %>% 
  select(site_id, date, survey_type) |> 
  bind_rows(survey_10114_build |> 
              select(site_id, date, survey_type) 
            ) %>% 
  mutate(date = case_when(date==as.Date("2026-07-17") & survey_type=="release" ~ as.Date("2026-07-18"), .default = date ), 
         survey_type = case_when(survey_type=="zoo_reintroduction" ~ "release", .default = survey_type ),
         ) |> 
  arrange(date)
```
```{r}
view(survey_10114_build)
```

### Survey data checks

- Run checks before and after data cleaning (next code chunk) to ensure that all issues have been resolved.  
- Run lines of code one at a time to allow easy evaluation of messages and outputs.  

```{r survey-data-checks}
# Check for duplicate survey dates
survey_10114_build %>% count(site_id, date) %>% filter(n > 1)

# Check for sites that were unsuitable for survey
# survey_10114_build %>% filter(visit_status != "suitable")
# Based on my visual inspection of the CMR survey data and the fact that they are CMR data, the surveys were all suitable
```

### Clean survey data

- Edit data-cleaning code as necessary. 

```{r clean-survey-data}
# Drop duplicate surveys
survey_10114_build <- survey_10114_build %>% 
  distinct(date, .keep_all = TRUE)
```

### Create primary and secondary period columns in mrmr format

- Edit case_when steps to reflect actual translocation/reintroduction and survey history.  

#### For robust design dataset

```{r create-period-columns1}
survey_10114 <- survey_10114_build %>%
  rename(survey_date=date) |> 
  select(site_id, survey_date, survey_type) %>% 
  mutate(primary_period = case_when(# 2021
                                    survey_date == "2021-07-28" ~ 1,
                                    # 2022
                                    between(survey_date, ymd("2022-07-18"), ymd("2022-07-20")) ~ 2,
                                    survey_date == "2022-07-21" ~ 3,
                                    # 2023
                                    between(survey_date, ymd("2023-08-25"), ymd("2023-08-27")) ~ 4,
                                    survey_date == "2023-08-28" ~ 5,
                                    # 2024
                                    between(survey_date, ymd("2024-07-15"), ymd("2024-07-17")) ~ 6,
                                    survey_date == "2024-07-19" ~ 7,
                                    between(survey_date, ymd("2024-09-06"), ymd("2024-09-07")) ~ 8, #,
                                    # 2025
                                    between(survey_date, ymd("2025-08-08"), ymd("2025-08-10")) ~ 9, #,
                                    # 2026
                                    between(survey_date, ymd("2026-07-14"), ymd("2026-07-17")) ~ 10, #, One additional frog NEWLY TAGGED on 7/17 outside of normal CMR survey means there was essentially a fourth CMR day, but it isn't a true replicate
                                    survey_date == "2026-07-18" ~ 11, # Given the above 'survey' on July 17, I displaced the release date to one-day later
                                    between(survey_date, ymd("2026-08-15"), ymd("2026-08-17")) ~ 12 #,
                                    )
         )
                                    
survey_10114 <- survey_10114 %>% 
  arrange(primary_period, survey_date) %>% 
  group_by(primary_period) %>% 
  mutate(secondary_period = seq_len(n()),
         secondary_period = replace(secondary_period, survey_type != "cmr", 0)) 
```
```{r}
view(survey_10114)
```


### Save cleaned survey file

```{r save-cleaned-survey-data}
survey_10114 %>% write_csv(here::here("data", "clean", "10114_survey.csv"))
```

## Create capture dataset

### Retain only relevant columns

```{r capture-relevant-columns}
# tentatively, this switches all captures to site_id == 10114, regardless of occurring at the other two sites 11896 and 10110.
# in 2022 and 2023, 6 out of 16 day time CMR captures occurred outside of 10114

capture_10114 <- capture_10114 %>% 
  # select(-visit_id, -survey_id, -capture_survey_id) %>% 
  rename(visit_date = date) |> 
  mutate(site_id = 10114 )
```

### Append current-year capture records

- Copy the following code chunk and insert immediately following the "capture-relevant-columns" code chunk in "xxxxx_createdata_mrmr.Rmd".
```{r append_currentyear_capture_records}
capture_10114_build <- capture_append_10114 %>% 
  select(site_id, visit_date, visit_status, survey_type, pit_tag_ref, tag_new, species, capture_life_stage, capture_animal_state, length, comment_visit, comment_survey, comment_capture) %>% 
  rename(pit_tag_id=pit_tag_ref) |> 
  bind_rows(capture_10114)
```

### Capture data checks

- Run checks before and after data cleaning (next code chunk) to ensure that all issues have been resolved.  
- Run lines of code one at a time to allow easy evaluation of messages and outputs.  

```{r capture-data-checks, message=FALSE}
# Check for records where pit_tag_ref is NULL
assert_that(noNA(capture_10114_build$pit_tag_id))

# Check for records containing pit tags shorter than 15 characters (or 9 for early sites)
assert_that(!any(nchar(capture_10114_build$pit_tag_id) != 15))

# Check for records containing pit tags that end with ".1" (known to be potentially incorrect) 
assert_that(!any(str_detect(capture_10114_build$pit_tag_id, "\\.1$")))

# Check for records where species != ramu OR rasi - this is an issue in August 2026 wherein one site may be categorized with both RASI and RAMU
capture_10114_build %>% drop_na(species) %>% filter(species != "ramu")
capture_10114_build %>% drop_na(species) %>% filter(species != "rasi")

# Check for records were survey_type = swab
assert_that(!any(capture_10114_build$survey_type == "swab"))

# Check for records where capture_animal_state is NULL
assert_that(noNA(capture_10114_build$capture_animal_state))

# Check for records where capture_animal_state == "dead"
assert_that(!any(capture_10114_build$capture_animal_state == "dead"))
### resolved by removing dead frogs in next chunk

# Check for records of subadult frogs
capture_10114_build %>% drop_na(length) %>% filter(length < 40)

# Check for sites that were unsuitable for survey
assert_that(!any(capture_10114_build$visit_status != "suitable"))

# Check for duplicate capture records on each date
capture_10114_build %>% count(visit_date, pit_tag_id) %>% filter(n > 1)

# Check for captures that lack associated surveys
survey_10114 %>% 
  select(survey_date, primary_period) %>% 
  rename(visit_date = survey_date) %>% 
  right_join(capture_10114_build, by = c("visit_date")) %>% 
  filter(is.na(primary_period))

# Simpler alternative
setdiff(capture_10114_build$visit_date, survey_10114$survey_date) %>% as.Date(origin = "1970-01-01")

# Check for surveys that lack associated captures
survey_10114 %>% 
  filter(survey_type == "cmr") %>% 
  select(survey_date, primary_period) %>% 
  rename(visit_date = survey_date) %>% 
  left_join(capture_10114_build, by = c("visit_date")) %>% 
  filter(is.na(species))
## output: no frogs observed at 10114 or neighbors on first-ever CMR survey date 2022-07-18
# 2022-07-18
# 2024-07-16
# 2025-08-08
# 2026-07-16
# 2026-08-15
```

### Clean capture data

- Edit data-cleaning code as necessary.  

```{r clean-capture-data}
capture_10114_build <- capture_10114_build %>% 
  filter(capture_animal_state != "dead")
```


```{r final-capture-data}
capture_10114 <- capture_10114_build %>%
  rename(survey_date = visit_date #, 
         # pit_tag_id = pit_tag_ref
         ) %>%
  arrange(survey_date, pit_tag_id) |> 
  select(site_id, survey_date, pit_tag_id)
```

### Save cleaned capture data

```{r save-cleaned-capture-data}
capture_10114 %>% write_csv(here::here("data", "clean", "10114_capture.csv"))
```

# Covariates

### Survey-level covariates
Covariates that might affect the probability of detection can be included via the capture_formula argument to clean_data.

### Individual-level survival covariates 
Covariates that may affect individual level survival can be included optionally via the survival_formula and survival_fill_value arguments

```{r build-covariates as needed}

```


