---
title: "10486 - Analyze capture-mark-recapture data using *mrmr*"
author: "Roland Knapp"
output: html_notebook
editor_options: 
  chunk_output_type: inline
---
This notebook analyzes capture-mark-recapture data using the *mrmr* package. The code creates the required survey, capture, and translocation data frames from saved CSV files (created by xxxxx_createdata_mrmr.Rmd), checks the datasets, fits the model, evaluates potential non-convergence of the MCMC algorithm, and visualizes model results. The code is written to be run on an ERI HPC server ("tong"). 

## Install latest versions of cmdstanr, cmdstan, and mrmr packages
Run only if updated versions were released since previous install
```{r install-stan-mrmr-packages}
# install cmdstanr and cmdstan
install.packages("cmdstanr", repos = c("https://mc-stan.org/r-packages/", getOption("repos")))
cmdstanr::install_cmdstan()

# install mrmr
remotes::install_github("SNARL1/mrmr")
```

## Load required packages
```{r load-packages}
library(dplyr)
library(readr)
library(tibble)
library(cmdstanr)
library(mrmr)
library(patchwork)
```

## Create required data frames
```{r create-data-objects}
captures <- read_csv(here::here("data", "clean", "10486_capture.csv"))
surveys <- read_csv(here::here("data", "clean", "10486_survey.csv"))
```

## Check datasets
```{r check-data}
data_10486 <- clean_data(captures = captures, surveys = surveys)
```

## Fit model
Code assumes that there are 8 physical cores available. Update parallel_chains and threads_per_chain as necessary based on actual number of cores. 
Add "control = list(adapt_delta = 0.99)" or similar if divergent transitions occur. 
```{r fit-model}
model_10486 <- fit_model(
                    data = data_10486, 
                    chains = 4, 
                    parallel_chains = 4, 
                    cpp_options = list(stan_threads = TRUE),
                    threads_per_chain = 2) 
```
* Running model with default adapt_delta (0.8) resulted in zero divergent transition (0/4000).

### Save/retrieve model object
This step prevents having to rerun the model in the event of a crash during model visualizations in subsequent code blocks. Model files can be large, so this saves the file in a compressed format.
```{r write-model}
write_rds(model_10486, here::here("out", "model", "10486_model.rds"), "xz", compression = 9L)
```
```{r read-model}
# model_10486 <- read_rds(here::here("out", "model", "10486_model.rds"))
```

### Diagnose potential non-convergence of the MCMC algorithm
Inspect traceplots and parameter summary table. In table, are any R~hat~ values $\ge 1.01$?
```{r model-diagnostics}
pars_to_plot <- c('alpha_lambda', 
                  'sigma_lambda', 
                  'beta_phi', 
                  'sigma_phi', 
                  'beta_detect')
bayesplot::mcmc_trace(model_10486$m_fit$draws(pars_to_plot))

model_10486$m_fit$summary(pars_to_plot)
```

## Visualize model results
### Default plots
Note: Custom plots (produced in next code chunk: "plots_custom") are often easier to interpret than the default plots produced by this code chunk.
```{r plots-default}
# plot_model(model_10486, what = "abundance")
# plot_model(model_10486, what = "recruitment")
# plot_model(model_10486, what = "survival")
```

### Customize plots
```{r plots-custom, fig.height=5, fig.width=7}
library(ggplot2)
p1 <- plot_model(model_10486, what = "abundance") 
p1 <- p1 + ggtitle("A. Population size") +
    ylab("Number of adults") + 
    expand_limits(y = 0)
  
p2 <- plot_model(model_10486, what = "recruitment")
p2 <- p2 + ggtitle("B. Recruitment") +
    ylab("Number of adults")
  
p1 / p2
```

### Save custom plots
```{r save-plot}
ggsave(here::here("out", "plots", "10486_mrmr_plots.png"), width = 6.5, height = 4.8)
```
