---
title: "20135 - Analysis of CMR data using *mrmr*"
author: "Roland Knapp"
output: html
---
Analysis of Ranger Station Pond CMR dataset - Tyndall area
- "treatment" covariate is included in translocation file to evaluate survival of Bd-exposed and unexposed frogs

## Install mrmr package if current version is not available

```{r install_mrmr}
remotes::install_github("SNARL1/mrmr")
```

## Install cmdstanr package if current version is not available

```{r}
install.packages("cmdstanr", repos = c("https://stan-dev.r-universe.dev", getOption("repos")))
cmdstanr::install_cmdstan()
```

## Load packages

```{r load_packages}
if (!require(librarian)){
  install.packages("librarian")
  library(librarian)
}
librarian::shelf(here, cmdstanr, mrmr, bayesplot, posterior, tidyverse)
```

## Create required datasets

```{r mrmr_datasets}
captures <- read_csv(here("data", "clean", "20135_capture.csv")) # should I have used 20135_bd_capture.csv instead? Check. 
surveys <- read_csv(here("data", "clean", "20135_survey.csv"))
translocations <- read_csv(here("data", "clean", "20135_reintroduction.csv"))
```

## Specify model formula and check data

```{r mrmr_model_formula}
data_20135 <- clean_data(captures, surveys, translocations,
                                survival_formula = ~ treatment,
                                survival_fill_value = c(treatment = "unexposed"))
```

## Fit model

```{r mrmr_fit_model}
# Used default of 1000 iterations for warmup and 1000 iterations for sampling
# "threads per chain" option and "cpp_options" did not speed up fit_model step
model_20135 <- fit_model(
  data_20135, 
  chains = 4,
  parallel_chains = 4,
  threads_per_chain = 4,
  cpp_options = list(stan_threads = TRUE)
)
```

## Save model object

```{r}
model_20135 %>% write_rds(here("out", "model", "20135_model.rds"), "xz", compression = 9L)
```

## Read model objects back in (as necessary)

```{r eval=FALSE}
model_20135 <- read_rds(here("out", "model", "20135_model.rds"))
```

## Evaluate convergence of the MCMC algorithm

- alpha_lamda: Baseline (intercept) of the recruitment process. In a Jolly-Seber model, recruitment (entry of new individuals into the population) in each primary period is modeled using a log link or logit link.
For instance, if log⁡(λt )=αλ+ϵλ, then alpha_lambda is the overall mean log recruitment rate across time period.
- sigma_lamda: Standard deviation of temporal random effects on recruitment (ε_λ,t). Therefore, sigma_lamda describes variability in recruitment through time. 
- beta_phi[1]: Intercept (baseline effect) for survival probability (φ). Survival is typically modeled on the logit scale: logit(ϕi,t)=βϕ,1+βϕ,2⋅Treatmenti+ϵϕ,t. beta_phi[1] is the average (logit-scale) survival for the reference group — typically "unexposed".
- beta_phi[2]: Treatment effect on survival probability. Continuing from beta_phi[1] above, beta_phi[2] modifies the baseline logit of survival for the treatment group (e.g., Bd-exposed vs. unexposed). Positive versus negative values indicate that the treatment increased or decreased survival, respectively. 
- sigma_phi: Standard deviation of temporal random effects on survival (ε_φ,t). Captures variability in survival between time periods.
- beta_detect[1]: Intercept (baseline effect) for detection probability (p). Represents the average log-odds of detecting an individual per secondary period, in the baseline group (e.g., unexposed).

```{r mrmr_convergence}
# Traceplots
pars_to_plot <- c("alpha_lambda", 
                  "sigma_lambda", 
                  "beta_phi", 
                  "sigma_phi", 
                  "beta_detect")

mcmc_trace(model_20135$m_fit$draws(pars_to_plot))

# Summary statistics, include Rhat (any Rhat > 1.01?)
model_20135$m_fit$summary(pars_to_plot)
```

## Built-in visualizations

```{r mrmr_visualizations}
plot_model(model_20135, what = "abundance")
plot_model(model_20135, what = "recruitment")
plot_model(model_20135, what = "survival")
```

## Compare posterior estimates for survival: Bd-exposed vs. unexposed individudals

- rstan saves the survival value as phi. In contrast, cmdstanr stores the components of the survival variable: beta_phi[1] = intercept, beta_phi[2] = treatment_effect, and eps_phi[1:13] = time (primary period) random effects. The following code calculates phi from these components. 

```{r}
# Extract phi components directly from cmdstanr fit
beta_phi_mat <- posterior::as_draws_matrix(model_20135$m_fit$draws("beta_phi"))
eps_phi_mat  <- posterior::as_draws_matrix(model_20135$m_fit$draws("eps_phi"))

# Dimensions
n_draws   <- nrow(beta_phi_mat)
n_periods <- ncol(eps_phi_mat)
treatment_levels <- unique(data_20135$survival_covariate_df$treatment)

# Compute posterior phi for each treatment (vectorized)
phi_list <- vector("list", length(treatment_levels))
names(phi_list) <- treatment_levels

for (i in seq_along(treatment_levels)) {
  treatment <- treatment_levels[i]
  
  # Treatment coding: 0 = unexposed, 1 = exposed
  treat_effect <- ifelse(treatment == "unexposed", 0, 1)
  
  # Create matrices for vectorized addition
  intercept_mat <- matrix(beta_phi_mat[,1], nrow = n_draws, ncol = n_periods)
  treatment_mat <- matrix(beta_phi_mat[,2] * treat_effect, nrow = n_draws, ncol = n_periods)
  
  # Compute logit(phi) = intercept + treatment effect + random effect
  logit_phi <- intercept_mat + treatment_mat + eps_phi_mat
  
  # Convert to probability
  phi_list[[treatment]] <- plogis(logit_phi)
}

# Combine into 3D array: draws × periods × treatment
phi_array <- abind::abind(phi_list, along = 3)
dimnames(phi_array) <- list(
  iteration = NULL,
  primary_period = 1:n_periods,
  treatment = treatment_levels
)

# Combine and tidy for plotting
survival_df <- reshape2::melt(
  phi_array,
  varnames = c("iteration", "primary_period", "treatment"),
  value.name = "phi"
) %>% as_tibble()

# Merge with survey metadata (e.g., survey_date)
survival_df <- survival_df %>%
  left_join(
    data_20135$surveys %>% filter(secondary_period < 2),
    by = "primary_period"
  )

# Plot posterior survival densities
ggplot(survival_df, aes(x = phi, color = treatment)) +
  stat_density(geom = "line") +
  facet_wrap(~ survey_date, scales = "free_y") +
  xlab("Survival probability") +
  ylab("Posterior density") +
  scale_color_discrete("Treatment")

# Posterior probability that treatment increased survival (using beta_phi)
treatment_prob <- mean(beta_phi_mat[,2] > 0) 
# for each draw from posterior samples of the treatment effect on survival, checks whether the effect is positive, i.e., treatment increases survival

cat("Posterior probability that treatment increased survival:", treatment_prob, "\n")
```

- Posterior probability that treatment increased survival: 0.206. Therefore, there is a 21% posterior probability that the treatment increased survival. Weak evidence at best.

## Plot the posterior distribution of the treatment effect

```{r}
beta_df <- data.frame(beta_treatment = beta_phi_mat[, "beta_phi[2]"]) 
colnames(beta_df)

beta_df <- beta_df %>% 
  rename(beta_treatment = beta_phi.2.)
colnames(beta_df)

ggplot(beta_df, aes(x = beta_treatment)) +
  geom_density(fill = "skyblue", alpha = 0.5) +
  geom_vline(xintercept = 0, color = "red", linetype = "dashed") +
  xlab("Treatment effect (logit scale)") +
  ylab("Posterior density") +
  ggtitle("Posterior distribution of treatment effect on survival")

```

## Custom plots of cohort survival by treatment group

- Plot cohort survival by treatment group, with 95% credible intervals and lines/points on the posterior median.

```{r}
primary_period_dates <- model_20135$data$surveys %>%
  group_by(primary_period) %>%
  summarize(date = min(survey_date),
            year = min(year))

transloc_dates <- distinct(
      model_20135$data$translocations,
      pit_tag_id,
      release_date
    )

p1 <- model_20135$m_fit$draws('s', format = "draws_df") %>%
  tidyr::pivot_longer(tidyselect::starts_with('s')) %>%
  mutate(get_numeric_indices(name)) %>%
      rename(
        primary_period = index_2
      ) %>% 
  mutate(pit_tag_id = dimnames(model_20135$data$stan_d$Y)[[1]][index_1]) %>%
  filter(pit_tag_id %in% as.character(model_20135$data$translocations$pit_tag_id)) %>%
  left_join(transloc_dates, by = "pit_tag_id") %>%
  left_join(model_20135$data$survival_covariate_df) %>%
  group_by(release_date, primary_period, .draw, treatment) %>%
  summarize(fraction_alive = mean(value == 2), .groups = "drop") %>%
  filter(primary_period > 1) %>%
  left_join(primary_period_dates) %>%
  filter(date >= release_date) %>%
  group_by(release_date, date, treatment) %>%
  summarize(lo = quantile(fraction_alive, .025),
            med = median(fraction_alive),
            hi = quantile(fraction_alive, .975),
            .groups = "drop") %>%
  mutate(grp = paste(treatment, release_date), 
         release_date = paste("Released on", release_date)) %>%
  ggplot(aes(date, med,
             color = tools::toTitleCase(treatment),
             fill = tools::toTitleCase(treatment), 
             group = grp)) +
  geom_ribbon(aes(ymin = lo, ymax = hi),
              alpha = .4, color = NA) +
  geom_point() +
  geom_line() +
  ylab("Fraction surviving") +
  xlab("Date") +
  scale_color_brewer("Treatment", palette = "Dark2") +
  scale_fill_brewer("Treatment", palette = "Dark2") +
  scale_x_date(date_breaks = "1 year") + 
  facet_wrap(~release_date, ncol = 1)

p1 <- p1 + ggtitle('A. Survival of reintroduced frogs') +
  theme(legend.position = c(0.6, 0.8))

p1
```

## Arrange survival, abundance, and recruitment plots

```{r}
# survival plot
p1 <- p1 +
#  guides(fill = guide_legend(nrow = 1)) + 
  theme(legend.position = "right")

# abundance plot
p2 <- plot_model(model_20135, what = "abundance") 
p2 <- p2 + ggtitle("B. Population size") +
    ylab("Number of adults")

# recruitment plot
p3 <- plot_model(model_20135, what = "recruitment")
p3 <- p3 + ggtitle("C. Recruitment") +
    ylab("Number of adults")

library(patchwork) 

p1 / p2 / p3 +
  plot_layout(heights = c(3, 1, 1))

ggsave(here("out", "plots", "20135_mrmr_plots_treatment.png"), width = 6.5, height = 7.2)
```

## Quick model check

- If the treatment increased survival, then in 2018 and 2019, we should have seen a greater proportion of treated than untreated frogs from the initial translocation.

```{r}
cohort_transloc <- translocations %>%
  filter(release_date == as.Date("2016-07-18"))

count(cohort_transloc, treatment) %>%
  mutate(frac = n / sum(n))
  
captures %>%
  inner_join(translocations, by = "pit_tag_id") %>% 
  select(survey_date, pit_tag_id, treatment) %>% 
  filter(survey_date > as.Date("2018-01-01"), 
         pit_tag_id %in% cohort_transloc$pit_tag_id) %>%
  group_by(survey_date) %>%
  count(treatment) %>%
  mutate(frac = n / sum(n))
```

- Our null expectation is that we should see approximately 71% exposed and 29% unexposed individuals based on the proportions of frogs in each category in this first translocation.
- For each survey, we see a somewhat larger fraction of exposed individuals than expected based on our null expectation.
