  ---
title: "An introduction to mrmr: Ranger Station"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{intro-to-mrmr-tyndall}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>", 
  fig.width = 7, 
  fig.height = 3.5
)
```

```{r setup}
library(dplyr)
library(readr)
library(tibble)
library(rstan)
library(mrmr)
```

## Data specifications

The first file specifies capture data:

```{r show-capture-data}
captures <- "../data/captures.rangerstapond.mrmr.csv" %>%
  read_csv

glimpse(captures)
```

The second file specifies translocation data: 

```{r show-trans-data}
translocations <- "../data/translocations.rangerstapond.mrmr.csv" %>%
  read_csv

glimpse(translocations)
```

Each row corresponds to a translocation event of one unique individual. 

The third file specifies survey-level data:

```{r show-survey-data}
surveys <- "../data/surveys.rangerstapond.mrmr.csv" %>%
  read_csv
    
glimpse(surveys)
```

Here, each row is a unique survey.


### Including survival covariates

Individual-level survival covariates can be included optionally via the 
`survival_formula` and `survival_fill_value` arguments. 
Both of these must be specified, because covariate values for pseudo-individuals 
must be filled in (they are never observed).
So, for example, to evaluate the effect of an experimental treatment, if some 
individuals belong to a "treatment" group and others belong to an "control" 
group, then a group must be specified as a fill value (e.g., "wild-caught" or 
"control", depending on the experiment):


```{r survival-covariate, message=FALSE, eval=FALSE}
# specify the formula
data <- clean_data(captures, surveys, translocations,
                   survival_formula = ~ treatment, 
                   survival_fill_value = c(treatment = "unexposed"))
```


## Model structure

The `mrmr` package implements a Bayesian open-population Jolly-Seber mark recapture model with known additions to the population (introduced adults).
The model tracks the states of $M$ individuals that comprise a superpopulation made up of real and pseudo-individuals (see Joseph and Knapp 2018, *Ecosphere* for details).

We assume a robust sampling design where the states of individuals are constant within primary periods, which have repeat secondary periods within them (during which observations are made). 
The possible states of individuals include "not recruited", "alive", and "dead". 
The possible observations of individuals include "detected" and "not detected". 
We assume that individuals that are in the "not recruited" or "dead" states are never detected (i.e., there are no mistakes in the individual PIT tag ID records).

## Model fitting

To fit a mark-recapture model, use `mrmr::fit_model()`. 
This model accounts for known introductions into the population, and has 
random effects to account for variation in survival and recruitment through time. 

```{r fit-model}
mod <- fit_model(data, cores = parallel::detectCores())
```

Save model object
```{r}
# compressed file

write_rds(mod, "../out/model_rangersta_mrmr_24apr2020.rds", "xz", compression = 9L)
```

To diagnose potential non-convergence of the MCMC algorithm, inspect traceplots:

```{r plot-traceplots}
pars_to_plot <- c('alpha_lambda', 
                  'sigma_lambda', 
                  'beta_phi', 
                  'sigma_phi', 
                  'beta_detect')
traceplot(mod$m_fit, pars = pars_to_plot)
```

It is also a good idea to look at Rhat estimates to see whether they 
indicate of a lack of convergence (Rhat values $\geq$ 1.01):

```{r print-model}
print(mod$m_fit, pars = pars_to_plot)
```


Compare posterior estimates for survival between Bd-exposed and unexposed 
individudals. 

```{r}
post <- rstan::extract(mod$m_fit)

# Check that all translocation treatments match
t1 <- data$survival_covariate_df %>%
  filter(pit_tag_id %in% translocations$pit_tag_id) %>%
  arrange(pit_tag_id)
t2 <- distinct(translocations, pit_tag_id, treatment) %>%
  mutate(pit_tag_id = as.character(pit_tag_id)) %>%
  arrange(pit_tag_id)
stopifnot(all_equal(t1$treatment, t2$treatment))


survival_df <- post$phi %>%
  reshape2::melt(varnames = c("iter", "individual", "primary_period")) %>%
  as_tibble %>%
  mutate(pit_tag_id = data$survival_covariate_df$pit_tag_id[individual]) %>%
  left_join(data$survival_covariate_df) %>%
  distinct(iter, primary_period, value, treatment) %>%
  left_join(data$surveys %>% filter(secondary_period < 2))

survival_df %>%
  ggplot(aes(x = value, color = treatment)) + 
  stat_density(geom = "line", position = position_dodge(width=0)) + 
  facet_wrap(~ survey_date, scales = 'free_y') + 
  xlab("Survival probability") + 
  ylab("Posterior density") + 
  scale_color_discrete("Treatment")

paste("Posterior probability that treatment increased survival:", 
      mean(post$beta_phi[, 2] < 0))
```


## Built-in visualizations

Time series of abundance, recruitment, and survival of introduced cohorts are
available through the `mrmr::plot_model()` function. 

```{r plot-abundance}
plot_model(mod, what = 'abundance')
```

```{r plot-recruitment}
plot_model(mod, what = 'recruitment')
```

```{r plot-survival}
plot_model(mod, what = 'survival')
```

## Survival table
```{r}
survival_table(mod)
```


## Custom visualizations

Any of the plotting functionality that you would expect from a `stanfit` model
is available as well, by accessing the `m_fit` list element from a model object. 
For example, we could assess the posterior for the superpopulation size:

```{r}
traceplot(mod$m_fit, pars = "Nsuper")
```


Plot cohort survival by treatment group, with 95% credible intervals and 
lines/points on the posterior median.

```{r}
primary_period_dates <- mod$data$surveys %>%
  group_by(.data$primary_period) %>%
  summarize(date = min(.data$survey_date),
            year = min(.data$year)) %>%
  ungroup

mod$post$s %>%
  reshape2::melt(varnames = c('iter', 'i', 'primary_period')) %>%
  as_tibble %>%
  mutate(pit_tag_id = dimnames(mod$data$stan_d$Y)[[1]][.data$i]) %>%
  filter(.data$pit_tag_id %in% as.character(mod$data$translocations$pit_tag_id)) %>%
  left_join(distinct(mod$data$translocations,
                     .data$pit_tag_id, .data$release_date)) %>%
  left_join(mod$data$survival_covariate_df) %>%
  group_by(.data$release_date, .data$primary_period, .data$iter, .data$treatment) %>%
  summarize(fraction_alive = mean(.data$value == 2)) %>%
  ungroup %>%
  filter(.data$primary_period > 1) %>%
  left_join(primary_period_dates) %>%
  filter(.data$date >= .data$release_date) %>%
  group_by(.data$release_date, .data$date, .data$treatment) %>%
  summarize(lo = quantile(.data$fraction_alive, .025),
            med = median(.data$fraction_alive),
            hi = quantile(.data$fraction_alive, .975)) %>%
  ungroup %>%
  mutate(grp = paste(treatment, release_date), 
         release_date = paste("Released on", release_date)) %>%
  ggplot(aes(.data$date, .data$med,
             color = tools::toTitleCase(.data$treatment),
             fill = tools::toTitleCase(.data$treatment), 
             group = .data$grp)) +
  geom_ribbon(aes(ymin = .data$lo, ymax = .data$hi),
              alpha = .4, color = NA) +
  geom_line() +
  geom_point() +
  ylab('Fraction surviving') +
  xlab('Date') +
  scale_color_brewer('Treatment', palette = "Dark2") +
  scale_fill_brewer('Treatment', palette = "Dark2") +
  scale_x_date(date_breaks = "1 year") + 
  facet_wrap(~release_date, ncol = 1)
```

## Export plot
```{r}
ggsave("../out/rangersta_survivalbygroup_plot.png", width = 6.5, height = 4.5)
```

## Customize and arrange plots
```{r}
# survival plot

primary_period_dates <- mod$data$surveys %>%
  group_by(.data$primary_period) %>%
  summarize(date = min(.data$survey_date),
            year = min(.data$year)) %>%
  ungroup

p1 <- mod$post$s %>%
  reshape2::melt(varnames = c('iter', 'i', 'primary_period')) %>%
  as_tibble %>%
  mutate(pit_tag_id = dimnames(mod$data$stan_d$Y)[[1]][.data$i]) %>%
  filter(.data$pit_tag_id %in% as.character(mod$data$translocations$pit_tag_id)) %>%
  left_join(distinct(mod$data$translocations,
                     .data$pit_tag_id, .data$release_date)) %>%
  left_join(mod$data$survival_covariate_df) %>%
  group_by(.data$release_date, .data$primary_period, .data$iter, .data$treatment) %>%
  summarize(fraction_alive = mean(.data$value == 2)) %>%
  ungroup %>%
  filter(.data$primary_period > 1) %>%
  left_join(primary_period_dates) %>%
  filter(.data$date >= .data$release_date) %>%
  group_by(.data$release_date, .data$date, .data$treatment) %>%
  summarize(lo = quantile(.data$fraction_alive, .025),
            med = median(.data$fraction_alive),
            hi = quantile(.data$fraction_alive, .975)) %>%
  ungroup %>%
  mutate(grp = paste(treatment, release_date), 
         release_date = paste("Released on", release_date)) %>%
  ggplot(aes(.data$date, .data$med,
             color = tools::toTitleCase(.data$treatment),
             fill = tools::toTitleCase(.data$treatment), 
             group = .data$grp)) +
  geom_ribbon(aes(ymin = .data$lo, ymax = .data$hi),
              alpha = .4, color = NA) +
  geom_line() +
  geom_point() +
  ylab('Fraction surviving') +
  xlab('Date') +
  scale_color_brewer('Treatment', palette = "Dark2") +
  scale_fill_brewer('Treatment', palette = "Dark2") +
  scale_x_date(date_breaks = "1 year") + 
  facet_wrap(~release_date, ncol = 1)

p1 <- p1 + ggtitle('A. Survival of reintroduced frogs') +
  theme(legend.position = c(0.15, 0.2))

# abundance plot

p2 <- plot_model(mod, what = 'abundance') 
p2 <- p2 + ggtitle('B. Population size') +
    ylab("Number of adults")

# recruitment plot

p3 <- plot_model(mod, what = 'recruitment')
p3 <- p3 + ggtitle('C. Recruitment') +
    ylab("Number of adults")

library(patchwork) 

p1 / p2 / p3 +
  plot_layout(heights = c(2.5, 1, 1))
ggsave("../out/rangersta_plot_all.png", width = 6.5, height = 7.2)
```


Quick sanity check: if the treatment increased survival, then in 2019, we 
should have seen more treated animals than untreated from the initial 
translocation.

```{r}
cohort_transloc <- translocations %>%
  filter(release_date == as.Date("2016-07-18"))

count(cohort_transloc, treatment) %>%
  mutate(frac = n / sum(n))
# our null expectation is that we should see ~69% exposed individuals, 
# and 31% unexposed from this first translocation

captures %>%
  filter(survey_date > as.Date("2019-01-01"), 
         pit_tag_id %in% cohort_transloc$pit_tag_id) %>%
  group_by(survey_date) %>%
  count(category) %>%
  mutate(frac = n / sum(n))
# for each survey, we see a larger fraction of exposed individuals than
# expected based on our null expectation.
```