---
title: "20135 - Create datasets for analysis using *mrmr*"
author: "Roland Knapp"
output: html_notebook
editor_options: 
  chunk_output_type: inline
---

This notebook creates the survey, capture, and translocation files for use in analyzing capture-mark-recapture data using the *mrmr* package. The code retrieves data from the PostgreSQL amphibians database, runs error checks, and creates the relevant files in *mrmr* format. A treatment column is included in the reintroduction file (exposed, unexposed) for inclusion as a survival covariate. 

Users without credentials to access the database will not be able to create or use the raw data file that is central to this notebook. See cmr-analysis/README.md for details. 

Note that data from the most recent field season may not yet be appended to tables in the amphibians database. If these recently-collected data are necessary (e.g., for preliminary analyses), complete datasets can be created by copying code chunks in "fulcrum_data_append.Rmd" into this notebook and executing all code chunks. 

## Load packages

```{r load_packages}
if (!require(librarian)){
  install.packages("librarian")
  library(librarian)
}
librarian::shelf(DBI, config, here, readr, connections, dplyr, tidyr, stringr, magrittr, lubridate, assertthat)
```

## Connect to database

- # Before running this code chunk, establish connection to database server (via VPN or SSH)

```{r db_connect}
# Select configuration of database of interest (connection paramaters for all databases provided in config.yml - in working directory)
Sys.setenv(R_CONFIG_ACTIVE = "amphibians-eri")

# Using connection_open from connections package shows structure of database tables in Connections pane
con <- connection_open(RPostgres::Postgres(),
                       dbname = config::get("dbname"),
                       host = config::get("host"),
                       port = config::get("port"),
                       user = config::get("user"),
                       password = rstudioapi::askForPassword("password")) # Enter password from password manager ("ERI GRIT Postgres")

# Set search path to schema of interest
dbExecute(con, "SET search_path TO public")

```

## Retrieve data subsets from PostgreSQL database

### Reintroduction data

```{r reintroduction_data}
relocate <- tbl(con, "relocate") %>%
  rename(
    relocate_comment = comment,
    relocate_id = id
  )

relocate_frog <- tbl(con, "relocate_frog") %>%
  rename(
    relocate_frog_comment = comment,
    relocate_frog_id = id
  )

reintroduction_20135 <- relocate %>%
  inner_join(relocate_frog, by = "relocate_id") %>%
  filter(
    release_siteid1 %in% c(20135, 21521, 21519, 21520)
  ) %>%
  select(
    release_siteid1,
    release_date,
    type,
    pit_tag_ref,
    tag_new,
    length,
    weight,
    swab_id,
    bd_exposure,
    relocate_comment,
    relocate_frog_comment
  ) %>%
  collect()
```

Review the reintroduction data subset for potential problems. 
- There is one 9 digit pit tag in dataset. Is it missing a 6-digit prefix?
- tag_new and comment fields are empty. 

### Survey data

To allow subsequent checks of whether surveys at which no frogs were captured are recorded as "visual" instead of "cmr", include visual surveys

```{r survey_data}
visit <- tbl(con, "visit") %>%
  rename(
    visit_comment = comment,
    visit_id = id
  )

survey <- tbl(con, "survey") %>%
  rename(
    survey_comment = comment,
    survey_id = id
  )

survey_20135 <- visit %>% 
  inner_join(survey, by = "visit_id") %>% 
  filter(
    site_id %in% c(20135, 21521, 21519, 21520),
    survey_type %in% c("cmr", "visual")
  ) %>% 
select( 
  site_id, 
  visit_date, 
  visit_status, 
  survey_type, 
  description, 
  survey_quality,
  wind,
  sky,
  duration,
  visit_comment,
  survey_comment
) %>% 
collect()

survey_20135 <- survey_20135 %>%  # convert ENUM columns to character
  mutate(across(where(is_pq_enum), as.character))
```

Review the survey data subset for potential problems. 
- Based on survey comments, few RAMU seen after 2019. Consider truncating capture data.
- Some surveys with inclement weather, likely worth retaining all surveys regardless. 

### Capture data

```{r capture_data}
visit <- tbl(con, "visit") %>%
  rename(
    visit_id = id
  )

survey <- tbl(con, "survey") %>%
  rename(
    survey_id = id
  )

capture <- tbl(con, "capture_survey") %>% 
  rename(
    capture_survey_comment = comment
  )

bdload <- tbl(con, "bd_load") %>% 
  rename(
    swab_id = sample_id
  )

capture_20135 <- visit %>% 
  inner_join(survey, by = "visit_id") %>% 
  inner_join(capture, by = "survey_id") %>% 
  left_join(bdload, by = "swab_id") %>% 
  filter(
    site_id %in% c(20135, 21521, 21519, 21520),
    survey_type == "cmr"
  ) %>% 
select( 
  site_id, 
  visit_date, 
  visit_status, 
  survey_type, 
  pit_tag_ref, 
  tag_new, 
  species, 
  capture_life_stage, 
  capture_animal_state, 
  length,
  swab_id,
  replicate,
  std_type,
  dilution,
  bd_load,
  capture_survey_comment
) %>% 
collect()

capture_20135 <- capture_20135 %>%  # convert ENUM columns to character
  mutate(across(where(is_pq_enum), as.character))
```

Review the capture data subset for potential problems. 
- No frogs captured after 2019.
- qPCR used gblock and plasmid standards, all samples run singly.

## Save raw data files

```{r save_raw_data, message=FALSE}
reintroduction_20135 %>% 
  write_csv(here::here("data", "raw", "20135_reintroduction_raw.csv"))

survey_20135 %>% 
  write_csv(here::here("data", "raw", "20135_survey_raw.csv"))

capture_20135 %>% 
  write_csv(here::here("data", "raw", "20135_capture_raw.csv"))
```

## Disconnect from database 

```{r db_disconnect}
connection_close(con) 
rm(con)
```

## Read in raw data files (if necessary)

```{r read_raw_data, eval=FALSE, message=FALSE}
# Change to eval=TRUE
reintroduction_20135 <- read_csv(here("data", "raw", "20135_reintroduction_raw.csv")) %>% 
  mutate(pit_tag_ref = as.character(pit_tag_ref))

survey_20135 <- read_csv(here("data", "raw", "20135_survey_raw.csv"))

capture_20135 <- read_csv(here("data", "raw", "20135_capture_raw.csv")) %>% 
  mutate(pit_tag_ref = as.character(pit_tag_ref))
```

## Clean reintroduction data

Run checks before and after data cleaning (next code chunk) to ensure that all issues have been resolved. Run lines of code one at a time to allow easy evaluation of messages and outputs. 

### Reintroduction data checks

```{r reintroduction_data_checks, message=FALSE}
# Check for records where pit_tag_ref is NULL
assert_that(noNA(reintroduction_20135$pit_tag_ref))

# Check for records where bd_exposure is NULL
assert_that(noNA(reintroduction_20135$bd_exposure))

# Check for records containing pit tags shorter than 15 characters (or 9 for early sites)
assert_that(!any(str_length(reintroduction_20135$pit_tag_ref) != 15))

# Check for records containing pit tags that end with ".1" (known to be potentially incorrect) 
assert_that(!any(str_detect(reintroduction_20135$pit_tag_ref, "\\.1$")))

# Check for type != "reintroduction"
assert_that(all(reintroduction_20135$type == "reintroduction"))

# Check for frogs that died during reintroduction
assert_that(!any(str_detect(reintroduction_20135$relocate_frog_comment, "dead|died")))

# Check for duplicate pit tags on each date
assert_that(!any(reintroduction_20135 %>% count(release_date, pit_tag_ref) %>% pull(n) > 1))

# Are release dates correct?
reintroduction_20135 %>% distinct(release_date) %>% glimpse()
```

### Update reintroduction data

```{r update_reintroduction_data}
# Check for 9-digit PIT tag in capture data. Frog was never recaptured, left tag as is.
bad_ids <- reintroduction_20135 %>% 
  filter(str_length(pit_tag_ref) != 15) %>% 
  pull(pit_tag_ref)

capture_20135 %>% 
  filter(pit_tag_ref == bad_ids)

rm(bad_ids)
```

### Create final reintroduction dataset formatted for mrmr

```{r final_reintroduction_data}
reintroduction_20135_mrmr <- reintroduction_20135 %>% 
  rename(pit_tag_id = pit_tag_ref,
         site_id = release_siteid1,
         treatment = bd_exposure) %>% 
  arrange(release_date, pit_tag_id) %>% 
  select(site_id, type, release_date, pit_tag_id, treatment)
```

### Save final reintroduction data

```{r save_final_reintroduction_data, message=FALSE}
reintroduction_20135_mrmr %>% write_csv(here("data", "clean", "20135_reintroduction.csv"))
```

### Read in cleaned reintroduction data (if necessary)

```{r read_cleaned_reintroduction_data, eval=FALSE, message=FALSE}
reintroduction_20135 <- read_csv(here::here("data", "clean", "20135_reintroduction.csv")) %>% 
  mutate(pit_tag_id = as.character(pit_tag_id))
```

## Clean survey dataset

### Review cmr survey history for all sites in project area

```{r} 
cmr_survey_check <- survey_20135 %>% 
  filter(survey_type == "cmr",
         visit_date > min(reintroduction_20135$release_date) + days(5)) %>% # Remove CMR surveys from before or soon after frogs were reintroduced
  distinct(visit_date) %>%
  arrange(visit_date) %>%
  mutate(primary_period = cumsum(visit_date - lag(visit_date, default = first(visit_date)) > 5) + 1) %>% # Identify primary periods based on a gap between CMR surveys of > 5 days
  inner_join(survey_20135, by = "visit_date") %>% 
  filter(survey_type == "cmr") %>% 
  select(site_id, primary_period) %>% 
  group_by(primary_period) %>% 
  count(site_id) %>% 
  mutate(num_sec_periods = max(n),
         diff = num_sec_periods - n)
```

- Primary site 20135 surveyed during each secondary period, peripheral sites sometimes surveyed less frequently. 

### Identify VES with no associated CMR period for possible inclusion as a CMR survey - customize for each site

```{r}
# For each primary period, create all dates in period +/- 2 days
cmr_dates <- survey_20135 %>% 
  filter(survey_type == "cmr",
         visit_date > min(reintroduction_20135$release_date) + days(5)) %>% # Remove CMR surveys from before or soon after frogs were reintroduced
  distinct(visit_date) %>%
  arrange(visit_date) %>%
  mutate(primary_period = cumsum(visit_date - lag(visit_date, default = first(visit_date)) > 5) + 1) %>% # Identify primary periods based on a gap between CMR surveys of > 1 day
  group_by(primary_period) %>%
  summarise(  # buffer primary periods by 2 days at start and end
    start = min(visit_date) - days(2),
    end   = max(visit_date) + days(2),
    .groups = "drop"
  ) %>%
  rowwise() %>% 
  mutate(visit_date = list(seq.Date(start, end, by = "day"))) %>%  # Create all dates in each expanded period
  unnest(visit_date) %>%
  select(primary_period, visit_date)

# Select VES dates
ves_dates <- survey_20135 %>%
  filter(
    survey_type == "visual",
    visit_date > min(reintroduction_20135$release_date + days(5)), # Remove VES surveys from before or soon after frogs were reintroduced
    visit_date <= max(cmr_dates$visit_date)
  ) %>% 
  select(visit_date) %>% 
  arrange(visit_date)

# Identify all visual surveys that fall outside of CMR primary periods (expanded)
unmatched_ves <- ves_dates %>% 
  anti_join(cmr_dates, by = "visit_date") %>% 
  distinct(visit_date)
```

- If no adult frogs found during unmatched VES, consider including as CMR survey.
- No unmatched VES detected. 

### Remove unneeded objects

```{r}
rm(cmr_survey_check, cmr_dates, ves_dates, unmatched_ves)
```

### Remove visual surveys from survey records

```{r remove_visual_surveys}
survey_20135 <- survey_20135 %>% 
  filter(survey_type == "cmr", 
         visit_date > min(reintroduction_20135$release_date) + days(5)) %>% # Remove CMR surveys from before or soon after frogs were reintroduced
  rename(survey_date = visit_date)
```

- Review records for potential problematic surveys, for example, based on comment fields and fields describing current conditions.

### Select unique survey dates

```{r unique_survey_dates}
survey_20135_mrmr <- survey_20135 %>% 
  distinct(survey_date, survey_type, visit_status)
```

### Survey data checks

```{r survey_data_checks}
# Check for duplicate survey dates
assert_that(!any(survey_20135_mrmr %>% count(survey_date) %>% pull(n) > 1))

# Check for sites that were unsuitable for survey
assert_that(all(survey_20135_mrmr$visit_status == "suitable"))

# Check for non-cmr and non_reintroduction survey types
assert_that(all(survey_20135_mrmr$survey_type == "cmr"))
```

### To meet mrmr format requirement, add reintroduction dates as surveys

```{r add_reintroductions}
survey_20135_mrmr <- reintroduction_20135_mrmr %>% 
  distinct(release_date, type) %>% 
  rename(survey_type = type, 
         survey_date = release_date) %>% 
  bind_rows(survey_20135_mrmr) %>% 
  mutate(site_id = 20135) %>%  # assign all surveys to primary site
  relocate(site_id, .before = survey_date) %>% 
  arrange(survey_date) 
```

### Assign primary and secondary periods

```{r assign_cmr_periods}
survey_20135_mrmr <- survey_20135_mrmr %>%
  arrange(survey_date) %>% 
  # Identify primary periods using breaks between CMR surveys (> 4 day gap) - adjust gap as necessary
  mutate(primary_period = cumsum(survey_date - lag(survey_date, default = first(survey_date)) > 4) + 1) %>%   
  # Assign secondary periods within each primary period
  group_by(primary_period) %>%
  mutate(secondary_period = row_number()) %>%
  ungroup() %>% 
  mutate(secondary_period = replace(secondary_period, survey_type == "reintroduction", 0)) %>% 
  select(-visit_status)
```

### Save cleaned survey file

```{r save_cleaned_survey_data}
survey_20135_mrmr %>% write_csv(here("data", "clean", "20135_survey.csv"))
```

## Clean capture dataset

### Capture data checks

```{r capture_data_checks, message=FALSE}
# Verify that all dates in capture are present in survey 
assert_that(nrow(capture_20135 %>% anti_join(survey_20135_mrmr, by = join_by(visit_date == survey_date))) == 0)

# Verify that visit_status is "suitable" for all records
assert_that(all(capture_20135$visit_status == "suitable"))

# Verify that survey_type is "cmr" for all records
assert_that(all(capture_20135$survey_type == "cmr"))

# Check for records where pit_tag_ref is NULL
assert_that(noNA(capture_20135$pit_tag_ref))

# Check for records containing pit tags shorter than 15 characters (or 9 for early sites)
assert_that(!any(nchar(capture_20135$pit_tag_ref) != 15))

# Check for records containing pit tags that end with ".1" (known to be potentially incorrect) 
assert_that(!any(str_detect(capture_20135$pit_tag_ref, "\\.1$")))

# Check for records where species != ramu
assert_that(all(capture_20135$species == "ramu"))

# Check for records where capture_animal_state is NULL
assert_that(noNA(capture_20135$capture_animal_state))

# Check for records where capture_animal_state == "dead"
assert_that(!any(capture_20135$capture_animal_state == "dead"))

# Check for records of subadult frogs 
assert_that(all(capture_20135$length >= 40)) # some captured frogs are < 40 mm, but all were retained

# Check for duplicate pit tags on each date
assert_that(nrow(capture_20135 %>% count(visit_date, pit_tag_ref) %>% filter(n > 1)) == 0)
```

### Clean capture data
```{r clean_capture_data}
# Drop dead frogs
capture_20135 <- capture_20135 %>% 
  filter(capture_animal_state != "dead")
```

### Create final capture dataset formatted for mrmr
```{r final_capture_data}
capture_20135_mrmr <- capture_20135 %>%
  rename(survey_date = visit_date, 
         pit_tag_id = pit_tag_ref) %>%
  mutate(site_id == 20135) %>%  # assign all captures to primary site
  select(site_id, survey_date, pit_tag_id, bd_load) %>% 
  arrange(survey_date, pit_tag_id)
```

### Save cleaned capture data (one file with Bd load and one without)
```{r save_cleaned_capture_data}
capture_20135_mrmr %>% write_csv(here("data", "clean", "20135_bd_capture.csv"))
capture_20135_mrmr %>% select(-bd_load) %>% write_csv(here("data", "clean", "20135_capture.csv"))
```








