---
title: "R Notebook"
output: html_notebook
---
Analysis of Ranger Station Pond CMR dataset - Tyndall area
* use 'treatment' covariate in translocation file to evaluate survival of Bd-exposed and unexposed frogs

Install mrmr package if current version is not available
```{r}
remotes::install_github("SNARL1/mrmr")
```

Load required packages
```{r}
library(dplyr)
library(readr)
library(tibble)
library(rstan)
library(mrmr)
```

Set working directory for Ranger Station Pond
```{r}
knitr::opts_knit$set(root.dir = '/media/rknapp/Data/Box Sync/SNARL shared folder/Projects/cmr_analysis/sites/tyndall/ranger_station')
getwd()
```

Create required data frames
```{r}
captures <- "captures.rangerstapond.mrmr.csv" %>% read_csv
translocations <- "translocations.rangerstapond.mrmr.csv" %>% read_csv
surveys <- "surveys.rangerstapond.mrmr.csv" %>% read_csv
```

Read and clean the data
```{r}
data_rangerstapond <- clean_data(captures, surveys, translocations,
                                survival_formula = ~ treatment,
                                survival_fill_value = c(treatment = "unexposed"))
```

Fit model
```{r}
# use 'chains = 1, iter = 10' for initial model check
# use 'chains = 3, iter = ~1000' for actual model run
model_rangerstapond <- fit_model(data_rangerstapond, cores = parallel::detectCores())
```

To diagnose potential non-convergence of the MCMC algorithm, inspect traceplots, Rhat estimates (any >= 1.01?)
```{r}
traceplot(model_rangerstapond$m_fit, par = "Nsuper")

pars_to_plot <- c('alpha_lambda', 
                  'sigma_lambda', 
                  'beta_phi', 
                  'sigma_phi', 
                  'beta_detect')
traceplot(model_rangerstapond$m_fit, pars = pars_to_plot)

print(model_rangerstapond$m_fit, pars = pars_to_plot)
```

Visualize results
```{r}
plot_model(model_rangerstapond, what = 'abundance')
plot_model(model_rangerstapond, what = 'recruitment')
plot_model(model_rangerstapond, what = 'survival')
```

Compare posterior estimates for survival between Bd-exposed and unexposed 
individudals. 
```{r}
post <- rstan::extract(model_rangerstapond$m_fit)

# Check that all translocation treatments match
t1 <- data_rangerstapond$survival_covariate_df %>%
  filter(pit_tag_id %in% translocations$pit_tag_id) %>%
  arrange(pit_tag_id)
t2 <- distinct(translocations, pit_tag_id, treatment) %>%
  mutate(pit_tag_id = as.character(pit_tag_id)) %>%
  arrange(pit_tag_id)
stopifnot(all_equal(t1$treatment, t2$treatment))


survival_df <- post$phi %>%
  reshape2::melt(varnames = c("iter", "individual", "primary_period")) %>%
  as_tibble %>%
  mutate(pit_tag_id = data_rangerstapond$survival_covariate_df$pit_tag_id[individual]) %>%
  left_join(data_rangerstapond$survival_covariate_df) %>%
  distinct(iter, primary_period, value, treatment) %>%
  left_join(data_rangerstapond$surveys %>% filter(secondary_period < 2))

survival_df %>%
  ggplot(aes(x = value, color = treatment)) + 
  stat_density(geom = "line", position = position_dodge(width=0)) + 
  facet_wrap(~ survey_date, scales = 'free_y') + 
  xlab("Survival probability") + 
  ylab("Posterior density") + 
  scale_color_discrete("Treatment")

paste("Posterior probability that treatment increased survival:", 
      mean(post$beta_phi[, 2] < 0))
```

Plot cohort survival by treatment group, with 95% credible intervals and 
lines/points on the posterior median.
```{r}
primary_period_dates <- model_rangerstapond$data$surveys %>%
  group_by(.data$primary_period) %>%
  summarize(date = min(.data$survey_date),
            year = min(.data$year)) %>%
  ungroup

model_rangerstapond$post$s %>%
  reshape2::melt(varnames = c('iter', 'i', 'primary_period')) %>%
  as_tibble %>%
  mutate(pit_tag_id = dimnames(model_rangerstapond$data$stan_d$Y)[[1]][.data$i]) %>%
  filter(.data$pit_tag_id %in% as.character(model_rangerstapond$data$translocations$pit_tag_id)) %>%
  left_join(distinct(model_rangerstapond$data$translocations,
                     .data$pit_tag_id, .data$release_date)) %>%
  left_join(model_rangerstapond$data$survival_covariate_df) %>%
  group_by(.data$release_date, .data$primary_period, .data$iter, .data$treatment) %>%
  summarize(fraction_alive = mean(.data$value == 2)) %>%
  ungroup %>%
  filter(.data$primary_period > 1) %>%
  left_join(primary_period_dates) %>%
  filter(.data$date >= .data$release_date) %>%
  group_by(.data$release_date, .data$date, .data$treatment) %>%
  summarize(lo = quantile(.data$fraction_alive, .025),
            med = median(.data$fraction_alive),
            hi = quantile(.data$fraction_alive, .975)) %>%
  ungroup %>%
  mutate(grp = paste(treatment, release_date), 
         release_date = paste("Released on", release_date)) %>%
  ggplot(aes(.data$date, .data$med,
             color = tools::toTitleCase(.data$treatment),
             fill = tools::toTitleCase(.data$treatment), 
             group = .data$grp)) +
  geom_ribbon(aes(ymin = .data$lo, ymax = .data$hi),
              alpha = .4, color = NA) +
  geom_line() +
  geom_point() +
  ylab('Fraction surviving') +
  xlab('Date') +
  scale_color_brewer('Treatment', palette = "Dark2") +
  scale_fill_brewer('Treatment', palette = "Dark2") +
  scale_x_date(date_breaks = "1 year") + 
  facet_wrap(~release_date, ncol = 1)
```


Save model objects
```{r}
write_rds(model_rangerstapond, "model_rangerstapond_mrmr.rds")

# Or compressed
write_rds(model_rangerstapond, "model_rangerstapond_mrmr_compressed.rds", "xz", compression = 9L)
```

Read model objects back in
```{r}
model_rangerstapond <- read_rds("model_rangerstapond_mrmr_compressed.rds")
```

