---
title: "20276 - Create datasets for analysis using *mrmr* WITH reinfection-reintroduction data"
author: "Thomas C Smith"
output: html_notebook
editor_options: 
  chunk_output_type: inline
---
```{r load-packages}
library(tidyverse)
library(RPostgreSQL)
library(lubridate)
library(assertthat)
library(DBI)
library(RPostgres)
library(here)
# options(readr.show_col_types = FALSE)
```

### Connect to database

- When connecting via SSH, open external connection to database, then replace "Sys.getenv" with "rstudioapi::askForPassword" for host and port, run chunk.   
- Users without credentials can use CSV files created in code chunk "save-raw-data-files" below as a starting point.  

```{r db-connect-set-schemas}
pg = dbDriver("Postgres")
con_amphibians_v2 = dbConnect(pg, 
                user = rstudioapi::askForPassword("user name"), 
                password = rstudioapi::askForPassword("password"),
                host = Sys.getenv("host"), 
                port = Sys.getenv("port"), 
                dbname = Sys.getenv("dbname")  
                # options = "-c search_path=field_surveys" # add this to access the schema for all tables; versus setting schema as each table is accessed and assigned
                )
DBI::dbExecute(con_amphibians_v2, "SET search_path TO field_surveys, qpcr") # now the schemas are set, in order of query priority
DBI::dbListTables(con_amphibians_v2) # to test the connection and schemas


```


# Create Data overview for 20276 in 2026.

For the site 20276, this notebook creates the survey, capture, and reintroduction data files for use in analyzing capture-mark-recapture data using the *mrmr* package. The code retrieves data from the PostgreSQL amphibians database, runs error checks, and creates the relevant files in *mrmr* format. The code is written in distinct chunks to allow review of results from one chunk before running the next. For example, review of the raw data files is necessary to identify potentially problematic records for review and possible removal from the data set. This code was built from a template, and all instances of "xxxxx" in the template were replaced with the 5-digit site_id of interest: 20276.

For the initial evaluation of CMR at 20276, we consider two pieces of context:
1) Frogs released into 20276 were exposed to Bd at the Oakland Zoo in years 2022 and 2023, prior to being cleared of fungal infections and prior to release/reintroduction. This was done with the goal of prompting an immune response, as is described in the work recorded in the GitHub repository: https://github.com/mqwilber/reinfection_revisited. 

2) As of spring 2026, frogs are released into 20276 and nearby site 20275. However at this time, the two sites seem sufficiently far apart and no evidence of frog emigration/dispersal is reported, so they are initially analyzed as separate sites. We have learned from other basins (e.g. 10102, 10101, 10100; see Knapp et al 2022 PeerJ) that even small amounts of frog migration impact that and require adjacent sites to be lumped. Until that is observed at 20276, we proceed treating it as an isolated site.

> Note for credentialed users: Data from the most recent field season may not yet be appended to tables in the amphibians database. If these recently-collected data are necessary (e.g., for preliminary analyses), complete datasets can be created by copying code chunks in "fulcrum_data_append.Rmd" into this notebook and executing all code chunks.

As of April 2026, the 2025 data are not appended to the database, and are added to the workspace before the database is accessed for. 
## Create current year data.
### Load current year data from local files.
Loading visits, surveys, captures, and Bd data from local sources and/or qpcr database allows us to build the data as needed prior to this year's data append.

```{r create_current_year_append_tables_from_local_source, eval=FALSE}
# visit
visit_20276_current <- read_csv("P:/MLRG/code/ErrorChecking/2025/cleaned_data/visit_final_251209.csv") %>% 
  rename(visit_comment = comment, 
         visit_id = id)

# survey
survey_20276_current <- read_csv("P:/MLRG/code/ErrorChecking/2025/cleaned_data/survey_final_251209.csv") %>% 
  rename(survey_id = id, 
         survey_comment = comment ) %>% 
  select(survey_id, visit_id, survey_type, description, survey_quality, wind, sky, air_temp, duration, survey_comment)

# capture
capture_20276_current <- read_csv("P:/MLRG/code/ErrorChecking/2025/cleaned_data/capture_final_daynight_251209.csv") %>% 
  mutate(pit_tag_ref = as.character(pit_tag_ref)) %>% 
  rename(capture_survey_id = id,
         capture_comment = comment, 
         pit_tag_id = pit_tag_ref) %>% 
  select(capture_survey_id, survey_id, pit_tag_id, tag_new, species, capture_life_stage, capture_animal_state, length, weight, capture_comment, swab_id)
```

### Retrieve current year Bd loads if relevant
Possible to use either the local files [location unknown] or to use the qPCR database and connect to the current year capture records via swab_id = sample_id
```{sql retrieve-capture-data, connection=con_amphibians_v2, output.var = "bd_loads_only_all" }
select 
  bd_load.sample_id AS swab_id,
  bd_load.bd_load
from bd_load
```

### Put current year data in long-term format to match survey data and capture histories
```{r joins-current-year-data-objects}
# visit_20276_current
# survey_20276_current
# capture_20276_current
# bd_loads_only_all

# survey
survey_20276_2025 <- visit_20276_current %>% 
  filter(site_id %in% c( 20276 , 20276, 23105, 23104, 23110, 21386, 20277, 20275) ) %>% 
  inner_join( survey_20276_current , by = "visit_id" ) %>% 
  filter(survey_type == "cmr") %>% 
  select(site_id, 
    visit_date, 
    visit_status, 
    survey_type, 
    description, 
    survey_quality,
    wind,
    sky,
    air_temp,
    duration,
    visit_comment,
    survey_comment)
  
# capture
capture_20276_2025 <- visit_20276_current %>% 
  filter(site_id %in% c( 20276 , 20276, 23105, 23104, 23110, 21386, 20277, 20275) ) %>% 
  inner_join( survey_20276_current , by = "visit_id" )  %>% 
  filter( survey_type=="cmr" ) %>%
  # filter( survey_type %in% c("cmr", "night") ) %>% # included for exploratory purposes, night captures net two additional individuals
  inner_join( capture_20276_current, by = "survey_id" ) %>% 
  select(site_id, 
    visit_date,
    visit_status, 
    survey_type, 
    pit_tag_id, 
    tag_new, 
    species, 
    capture_life_stage, 
    capture_animal_state, 
    length,
    weight,
    visit_comment,
    survey_comment,
    capture_survey_comment=capture_comment,
    visit_id,
    survey_id,
    capture_survey_id,
    swab_id
  ) %>% 
  left_join(bd_loads_only_all, by = "swab_id" )

```



## Retriev long-term data from PostgreSQL database
### Retrieve reintroduction data from database

```{sql retrieve_reintroduction_data, connection=con_amphibians_v2, output.var = "reintroduction_20276"}

select 
  visit.site_id, 
  release.release_type AS type,
  visit.visit_date AS release_date,
  release_frog.pit_tag_ref AS pit_tag_id, 
  release_frog.tag_new, 
  release_frog.sex, 
  release_frog.length, 
  release_frog.weight, 
  release_frog.swab_id, 
  release_frog.bd_exposure
from release_frog
INNER JOIN release ON release_frog.release_id = release.release_id
INNER JOIN visit ON release.visit_id = visit.visit_id
WHERE visit.site_id = 20276 
```

### Retrieve CMR survey data from database

- If some relevant surveys are classified as "ves", include in "where" statement. 
- Note that in come cases, 20276 received nighttime survey effort. As of April 2026, those night time surveys are categorized as "survey_type == swab". 
- Initially night captures are not included in the data due to difference in detectability; thus night surveys are not included either. For future reference, at 20276, those night surveys occurred on 2022-08-25, 2023-08-26, 2024-07-16, 2024-07-15, as indicated by the survey description or visit comments or survey comments. Per this design, to include night surveys in queries, include the survey_type='swab' in the `where` expression.

```{sql retrieve-survey-data, connection=con_amphibians_v2, output.var = "survey_20276"}
select 
  site_id, 
  visit_date, 
  visit_status, 
  survey_type, 
  description, 
  survey_quality,
  wind,
  sky,
  air_temp,
  duration,
  visit_comment,
  survey_comment
from visit
inner join survey on visit.visit_id = survey.visit_id
where site_id in (20276, 23105, 23104, 23110, 21386, 20277, 20275) AND 
      survey_type in ('cmr', 'visual') AND
      visit_date > '2021-01-01'
```
```{r}
view(survey_20276)
```

#### Add current year surveys if relevant

```{r current-year-surveys-bind-2025}

survey_20276 <- survey_20276 %>% 
  bind_rows( survey_20276_2025 %>% 
               select(site_id, 
                      visit_date, 
                      visit_status,
                      survey_type,
                      description, 
                      survey_quality,
                      wind,
                      sky,
                      air_temp,
                      duration,
                      visit_comment,
                      survey_comment
                      )
             )

```

### Retrieve frog capture data from database

Again, as stated above, for site 20276 as of April 2026, some frogs were captured in night-time surveys. So long as those captures are associated with survey_type='swab', the where expression that does/does not include 'swab' will/will not include those individuals. As of April 2026, this query does not include the night time captures.

```{sql retrieve-capture-data, connection=con_amphibians_v2, output.var = "capture_20276"}
select 
  visit.site_id, 
  visit.visit_date, 
  visit.visit_status, 
  survey.survey_type, 
  capture_survey.pit_tag_ref AS pit_tag_id, 
  capture_survey.tag_new, 
  capture_survey.species, 
  capture_survey.capture_life_stage, 
  capture_survey.capture_animal_state, 
  capture_survey.length,
  capture_survey.weight,
  visit.visit_comment,
  survey.survey_comment,
  capture_survey.capture_survey_comment,
  visit.visit_id,
  survey.survey_id,
  capture_survey.capture_survey_id,
  capture_survey.swab_id,
  bd_load.bd_load
from visit
inner join survey on visit.visit_id = survey.visit_id
inner join capture_survey on survey.survey_id = capture_survey.survey_id
left join bd_load on bd_load.sample_id = capture_survey.swab_id
where site_id in (20276, 23105, 23104, 23110, 21386, 20277, 20275) and
      survey_type in ('cmr','swab')
```
```{r}
view(capture_20276)
```

#### Add current year captures if relevant

```{r current-year-surveys-bind-2025}


```

```{r db-disconnect3}
dbDisconnect(con)
rm(con)
```

### Save raw data files
```{r save-raw-data-files, message=FALSE}
reintroduction_20276 %>% write_csv(here::here("data", "raw", "20276_reintroduction_raw.csv")) 
survey_20276 %>% write_csv(here::here("data", "raw", "20276_survey_raw.csv"))
capture_20276 %>% write_csv(here::here("data", "raw", "20276_capture_raw.csv"))
```

If data are created and up to date, read saved files back in to your session using this example code:
> translocation_xxxxx <- read_csv(here::here("data", "raw", "xxxxx_translocation_raw.csv")) %>% 
  mutate(pit_tag_ref = as.character(pit_tag_ref))

## Create reintroduction dataset
### Retain only relevant columns

```{r reintroduction-relevant-columns}
reintroduction_20276 <- reintroduction_20276 %>% 
  select(site_id, release_date, type, pit_tag_ref, bd_exposure)
```

### Data checks on reintroductions, surveys, and captures.

- Run checks before and after data cleaning (next code chunk) to ensure that all issues have been resolved.  
- Run lines of code one at a time to allow easy evaluation of messages and outputs.  

```{r reintroduction-data-checks, message=FALSE}
# Check for records where pit_tag_ref is NULL
assert_that(noNA(reintroduction_20276$pit_tag_id))

# Check for records containing pit tags shorter than 15 characters
assert_that(!any(nchar(reintroduction_20276$pit_tag_id) != 15))

# Check for records containing pit tags that end with ".1" (known to be potentially incorrect) 
assert_that(!any(str_detect(reintroduction_20276$pit_tag_id, "\\.1$")))

# Check for records of subadult frogs
reintroduction_20276 %>% drop_na(length) %>% filter(length < 40)

# Check for duplicate pit tags on each date
reintroduction_20276 %>% count(release_date, pit_tag_id) %>% filter(n > 1)

# Are release dates correct?
reintroduction_20276 %>% distinct(release_date) %>% glimpse()
```
Note the release dates for site 20276 are:
2022-07-21
2023-08-28
2024-07-19
2025-08-05

### Clean reintroduction data
Passed all checks. 

### Create final translocation dataset formatted for mrmr

```{r final-reintroduction-data}
reintroduction_20276 <- reintroduction_20276 %>% 
  arrange(release_date, pit_tag_id) %>% 
  select(site_id, type, release_date, pit_tag_id, bd_exposure)
```

### Save cleaned translocation data

```{r save-cleaned-translocation-data, message=FALSE}
reintroduction_20276 %>% write_csv(here::here("data", "clean", "20276_reintroduction.csv"))
```

- When reading dataset back in, be sure to format pit_tag_id as character.  

## Create survey dataset

### Include only relevant surveys

In 2026, the relevant surveys for this site are in 20276, but worth looking for surveys in 23105, 23104, 23110, 21386, 20277, 20275.

#### Primary site surveyed during each secondary period?

- If yes, code chunk will return zero records. Exclude other records as necessary. 
- In filter statement, edit site_id as necessary.

```{r check-survey-history}
survey_20276 %>% 
  select(site_id, visit_date, survey_type, visit_status) %>% 
  mutate(visit_status = case_when( visit_status=="suitable" ~ 1, 
                                   .default=0)) %>% 
  unique() %>% 
  pivot_wider(id_cols = c(visit_date, survey_type), names_from = site_id, values_from = visit_status  )
  
```

#### Include only surveys at primary site

- Run this code chunk as is when only CMR surveys need to be included. If VES surveys also need to be included, modify code accordingly. 

```{r include-relevant-surveys}
survey_20276 <- survey_20276 %>% 
  filter(site_id == 20276, 
         survey_type == 'cmr' ) %>%
  rename(survey_date = visit_date) %>% 
  arrange(survey_date)
```

### Add reintroduction dates as surveys

- Skip if reintroductions are not relevant.  

```{r add-translocation-surveys}
survey_20276 <- reintroduction_20276 %>% 
  distinct(site_id, release_date, type) %>% 
  rename(survey_type = type, survey_date = release_date) %>% 
  bind_rows(survey_20276) %>% 
  arrange(survey_date) 
```




### Survey data checks

- Run checks before and after data cleaning (next code chunk) to ensure that all issues have been resolved.  
- Run lines of code one at a time to allow easy evaluation of messages and outputs.  

```{r survey-data-checks}
# Check for duplicate survey dates
survey_xxxxx %>% count(site_id, survey_date) %>% filter(n > 1)

# Check for sites that were unsuitable for survey
survey_xxxxx %>% filter(visit_status != "suitable")
```

### Clean survey data

- Edit data-cleaning code as necessary. 

```{r clean-survey-data}
# Drop duplicate surveys
survey_20276 <- survey_20276 %>% 
  distinct(survey_date, .keep_all = TRUE)
```



### Create primary and secondary period columns in mrmr format

- Edit case_when steps to reflect actual translocation/reintroduction and survey history.  

#### For robust design dataset

```{r create-period-columns1}
survey_20276 <- survey_20276 %>%
  select(site_id, survey_date, survey_type) %>% 
  mutate(primary_period = case_when(survey_date == "2022-07-21" ~ 1,
                                    between(survey_date, ymd("2022-08-25"), ymd("2022-08-26")) ~ 2,
                                    between(survey_date, ymd("2023-08-25"), ymd("2023-08-27")) ~ 3,
                                    survey_date == "2023-08-28" ~ 4,
                                    between(survey_date, ymd("2024-07-15"), ymd("2024-07-17")) ~ 5,
                                    survey_date == "2024-07-19" ~ 6
                                    ) )
                                    
survey_20276 <- survey_20276 %>% 
  arrange(primary_period, survey_date) %>% 
  group_by(primary_period) %>% 
  mutate(secondary_period = seq_len(n()),
         secondary_period = replace(secondary_period, survey_type != "cmr", 0)) 
```

### Save cleaned survey file

```{r save-cleaned-survey-data}
survey_20276 %>% write_csv(here::here("data", "clean", "20276_survey.csv"))
```

## Create capture dataset

### Retain only relevant columns

```{r capture-relevant-columns}
capture_xxxxx <- capture_20276 %>% 
  select(-visit_id, -survey_id, -capture_id)
```

### Capture data checks

- Run checks before and after data cleaning (next code chunk) to ensure that all issues have been resolved.  
- Run lines of code one at a time to allow easy evaluation of messages and outputs.  

```{r capture-data-checks, message=FALSE}
# Check for records where pit_tag_ref is NULL
assert_that(noNA(capture_20276$pit_tag_ref))

# Check for records containing pit tags shorter than 15 characters (or 9 for early sites)
assert_that(!any(nchar(capture_20276$pit_tag_ref) != 15))

# Check for records containing pit tags that end with ".1" (known to be potentially incorrect) 
assert_that(!any(str_detect(capture_20276$pit_tag_ref, "\\.1$")))

# Check for records where species != ramu
capture_20276 %>% drop_na(species) %>% filter(species != "ramu")

# Check for records were survey_type = swab
### for 20276, this would be approached differently if NIGHT captures are included because they are coded as swab surveys as of April 2026
assert_that(!any(capture_20276$survey_type == "swab"))

# Check for records where capture_animal_state is NULL
assert_that(noNA(capture_20276$capture_animal_state))

# Check for records where capture_animal_state == "dead"
assert_that(!any(capture_20276$capture_animal_state == "dead"))
### Which PIT IDs and dates for dead individuals?
capture_20276 %>% 
  filter( capture_animal_state == "dead") %>% 
  select( visit_date, pit_tag_id)

# Check for records of subadult frogs
capture_20276 %>% drop_na(length) %>% filter(length < 40)

# Check for sites that were unsuitable for survey
assert_that(!any(capture_20276$visit_status != "suitable"))

# Check for duplicate capture records on each date
capture_20276 %>% count(visit_date, pit_tag_id) %>% filter(n > 1)

# Check for captures that lack associated surveys
survey_xxxxx %>% 
  select(survey_date, primary_period) %>% 
  rename(visit_date = survey_date) %>% 
  right_join(capture_xxxxx, by = c("visit_date")) %>% 
  filter(is.na(primary_period))

# Simpler alternative
setdiff(capture_20276$visit_date, survey_20276$survey_date) %>% as.Date(origin = "1970-01-01")

# Check for surveys that lack associated captures
survey_20276 %>% 
  filter(survey_type == "cmr") %>% 
  select(survey_date, primary_period) %>% 
  rename(visit_date = survey_date) %>% 
  left_join(capture_20276, by = c("visit_date")) %>% 
  filter(is.na(species))

# Simpler alternative
setdiff(survey_20276$survey_date, capture_20276$visit_date) %>% as.Date(origin = "1970-01-01")
```
#### Dead frogs.
Note that these individuals are recorded as "dead" on these dates:
2022-08-25	982091068563860			
2023-08-26	982091068034956			
2023-08-25	982091068562931			
2024-07-15	982091068554611			
2024-07-15	982091068563346
2024-07-15	982091068559959			

retain capture_animal_state in the subsequent data to preserve this record of dead frogs.

#### Surveys with no captures.
Note that the following dates had no frog captures
2022-07-21 - reintroduction day, not a CMR
2022-08-26 - day two of a CMR primary period, there were captures on day 1.
2023-08-28 - reintroduction day, not a CMR
2024-07-19 - reintroduction day, not a CMR

### Clean capture data

In some contexts we may wish to retain or drop the dead animals from analysis, but as of April 2026 they remain in the capture data to allow Mark/analysts to assess and remove them; in my opinion this is at a mimimum useful anecdotal information if we have a previous capture of the individual and know their preceding Bd loads.
```{r clean-capture-data}
capture_xxxxx <- capture_xxxxx %>% 
  filter(capture_animal_state != "dead")
```

### Create final capture dataset formatted for mrmr
For user information in the context of disease, I have left the capture animal state and capture comments in this initial version of the capture data. These are easily removed just before analysis, but could inform Bd loads or individual survival.
```{r final-capture-data}
capture_20276 <- capture_20276 %>%
  rename(survey_date = visit_date) %>%
  arrange(pit_tag_id, survey_date ) %>%
  select(site_id, survey_date, pit_tag_id, swab_id, bd_load, capture_animal_state, capture_survey_comment)
```

### Save cleaned capture data

```{r save-cleaned-capture-data}
capture_20276 %>% write_csv(here::here("data", "clean", "20276_capture.csv"))
```










# Night captures for 20276
###
Because at 20276, some frogs were captured in night time cmr-like surveys but are not included in the cmr capture data. In the database, frogs captured at night at 20276 were categorized as survey type = 'swab', but frogs captured at night do not appear in the database for previous survey years. Years 2022-2023 night captures are only in the pcloud fulcrum data or annual error-checking output. For now, the night surveys will not be written to file and will need additional work.

```{sql retrieve-capture-data, connection=con_amphibians_v2, output.var = "night_capture_20276"}
select 
  visit.site_id, 
  visit.visit_date, 
  visit.visit_status, 
  survey.survey_type, 
  capture_survey.pit_tag_ref as pit_tag_id, 
  capture_survey.tag_new, 
  capture_survey.species, 
  capture_survey.capture_life_stage, 
  capture_survey.capture_animal_state, 
  capture_survey.length,
  capture_survey.weight,
  visit.visit_comment,
  survey.survey_comment,
  capture_survey.capture_survey_comment,
  visit.visit_id,
  survey.survey_id,
  capture_survey.capture_survey_id,
  capture_survey.swab_id,
  bd_load.bd_load
from visit
inner join survey on visit.visit_id = survey.visit_id
inner join capture_survey on survey.survey_id = capture_survey.survey_id
left join bd_load on bd_load.sample_id = capture_survey.swab_id
where site_id in (20276, 23105, 23104, 23110, 21386, 20277, 20275) and
      visit_date > '2021-01-01' and
      survey_type = 'swab' and
      species = 'ramu'
```
```{r}
view(night_capture_20276)

night_capture_20276 %>% 
  select(pit_tag_ref) %>% 
  unique()
```

2024 night captures nets 16 additional captures of 16 individuals; more years need to be added. Visual inspection of the smaller set of night capture data suggests that night capture data is more or less in fine shape. From here, save it separately, and inspect a 'bound' version of day+night to preview what Mark could create quickly after pulling the two separate files.
```{r final-capture-data}
night_capture_20276 <- night_capture_20276 %>%
  rename(survey_date = visit_date) %>%
  arrange(pit_tag_id, survey_date ) %>%
  select(site_id, survey_date, pit_tag_id, swab_id, bd_load, capture_animal_state, capture_survey_comment)
```

```{r preview-day-night-bindrows  }
capture_20276 %>% 
  bind_rows(night_capture_20276)
```
This yields 68 capture events of 46 individual frogs.

<!--  Next steps:

create zoo exposure load trajectories
- this is somewhat of a separate project and probably a separate script and effort. 
- But it should be straight forward for Rocky and for Laurel because of the orderliness and isolation of oakland zoo swab data.

--> 