---
title: "22008 - Create datasets for analysis using *mrmr* - 2 treatment groups"
author: "Roland Knapp"
format: 
  html:
editor_options: 
  chunk_output_type: inline
---

This notebook creates the survey, capture, and translocation files for use in analyzing capture-mark-recapture data using the *mrmr* package. The code retrieves data from the PostgreSQL amphibians database, runs error checks, and creates the relevant files in *mrmr* format. A treatment column is included in the reintroduction file (exposed, unexposed). Wild frogs are excluded from the capture file. 

Users without credentials to access the database will not be able to create the raw data files, but can use the saved clean data files for analyses. See cmr-analysis/README.md for details. 

Note that data from the most recent field season may not yet be appended to tables in the amphibians database. If these recently-collected data are necessary (e.g., for preliminary analyses), complete datasets can be created by copying code chunks in "code\\fulcrum_data_append.Rmd" into this notebook and executing all code chunks. 

## Setup

```{r}
#| eval: true
#| output: false
#| code-fold: true
#| code-summary: "Packages"

# Install packages from remote - if necessary
remotes::install_github("SNARL1/pgutils")

# Load installed packages
if (!require(librarian)){
  install.packages("librarian")
  library(librarian)
}
librarian::shelf(here, readr, dplyr, tidyr, stringr, magrittr, lubridate, assertthat)
```

## Connect to database

Before running this code chunk, establish connection to database server (via VPN or SSH)

```{r}
#| eval: true
#| output: false
#| code-fold: true
#| code-summary: "Packages"

con <- pgutils::connect_to_pg()
```

## Retrieve data subsets from PostgreSQL database

- Before retrieving data, view table in Connections pane to determine presence of columns formatted as "pq_xxxx". If any exist, add code after "collect()" to format as character.

### Reintroduction data

```{r reintroduction_data}
relocate <- tbl(con, "relocate") %>%
  rename(
    relocate_comment = comment,
    relocate_id = id
  )

relocate_frog <- tbl(con, "relocate_frog") %>%
  rename(
    relocate_frog_comment = comment,
    relocate_frog_id = id
  )

reintroduction_22008 <- relocate %>%
  inner_join(relocate_frog, by = "relocate_id") %>%
  filter(
    release_siteid1 %in% c(22008, 22019)
  ) %>%
  select(
    release_siteid1,
    release_date,
    type,
    pit_tag_ref,
    tag_new,
    length,
    weight,
    swab_id,
    bd_exposure,
    relocate_comment,
    relocate_frog_comment
  ) %>%
  collect()
```

Review the reintroduction data subset for potential problems. 
- tag_new field is empty. 

### Survey data

To allow subsequent checks of whether surveys at which no frogs were captured are recorded as "visual" instead of "cmr", include visual surveys

```{r survey_data}
visit <- tbl(con, "visit") %>%
  rename(
    visit_comment = comment,
    visit_id = id
  )

survey <- tbl(con, "survey") %>%
  rename(
    survey_comment = comment,
    survey_id = id
  )

survey_22008 <- visit %>% 
  inner_join(survey, by = "visit_id") %>% 
  filter(
    site_id %in% c(22008, 22019),
    survey_type %in% c("cmr", "visual")
  ) %>% 
select( 
  site_id, 
  visit_date, 
  visit_status, 
  survey_type, 
  description, 
  survey_quality,
  wind,
  sky,
  duration,
  visit_comment,
  survey_comment
) %>% 
collect()

# convert ENUM columns to character
survey_22008 <- pgutils::convert_pg_enums(survey_22008) 
  
```

Review the survey data subset for potential problems. 
- Some surveys with inclement weather, likely worth retaining all surveys regardless. 

### Capture data

```{r capture_data}
visit <- tbl(con, "visit") %>%
  rename(
    visit_id = id
  )

survey <- tbl(con, "survey") %>%
  rename(
    survey_id = id
  )

capture <- tbl(con, "capture_survey") %>% 
  rename(
    capture_survey_comment = comment
  )

bdload <- tbl(con, "bd_load") %>% 
  rename(
    swab_id = sample_id
  )

capture_22008 <- visit %>% 
  inner_join(survey, by = "visit_id") %>% 
  inner_join(capture, by = "survey_id") %>% 
  left_join(bdload, by = "swab_id") %>% 
  filter(
    site_id %in% c(22008, 22019),
    survey_type == "cmr"
  ) %>% 
select( 
  site_id, 
  visit_date, 
  visit_status, 
  survey_type, 
  pit_tag_ref, 
  tag_new, 
  species, 
  capture_life_stage, 
  capture_animal_state, 
  length,
  swab_id,
  replicate,
  std_type,
  dilution,
  bd_load,
  capture_survey_comment
) %>% 
collect()

capture_22008 <- pgutils::convert_pg_enums(capture_22008) 
```

Review the capture data subset for potential problems. 
- qPCR used gblock and plasmid standards, all samples run singly.

## Save raw data files

```{r save_raw_data, message=FALSE}
reintroduction_22008 %>% 
  write_csv(here::here("data", "raw", "22008_reintroduction_raw.csv"))

survey_22008 %>% 
  write_csv(here::here("data", "raw", "22008_survey_raw.csv"))

capture_22008 %>% 
  write_csv(here::here("data", "raw", "22008_capture_raw.csv"))
```

## Disconnect from database 

```{r db_disconnect}
connections::connection_close(con) 
rm(con)
```

- Remember to close SSH connection too. 

## Read in raw data files (if necessary)

```{r read_raw_data, eval=FALSE, message=FALSE}
# Change to eval=TRUE
reintroduction_22008 <- read_csv(here("data", "raw", "22008_reintroduction_raw.csv")) %>% 
  mutate(pit_tag_ref = as.character(pit_tag_ref))

survey_22008 <- read_csv(here("data", "raw", "22008_survey_raw.csv"))

capture_22008 <- read_csv(here("data", "raw", "22008_capture_raw.csv")) %>% 
  mutate(pit_tag_ref = as.character(pit_tag_ref))
```

## Clean reintroduction data

Run checks before and after data cleaning (next code chunk) to ensure that all issues have been resolved. Run lines of code one at a time to allow easy evaluation of messages and outputs. 

### Reintroduction data checks

```{r reintroduction_data_checks, message=FALSE}
# Check for records where pit_tag_ref is NULL
see_if(noNA(reintroduction_22008$pit_tag_ref))

# Check for records where bd_exposure is NULL
see_if(noNA(reintroduction_22008$bd_exposure))

# Check for records containing pit tags shorter than 15 characters (or 9 for early sites)
see_if(!any(str_length(reintroduction_22008$pit_tag_ref) != 15))

# Check for records containing pit tags that end with ".1" (known to be potentially incorrect) 
see_if(!any(str_detect(reintroduction_22008$pit_tag_ref, "\\.1$")))

# Check for type != "reintroduction"
see_if(all(reintroduction_22008$type == "reintroduction"))

# Check for frogs that died during reintroduction
see_if(!any(str_detect(reintroduction_22008$relocate_frog_comment, "dead|died")))

# Check for duplicate pit tags on each date
see_if(!any(reintroduction_22008 %>% count(release_date, pit_tag_ref) %>% pull(n) > 1))

# Are release dates correct?
reintroduction_22008 %>% distinct(release_date) %>% glimpse()
```

### Update reintroduction data

```{r update_reintroduction_data}
# No updates to make
```

### Create final reintroduction dataset formatted for mrmr

```{r final_reintroduction_data}
reintroduction_22008_mrmr <- reintroduction_22008 %>% 
  rename(pit_tag_id = pit_tag_ref,
         site_id = release_siteid1,
         treatment = bd_exposure) %>% 
  arrange(release_date, pit_tag_id) %>% 
  select(site_id, type, release_date, pit_tag_id, treatment)
```

### Save final reintroduction data

```{r save_final_reintroduction_data, message=FALSE}
reintroduction_22008_mrmr %>% write_csv(here("data", "clean", "22008_reintroduction_2groups.csv"))
```

### Read in cleaned reintroduction data (if necessary)

```{r read_cleaned_reintroduction_data, eval=FALSE, message=FALSE}
reintroduction_22008 <- read_csv(here::here("data", "clean", "22008_reintroduction_2groups.csv")) %>% 
  mutate(pit_tag_id = as.character(pit_tag_id))
```

## Clean survey dataset

### Review cmr survey history for all sites in project area

```{r} 
cmr_survey_check <- survey_22008 %>% 
  filter(survey_type == "cmr",
         visit_date > min(reintroduction_22008$release_date)) %>% 
  distinct(visit_date) %>%
  arrange(visit_date) %>%
  mutate(primary_period = cumsum(visit_date - lag(visit_date, default = first(visit_date)) > 5) + 1) %>% # Identify primary periods based on a gap between CMR surveys of > 5 days
  inner_join(survey_22008, by = "visit_date") %>% 
  filter(survey_type == "cmr") %>% 
  select(site_id, primary_period) %>% 
  group_by(primary_period) %>% 
  count(site_id) %>% 
  mutate(num_sec_periods = max(n), # of times each site was visited during primary period
         diff = num_sec_periods - n) %>% # difference between number of secondary periods and the number of visits to each site
  relocate(site_id) %>% 
  arrange(site_id, primary_period)
```

- Primary site 22008 surveyed during each secondary period. In most primary periods, 22019 was not visited, and when it was visited, it was usually visited less frequently. 

### Identify VES with no associated CMR period for possible inclusion as a CMR survey - customize for each site

```{r}
# For each primary period, create all dates in period +/- 2 days
cmr_dates <- survey_22008 %>% 
  filter(survey_type == "cmr",
         visit_date > min(reintroduction_22008$release_date) + days(5)) %>% # Remove CMR surveys from before or soon after frogs were first reintroduced
  distinct(visit_date) %>%
  arrange(visit_date) %>%
  mutate(primary_period = cumsum(visit_date - lag(visit_date, default = first(visit_date)) > 5) + 1) %>% # Identify primary periods based on a gap between CMR surveys of > 1 day
  group_by(primary_period) %>%
  summarise(  # buffer primary periods by 2 days at start and end
    start = min(visit_date) - days(2),
    end   = max(visit_date) + days(2),
    .groups = "drop"
  ) %>%
  rowwise() %>% 
  mutate(visit_date = list(seq.Date(start, end, by = "day"))) %>%  # Create all dates in each expanded period
  unnest(visit_date) %>%
  select(primary_period, visit_date)

# Select VES dates
ves_dates <- survey_22008 %>%
  filter(
    survey_type == "visual",
    visit_date > min(reintroduction_22008$release_date + days(5)), # Remove VES surveys from before or soon after frogs were first reintroduced
    visit_date <= max(cmr_dates$visit_date)
  ) %>% 
  select(visit_date) %>% 
  arrange(visit_date)

# Identify all visual surveys that fall outside of CMR primary periods (expanded)
unmatched_ves <- ves_dates %>% 
  anti_join(cmr_dates, by = "visit_date") %>% 
  distinct(visit_date)
```

- Unmatched VES on 2016-08-11, no adult frogs counted during survey. Consider for inclusion in survey history. 
- CMR surveys were conducted 8-14 to 8-16. Given temporal proximity to unmatched VES, no reason to include this VES.

### Remove unneeded objects

```{r}
rm(cmr_survey_check, cmr_dates, ves_dates, unmatched_ves)
```

### Remove irrelevant surveys from survey records

```{r remove_visual_surveys}
survey_22008_mrmr <- survey_22008 %>% 
  filter(survey_type == "cmr", 
         visit_date > min(reintroduction_22008$release_date)) %>% 
  rename(survey_date = visit_date)
```

- Review records for potential problematic surveys, for example, based on comment fields and fields describing current conditions.
  - Cold conditions during at least some secondary periods of 2022 survey. Consider dropping these.

### Select unique survey dates

```{r unique_survey_dates}
survey_22008_mrmr <- survey_22008_mrmr %>% 
  distinct(survey_date, survey_type, visit_status)
```

### Survey data checks

```{r survey_data_checks}
# Check for duplicate survey dates
see_if(!any(survey_22008_mrmr %>% count(survey_date) %>% pull(n) > 1))

# Check for sites that were unsuitable for survey
see_if(all(survey_22008_mrmr$visit_status == "suitable"))

# Check for non-cmr and non-reintroduction survey types
see_if(all(survey_22008_mrmr$survey_type == "cmr"))
```

### To meet mrmr format requirement, add reintroduction dates as surveys

```{r add_reintroductions}
survey_22008_mrmr <- reintroduction_22008_mrmr %>% 
  distinct(release_date, type) %>% 
  rename(survey_type = type, 
         survey_date = release_date) %>% 
  bind_rows(survey_22008_mrmr) %>% 
  mutate(site_id = 22008) %>%  # assign all surveys to primary site
  relocate(site_id, .before = survey_date) %>% 
  arrange(survey_date) 
```

### Assign primary and secondary periods

```{r assign_cmr_periods}
survey_22008_mrmr <- survey_22008_mrmr %>% 
  arrange(survey_date) %>% 
  # Create a flag for breaks between surveys: gap > 4 days OR survey_type == "reintroduction"
  mutate(gap_break = (survey_date - lag(survey_date, default = first(survey_date)) > 4),
    reintr_break = lag(survey_type == "reintroduction", default = FALSE),
    primary_period = cumsum(gap_break | reintr_break) + 1) %>%   
  # Assign secondary periods within each primary period
  group_by(primary_period) %>%
  mutate(secondary_period = row_number()) %>%
  ungroup() %>% 
  mutate(secondary_period = replace(secondary_period, survey_type == "reintroduction", 0)) %>% 
  select(site_id, survey_date, survey_type, primary_period, secondary_period)
```

### Save cleaned survey file

```{r save_cleaned_survey_data}
survey_22008_mrmr %>% write_csv(here("data", "clean", "22008_survey_2groups.csv"))
```

## Clean capture dataset

### Capture data checks

```{r capture_data_checks, message=FALSE}
# Verify that all dates in capture are present in survey 
see_if(nrow(capture_22008 %>% anti_join(survey_22008_mrmr, by = join_by(visit_date == survey_date))) == 0)

# Verify that visit_status is "suitable" for all records
see_if(all(capture_22008$visit_status == "suitable"))

# Verify that survey_type is "cmr" for all records
see_if(all(capture_22008$survey_type == "cmr"))

# Check for records where pit_tag_ref is NULL
see_if(noNA(capture_22008$pit_tag_ref))

# Check for records containing pit tags shorter than 15 characters (or 9 for early sites)
see_if(!any(nchar(capture_22008$pit_tag_ref) != 15))

# Check for records containing pit tags that end with ".1" (known to be potentially incorrect) 
see_if(!any(str_detect(capture_22008$pit_tag_ref, "\\.1$")))

# Check for records where species != ramu
see_if(all(capture_22008$species == "ramu"))

# Check for records where capture_animal_state is NULL
see_if(noNA(capture_22008$capture_animal_state))

# Check for records where capture_animal_state == "dead"
see_if(!any(capture_22008$capture_animal_state == "dead"))

# Check for records of subadult frogs 
see_if(all(capture_22008$length >= 40)) # 1 captured frog is < 40 mm, but was retained

# Check for duplicate pit tags on each date
see_if(nrow(capture_22008 %>% count(visit_date, pit_tag_ref) %>% filter(n > 1)) == 0)
```

### Clean capture data

```{r clean_capture_data}
# Drop dead frogs
capture_22008 <- capture_22008 %>% 
  filter(capture_animal_state != "dead")
```

- 1 frog was < 40 mm. Opted to retain it.  

### Create final capture dataset formatted for mrmr

```{r final_capture_data}
capture_22008_mrmr <- capture_22008 %>%
  rename(survey_date = visit_date, 
         pit_tag_id = pit_tag_ref) %>%
  mutate(site_id == 22008) %>%  # assign all captures to primary site
  select(site_id, survey_date, pit_tag_id, bd_load) %>% 
  inner_join(reintroduction_22008_mrmr %>%  # remove wild frogs
              select(pit_tag_id),
            by = "pit_tag_id") %>% 
  arrange(survey_date, pit_tag_id)
```

### Save cleaned capture data (one file with Bd load and one without)

```{r save_cleaned_capture_data}
capture_22008_mrmr %>% write_csv(here("data", "clean", "22008_bd_capture_2groups.csv"))
capture_22008_mrmr %>% select(-bd_load) %>% write_csv(here("data", "clean", "22008_capture_2groups.csv"))
```





