---
title: "22008 - Create datasets for analysis using *mrmr*"
author: "Roland Knapp"
output: html_notebook
editor_options: 
  chunk_output_type: inline
---

This notebook creates the survey, capture, and translocation files for use in analyzing capture-mark-recapture data using the *mrmr* package. The code retrieves data from the PostgreSQL amphibians database, runs error checks, and creates the relevant files in *mrmr* format. A treatment column is included in the reintroduction file (exposed, unexposed) and in the capture file (wild, exposed, unexposed) for inclusion as survival covariates. 

Users without credentials to access the database will not be able to create or use the raw data file that is central to this notebook. See cmr-analysis/README.md for details. 

Note that data from the most recent field season may not yet be appended to tables in the amphibians database. If these recently-collected data are necessary (e.g., for preliminary analyses), complete datasets can be created by copying code chunks in "fulcrum_data_append.Rmd" into this notebook and executing all code chunks. 

## Load packages

```{r load_packages}
if (!require(librarian)){
  install.packages("librarian")
  library(librarian)
}
librarian::shelf(DBI, config, here, readr, connections, dplyr, tidyr, stringr, magrittr, lubridate, assertthat)
```

## Connect to database

- Before running this code chunk, establish connection to database server (via VPN or SSH)

```{r db_connect}
# Select configuration of database of interest (connection paramaters for all databases provided in config.yml - in working directory)
Sys.setenv(R_CONFIG_ACTIVE = "amphibians-eri")

# Using connection_open from connections package shows structure of database tables in Connections pane
con <- connection_open(RPostgres::Postgres(),
                       dbname = config::get("dbname"),
                       host = config::get("host"),
                       port = config::get("port"),
                       user = config::get("user"),
                       password = rstudioapi::askForPassword("password")) # Enter password from password manager ("ERI GRIT Postgres")

# Set search path to schema of interest
dbExecute(con, "SET search_path TO public")

```

## Retrieve data subsets from PostgreSQL database

### Reintroduction data

```{r reintroduction_data}
relocate <- tbl(con, "relocate") %>%
  rename(
    relocate_comment = comment,
    relocate_id = id
  )

relocate_frog <- tbl(con, "relocate_frog") %>%
  rename(
    relocate_frog_comment = comment,
    relocate_frog_id = id
  )

reintroduction_22008 <- relocate %>%
  inner_join(relocate_frog, by = "relocate_id") %>%
  filter(
    release_siteid1 %in% c(22008, 22019)
  ) %>%
  select(
    release_siteid1,
    release_date,
    type,
    pit_tag_ref,
    tag_new,
    length,
    weight,
    swab_id,
    bd_exposure,
    relocate_comment,
    relocate_frog_comment
  ) %>%
  collect()
```

Review the reintroduction data subset for potential problems. 
- tag_new field is empty. 

### Survey data

To allow subsequent checks of whether surveys at which no frogs were captured are recorded as "visual" instead of "cmr", include visual surveys

```{r survey_data}
visit <- tbl(con, "visit") %>%
  rename(
    visit_comment = comment,
    visit_id = id
  )

survey <- tbl(con, "survey") %>%
  rename(
    survey_comment = comment,
    survey_id = id
  )

survey_22008 <- visit %>% 
  inner_join(survey, by = "visit_id") %>% 
  filter(
    site_id %in% c(22008, 22019),
    survey_type %in% c("cmr", "visual")
  ) %>% 
select( 
  site_id, 
  visit_date, 
  visit_status, 
  survey_type, 
  description, 
  survey_quality,
  wind,
  sky,
  duration,
  visit_comment,
  survey_comment
) %>% 
collect()

survey_22008 <- survey_22008 %>%  # convert ENUM columns to character
  mutate(across(where(is_pq_enum), as.character))

```

Review the survey data subset for potential problems. 
- Some surveys with inclement weather, likely worth retaining all surveys regardless. 

### Capture data

```{r capture_data}
visit <- tbl(con, "visit") %>%
  rename(
    visit_id = id
  )

survey <- tbl(con, "survey") %>%
  rename(
    survey_id = id
  )

capture <- tbl(con, "capture_survey") %>% 
  rename(
    capture_survey_comment = comment
  )

bdload <- tbl(con, "bd_load") %>% 
  rename(
    swab_id = sample_id
  )

capture_22008 <- visit %>% 
  inner_join(survey, by = "visit_id") %>% 
  inner_join(capture, by = "survey_id") %>% 
  left_join(bdload, by = "swab_id") %>% 
  filter(
    site_id %in% c(22008, 22019),
    survey_type == "cmr"
  ) %>% 
select( 
  site_id, 
  visit_date, 
  visit_status, 
  survey_type, 
  pit_tag_ref, 
  tag_new, 
  species, 
  capture_life_stage, 
  capture_animal_state, 
  length,
  swab_id,
  replicate,
  std_type,
  dilution,
  bd_load,
  capture_survey_comment
) %>% 
collect()

capture_22008 <- capture_22008 %>%  # convert ENUM columns to character
  mutate(across(where(is_pq_enum), as.character))
```

Review the capture data subset for potential problems. 
- At 22008, only 1 frog captured in 2022 survey. Was this a true CMR or only captured tagged frogs? Truncate dataset?
- qPCR used gblock and plasmid standards, all samples run singly.

## Save raw data files

```{r save_raw_data, message=FALSE}
reintroduction_22008 %>% 
  write_csv(here::here("data", "raw", "22008_reintroduction_raw.csv"))

survey_22008 %>% 
  write_csv(here::here("data", "raw", "22008_survey_raw.csv"))

capture_22008 %>% 
  write_csv(here::here("data", "raw", "22008_capture_raw.csv"))
```

## Disconnect from database 

```{r db_disconnect}
connection_close(con) 
rm(con)
```

## Read in raw data files (if necessary)

```{r read_raw_data, eval=FALSE, message=FALSE}
# Change to eval=TRUE
reintroduction_22008 <- read_csv(here("data", "raw", "22008_reintroduction_raw.csv")) %>% 
  mutate(pit_tag_ref = as.character(pit_tag_ref))

survey_22008 <- read_csv(here("data", "raw", "22008_survey_raw.csv"))

capture_22008 <- read_csv(here("data", "raw", "22008_capture_raw.csv")) %>% 
  mutate(pit_tag_ref = as.character(pit_tag_ref))
```

## Clean reintroduction data

Run checks before and after data cleaning (next code chunk) to ensure that all issues have been resolved. Run lines of code one at a time to allow easy evaluation of messages and outputs. 

### Reintroduction data checks

```{r reintroduction_data_checks, message=FALSE}
# Check for records where pit_tag_ref is NULL
assert_that(noNA(reintroduction_22008$pit_tag_ref))

# Check for records where bd_exposure is NULL
assert_that(noNA(reintroduction_22008$bd_exposure))

# Check for records containing pit tags shorter than 15 characters (or 9 for early sites)
assert_that(!any(str_length(reintroduction_22008$pit_tag_ref) != 15))

# Check for records containing pit tags that end with ".1" (known to be potentially incorrect) 
assert_that(!any(str_detect(reintroduction_22008$pit_tag_ref, "\\.1$")))

# Check for type != "reintroduction"
assert_that(all(reintroduction_22008$type == "reintroduction"))

# Check for frogs that died during reintroduction
assert_that(!any(str_detect(reintroduction_22008$relocate_frog_comment, "dead|died")))

# Check for duplicate pit tags on each date
assert_that(!any(reintroduction_22008 %>% count(release_date, pit_tag_ref) %>% pull(n) > 1))

# Are release dates correct?
reintroduction_22008 %>% distinct(release_date) %>% glimpse()
```

### Update reintroduction data

```{r update_reintroduction_data}
# No updates to make
```

### Create final reintroduction dataset formatted for mrmr

```{r final_reintroduction_data}
reintroduction_22008_mrmr <- reintroduction_22008 %>% 
  rename(pit_tag_id = pit_tag_ref,
         site_id = release_siteid1,
         treatment = bd_exposure) %>% 
  arrange(release_date, pit_tag_id) %>% 
  select(site_id, type, release_date, pit_tag_id, treatment)
```

### Save final reintroduction data

```{r save_final_reintroduction_data, message=FALSE}
reintroduction_22008_mrmr %>% write_csv(here("data", "clean", "22008_reintroduction.csv"))
```

### Read in cleaned reintroduction data (if necessary)

```{r read_cleaned_reintroduction_data, eval=FALSE, message=FALSE}
reintroduction_22008 <- read_csv(here::here("data", "clean", "22008_reintroduction.csv")) %>% 
  mutate(pit_tag_id = as.character(pit_tag_id))
```

## Clean survey dataset

### Review cmr survey history for all sites in project area

```{r} 
cmr_survey_check <- survey_22008 %>% 
  filter(survey_type == "cmr",
         visit_date >= "2016-01-01") %>% # Include CMR surveys from before 2016 but before first reintroduction (tagging of wild frogs), kept those conducted before and soon after frogs were reintroduced because captures included wild frogs
  distinct(visit_date) %>%
  arrange(visit_date) %>%
  mutate(primary_period = cumsum(visit_date - lag(visit_date, default = first(visit_date)) > 5) + 1) %>% # Identify primary periods based on a gap between CMR surveys of > 5 days
  inner_join(survey_22008, by = "visit_date") %>% 
  filter(survey_type == "cmr") %>% 
  select(site_id, primary_period) %>% 
  group_by(primary_period) %>% 
  count(site_id) %>% 
  mutate(num_sec_periods = max(n), # of times each site was visited during primary period
         diff = num_sec_periods - n) # difference between number of secondary periods and the number of visits to each site
```

- Primary site 22008 surveyed during each secondary period, 22019 visited less frequently in most primary periods. 

### Identify VES with no associated CMR period for possible inclusion as a CMR survey - customize for each site

```{r}
# For each primary period, create all dates in period +/- 2 days
cmr_dates <- survey_22008 %>% 
  filter(survey_type == "cmr",
         visit_date > min(reintroduction_22008$release_date) + days(5)) %>% # Remove CMR surveys from before or soon after frogs were first reintroduced
  distinct(visit_date) %>%
  arrange(visit_date) %>%
  mutate(primary_period = cumsum(visit_date - lag(visit_date, default = first(visit_date)) > 5) + 1) %>% # Identify primary periods based on a gap between CMR surveys of > 1 day
  group_by(primary_period) %>%
  summarise(  # buffer primary periods by 2 days at start and end
    start = min(visit_date) - days(2),
    end   = max(visit_date) + days(2),
    .groups = "drop"
  ) %>%
  rowwise() %>% 
  mutate(visit_date = list(seq.Date(start, end, by = "day"))) %>%  # Create all dates in each expanded period
  unnest(visit_date) %>%
  select(primary_period, visit_date)

# Select VES dates
ves_dates <- survey_22008 %>%
  filter(
    survey_type == "visual",
    visit_date > min(reintroduction_22008$release_date + days(5)), # Remove VES surveys from before or soon after frogs were first reintroduced
    visit_date <= max(cmr_dates$visit_date)
  ) %>% 
  select(visit_date) %>% 
  arrange(visit_date)

# Identify all visual surveys that fall outside of CMR primary periods (expanded)
unmatched_ves <- ves_dates %>% 
  anti_join(cmr_dates, by = "visit_date") %>% 
  distinct(visit_date)
```

- If no adult frogs found during unmatched VES, consider including as CMR survey.
- Unmatched VES on 2016-08-11 in which no adult frogs were captured, but CMR surveys were conducted 8-14 to 8-16. No justification to include this VES.

### Remove unneeded objects

```{r}
rm(cmr_survey_check, cmr_dates, ves_dates, unmatched_ves)
```

### Remove visual surveys from survey records

```{r remove_visual_surveys}
survey_22008 <- survey_22008 %>% 
  filter(survey_type == "cmr", 
         visit_date >= "2016-01-01") %>% # Remove CMR surveys from before 2016
  rename(survey_date = visit_date)
```

- Review records for potential problematic surveys, for example, based on comment fields and fields describing current conditions.
  - Cold conditions during at least some secondary periods of 2022 survey. Consider dropping these

### Select unique survey dates

```{r unique_survey_dates}
survey_22008_mrmr <- survey_22008 %>% 
  distinct(survey_date, survey_type, visit_status)
```

### Survey data checks

```{r survey_data_checks}
# Check for duplicate survey dates
assert_that(!any(survey_22008_mrmr %>% count(survey_date) %>% pull(n) > 1))

# Check for sites that were unsuitable for survey
assert_that(all(survey_22008_mrmr$visit_status == "suitable"))

# Check for non-cmr and non-reintroduction survey types
assert_that(all(survey_22008_mrmr$survey_type == "cmr"))
```

### To meet mrmr format requirement, add reintroduction dates as surveys

```{r add_reintroductions}
survey_22008_mrmr <- reintroduction_22008_mrmr %>% 
  distinct(release_date, type) %>% 
  rename(survey_type = type, 
         survey_date = release_date) %>% 
  bind_rows(survey_22008_mrmr) %>% 
  mutate(site_id = 22008) %>%  # assign all surveys to primary site
  relocate(site_id, .before = survey_date) %>% 
  arrange(survey_date) 
```

### Assign primary and secondary periods FIX THIS TO ALWAYS ASSIGN REINTRODUCTION SURVEYS THEIR OWN PERIOD

```{r assign_cmr_periods}
survey_22008_mrmr <- survey_22008_mrmr %>%
  arrange(survey_date) %>% 
  # Create a flag for "forced" breaks: gap > 4 days OR survey_type == "reintroduction"
  mutate(gap_break = (survey_date - lag(survey_date, default = first(survey_date)) > 4),
    reintr_break = lag(survey_type == "reintroduction", default = FALSE),
    primary_period = cumsum(gap_break | reintr_break) + 1) %>%   
  # Assign secondary periods within each primary period
  group_by(primary_period) %>%
  mutate(secondary_period = row_number()) %>%
  ungroup() %>% 
  mutate(secondary_period = replace(secondary_period, survey_type == "reintroduction", 0)) %>% 
  select(-visit_status)
```

### Save cleaned survey file

```{r save_cleaned_survey_data}
survey_22008_mrmr %>% write_csv(here("data", "clean", "22008_survey.csv"))
```

## Clean capture dataset

### Capture data checks

```{r capture_data_checks, message=FALSE}
# Verify that all dates in capture are present in survey 
assert_that(nrow(capture_22008 %>% anti_join(survey_22008_mrmr, by = join_by(visit_date == survey_date))) == 0)

# Verify that visit_status is "suitable" for all records
assert_that(all(capture_22008$visit_status == "suitable"))

# Verify that survey_type is "cmr" for all records
assert_that(all(capture_22008$survey_type == "cmr"))

# Check for records where pit_tag_ref is NULL
assert_that(noNA(capture_22008$pit_tag_ref))

# Check for records containing pit tags shorter than 15 characters (or 9 for early sites)
assert_that(!any(nchar(capture_22008$pit_tag_ref) != 15))

# Check for records containing pit tags that end with ".1" (known to be potentially incorrect) 
assert_that(!any(str_detect(capture_22008$pit_tag_ref, "\\.1$")))

# Check for records where species != ramu
assert_that(all(capture_22008$species == "ramu"))

# Check for records where capture_animal_state is NULL
assert_that(noNA(capture_22008$capture_animal_state))

# Check for records where capture_animal_state == "dead"
assert_that(!any(capture_22008$capture_animal_state == "dead"))

# Check for records of subadult frogs 
assert_that(all(capture_22008$length >= 40)) # 1 captured frog is < 40 mm, but was retained

# Check for duplicate pit tags on each date
assert_that(nrow(capture_22008 %>% count(visit_date, pit_tag_ref) %>% filter(n > 1)) == 0)
```

### Clean capture data
```{r clean_capture_data}
# Drop dead frogs
capture_22008 <- capture_22008 %>% 
  filter(capture_animal_state != "dead")
```

### Create final capture dataset formatted for mrmr
```{r final_capture_data}
capture_22008_mrmr <- capture_22008 %>%
  rename(survey_date = visit_date, 
         pit_tag_id = pit_tag_ref) %>%
  mutate(site_id == 22008) %>%  # assign all captures to primary site
  select(site_id, survey_date, pit_tag_id, bd_load) %>% 
  left_join(reintroduction_22008_mrmr %>% 
              select(pit_tag_id, treatment),
            by = "pit_tag_id") %>% 
  mutate(treatment = replace_na(treatment, "wild")) %>% 
  arrange(survey_date, pit_tag_id)
```

### Save cleaned capture data (one file with Bd load and one without)
```{r save_cleaned_capture_data}
capture_22008_mrmr %>% write_csv(here("data", "clean", "22008_bd_capture.csv"))
capture_22008_mrmr %>% select(-bd_load) %>% write_csv(here("data", "clean", "22008_capture.csv"))
```








