---
title: "R Notebook"
output: html_notebook
---
Analysis of 22008 CMR dataset
* use 'treatment' covariate in translocation file to evaluate survival of Bd-exposed and unexposed frogs

Install mrmr package if current version is not available
```{r}
remotes::install_github("SNARL1/mrmr")
```

Load required packages
```{r}
library(dplyr)
library(readr)
library(tibble)
library(rstan)
library(mrmr)
```

Set working directory for Tyndall Pond (may have to do this from console - following may not be working)
```{r}
knitr::opts_knit$set(root.dir = '/media/rknapp/Data/Box Sync/SNARL shared folder/Projects/cmr_analysis/sites/tyndall/tyndall')
getwd()
```

Create required data frames
```{r}
captures <- read_csv("captures.tyndallpond.mrmr.csv")
translocations <- read_csv("translocations.tyndallpond.mrmr.csv")
surveys <- read_csv("surveys.tyndallpond.mrmr.csv")
```

Read and clean the data
```{r}
    data_tyndallpond <- clean_data(captures, surveys, translocations,
                                    survival_formula = ~ treatment,
                                    survival_fill_value = c(treatment = "unexposed"))
```

Fit model
```{r}
# use 'chains = 1, iter = 10' for initial model check
# use 'chains = 3, iter = ~1000' for actual model run
model_tyndallpond <- fit_model(data_tyndallpond, cores = parallel::detectCores(), chains = 3, iter = 500)
```

Check model fit
```{r}
traceplot(model_tyndallpond$m_fit, par = "Nsuper")

pars_to_plot <- c('alpha_lambda', 
                  'sigma_lambda', 
                  'beta_phi', 
                  'sigma_phi', 
                  'beta_detect')
traceplot(model_tyndallpond$m_fit, pars = pars_to_plot)

print(model_tyndallpond$m_fit, pars = pars_to_plot)
```

Visualize results
```{r}
plot_model(model_tyndallpond, what = 'abundance')
plot_model(model_tyndallpond, what = 'recruitment')
plot_model(model_tyndallpond, what = 'survival')
```

Save model objects
```{r}
write_rds(model_tyndallpond, "model_tyndallpond_mrmr.rds")

# Or compressed
write_rds(model_tyndallpond, "model_tyndallpond_mrmr_compressed.rds", "xz", compression = 9L)
```

Read model objects back in
```{r}
model_tyndallpond <- read_rds("model_tyndallpond_mrmr_compressed.rds")
```

