---
title: "22008 - Analysis of CMR data using *mrmr* - 2 treatment groups"
author: "Roland Knapp"
output: html
---
Analysis of Tyndall Pond CMR dataset - Tyndall area
- "treatment" covariate is included in translocation and capture files to evaluate survival of zoo-raised Bd-exposed versus unexposed frogs

## Install mrmr package if current version is not available

```{r install_mrmr}
remotes::install_github("SNARL1/mrmr")
```

## Install cmdstanr package if current version is not available

```{r}
install.packages("cmdstanr", repos = c("https://stan-dev.r-universe.dev", getOption("repos")))
cmdstanr::install_cmdstan()
```

## Load packages

```{r load_packages}
if (!require(librarian)){
  install.packages("librarian")
  library(librarian)
}
librarian::shelf(here, cmdstanr, mrmr, bayesplot, posterior, tidyverse, reshape2, abind)
```

## Create required datasets

```{r mrmr_datasets}
captures <- read_csv(here("data", "clean", "22008_capture.csv")) %>% filter(treatment != "wild")  # Remove wild group
surveys <- read_csv(here("data", "clean", "22008_survey.csv"))
translocations <- read_csv(here("data", "clean", "22008_reintroduction.csv"))
```

### Specify model formula and check data

```{r mrmr_model_formula}
data_22008 <- clean_data(captures, surveys, translocations,
                                survival_formula = ~ treatment,
                                survival_fill_value = c(treatment = "unexposed")) 
```

### Fit model

```{r mrmr_fit_model}
# Used default of 1000 iterations for warmup and 1000 iterations for sampling
# "threads per chain" option and "cpp_options" did not speed up fit_model step
model_22008 <- fit_model(
  data_22008, 
  chains = 4,
  parallel_chains = 4
  ) 
```

### Save model object

```{r}
model_22008 %>% write_rds(here("out", "model", "22008_model_2groups.rds"), "xz", compression = 9L)
```

### Read model objects back in (as necessary)

```{r eval=FALSE}
model_22008 <- read_rds(here("out", "model", "22008_model_2groups.rds"))
```

### Evaluate convergence of the MCMC algorithm

- alpha_lamda: Baseline (intercept) of the recruitment process. In a Jolly-Seber model, recruitment (entry of new individuals into the population) in each primary period is modeled using a log link or logit link.
For instance, if log⁡(λt )=αλ+ϵλ, then alpha_lambda is the overall mean log recruitment rate across time period.
- sigma_lamda: Standard deviation of temporal random effects on recruitment (ε_λ,t). Therefore, sigma_lamda describes variability in recruitment through time. 
- beta_phi[1]: Intercept (baseline effect) for survival probability (φ). Survival is typically modeled on the logit scale: logit(ϕi,t)=βϕ,1+βϕ,2⋅Treatmenti+ϵϕ,t. beta_phi[1] is the average (logit-scale) survival for the reference group — typically "unexposed".
- beta_phi[2]: Treatment effect on survival probability. Continuing from beta_phi[1] above, beta_phi[2] modifies the baseline logit of survival for the treatment group (e.g., Bd-exposed vs. unexposed). Positive versus negative values indicate that the treatment increased or decreased survival, respectively. 
- sigma_phi: Standard deviation of temporal random effects on survival (ε_φ,t). Captures variability in survival between time periods.
- beta_detect[1]: Intercept (baseline effect) for detection probability (p). Represents the average log-odds of detecting an individual per secondary period, in the baseline group (e.g., unexposed).

```{r mrmr_convergence}
# Traceplots
pars_to_plot <- c("alpha_lambda", 
                  "sigma_lambda", 
                  "beta_phi", 
                  "sigma_phi", 
                  "beta_detect")

mcmc_trace(model_22008$m_fit$draws(pars_to_plot))

# Summary statistics, include Rhat (any Rhat > 1.01?)
model_22008$m_fit$summary(pars_to_plot)
```

### Built-in visualizations

```{r mrmr_visualizations}
plot_model(model_22008, what = "abundance")
plot_model(model_22008, what = "recruitment")
plot_model(model_22008, what = "survival")
```

### Compare posterior estimates for survival: Bd-exposed vs. unexposed individudals

- rstan saves the survival value as phi. In contrast, cmdstanr stores the components of the survival variable: beta_phi[1] = intercept, beta_phi[2] = treatment_effect, and eps_phi[1:13] = time (primary period) random effects. The following code calculates phi from these components. 

```{r}
# Extract phi components directly from cmdstanr fit
beta_phi_mat <- posterior::as_draws_matrix(model_22008$m_fit$draws("beta_phi"))
eps_phi_mat  <- posterior::as_draws_matrix(model_22008$m_fit$draws("eps_phi"))

# Dimensions
n_draws   <- nrow(beta_phi_mat)
n_periods <- ncol(eps_phi_mat)

# Make sure treatment is a factor with "unexposed" as reference
data_22008$survival_covariate_df$treatment <- 
  factor(data_22008$survival_covariate_df$treatment, levels = c("unexposed", "exposed"))

treatment_levels <- levels(data_22008$survival_covariate_df$treatment)

# Compute posterior phi for each treatment (vectorized)
phi_list <- vector("list", length(treatment_levels))
names(phi_list) <- treatment_levels

for (i in seq_along(treatment_levels)) {
  treatment <- treatment_levels[i]
  
  # Determine which columns of beta_phi_mat correspond to the treatment effect
  # Intercept = unexposed
  treat_idx <- match(treatment, treatment_levels)
  
  # Create matrices for vectorized addition
  intercept_mat <- matrix(beta_phi_mat[,1], nrow = n_draws, ncol = n_periods)
  
  # Treatment effects: exposed = beta_phi[,2], unexposed = 0
  if (treatment == "unexposed") {
    treatment_mat <- matrix(0, nrow = n_draws, ncol = n_periods)
  } else {
    treatment_mat <- matrix(beta_phi_mat[, treat_idx], nrow = n_draws, ncol = n_periods)
  }
  
  # Compute logit(phi) = intercept + treatment effect + random effect
  logit_phi <- intercept_mat + treatment_mat + eps_phi_mat
  
  # Convert to probability
  phi_list[[treatment]] <- plogis(logit_phi)
}

# Combine into 3D array: draws × periods × treatment
phi_array <- abind::abind(phi_list, along = 3)
dimnames(phi_array) <- list(
  iteration = NULL,
  primary_period = 1:n_periods,
  treatment = treatment_levels
)

# Combine and tidy for plotting
survival_df <- reshape2::melt(
  phi_array,
  varnames = c("iteration", "primary_period", "treatment"),
  value.name = "phi"
) %>% as_tibble()

# Merge with survey metadata (e.g., survey_date)
survival_df <- survival_df %>%
  left_join(
    data_22008$surveys %>% filter(secondary_period < 2),
    by = "primary_period"
  )

# Plot posterior survival densities
ggplot(survival_df, aes(x = phi, color = treatment)) +
  stat_density(geom = "line") +
  facet_wrap(~ survey_date, scales = "free_y") +
  xlab("Survival probability") +
  ylab("Posterior density") +
  scale_color_discrete("Treatment")

# Posterior probabilities of survival differences between groups 
p_exposed_gt_unexposed <- mean(beta_phi_mat[, 2] > 0)

cat("Posterior probability that Bd exposure increased survival vs unexposed:",
    p_exposed_gt_unexposed, "\n")
```

- Posterior probability that treatment increased survival: 0.144. That is, 14% of the posterior draws had a positive effect and 86% had a negative effect. Therefore, there is a 14% posterior probability that the treatment increased survival. 

### Plot the posterior distribution of the treatment effect

```{r}
# Extract posterior treatment effects
beta_exposed  <- as.numeric(beta_phi_mat[, "beta_phi[2]"])   # exposed vs unexposed

# Put into a tidy data frame
beta_df <- tibble(
  `Exposed − Unexposed` = beta_exposed
) %>%
  pivot_longer(
    cols = everything(),
    names_to = "comparison",
    values_to = "beta"
  )

# Plot
ggplot(beta_df, aes(x = beta)) +
  geom_density(fill = "skyblue", alpha = 0.5) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red") +
  facet_wrap(~ comparison, scales = "free") +
  xlab("Posterior effect size (logit scale)") +
  ylab("Density") +
  ggtitle("Posterior distributions of pairwise survival differences\n(logit-scale treatment effects)")

```

### Custom plots of cohort survival by treatment group

- Plot cohort survival by treatment group, with 95% credible intervals and lines/points on the posterior median.

```{r}
# --- Build helper objects used by the plot -----------------
# primary period -> first survey date and year
primary_period_dates <- model_22008$data$surveys %>%
  group_by(primary_period) %>%
  summarize(
    date = min(survey_date),
    year = min(year),
    .groups = "drop"
  )

# translocation release dates per individual (one row per pit_tag_id)
transloc_dates <- model_22008$data$translocations %>%
  distinct(pit_tag_id, release_date)

# --- Main pipeline: compute fraction alive by draw, summarize, and plot ---
p1 <- model_22008$m_fit$draws('s', format = "draws_df") %>%
  tidyr::pivot_longer(tidyselect::starts_with('s')) %>%
  # get_numeric_indices() is a helper from mrmr; if not available replace with your function
  mutate(get_numeric_indices(name)) %>%
  rename(primary_period = index_2) %>% 
  # map index_1 -> pit_tag_id (ensure indexing matches your model dims)
  mutate(pit_tag_id = dimnames(model_22008$data$stan_d$Y)[[1]][index_1]) %>%
  # keep only reintroduced individuals
  filter(pit_tag_id %in% as.character(model_22008$data$translocations$pit_tag_id)) %>%
  # add release_date for each pit_tag_id
  left_join(transloc_dates, by = "pit_tag_id") %>%
  # add treatment info for each individual (ensure by = "pit_tag_id")
  left_join(model_22008$data$survival_covariate_df, by = "pit_tag_id") %>%
  
  # compute fraction alive per draw + period + treatment + release cohort
  group_by(release_date, primary_period, .draw, treatment) %>%
  summarize(fraction_alive = mean(value == 2), .groups = "drop") %>%
  
  # remove primary period 1 if you don't want it (keeps periods > 1)
  filter(primary_period > 1) %>%
  
  # add survey dates for plotting, and keep only dates on/after release
  left_join(primary_period_dates, by = "primary_period") %>%
  filter(date >= release_date) %>%
  
  # summarize across posterior draws to get median and 95% CI per date/treatment/release
  group_by(release_date, date, treatment) %>%
  summarize(
    lo  = quantile(fraction_alive, probs = 0.025),
    med = median(fraction_alive),
    hi  = quantile(fraction_alive, probs = 0.975),
    .groups = "drop"
  ) %>%
  
  # prettify labels
  mutate(
    treatment = tools::toTitleCase(as.character(treatment)),
    facet_lab = paste("Released on", release_date)
  ) %>%
  
  # plot treatments, faceted by release cohort
  ggplot(aes(x = date, y = med, color = treatment, fill = treatment, group = treatment)) +
  geom_ribbon(aes(ymin = lo, ymax = hi), alpha = 0.30, color = NA) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 1.5) +
  facet_wrap(~ facet_lab, ncol = 1) +
  ylab("Fraction surviving") +
  xlab("Date") +
  scale_color_brewer("Treatment", palette = "Dark2") +
  scale_fill_brewer("Treatment", palette = "Dark2") +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
  ggtitle("A. Survival of reintroduced frogs") +
  theme_bw() +
  theme(
    legend.position = c(0.75, 0.85),
    legend.background = element_rect(fill = alpha("white", 0.8))
  )

# print plot
p1

```
- These cohort-specific results are not directly comparable to the global, treatment-level regression coefficients (posterior probabilities).
- Cohort-specific survival trajectories can look different from expectations from posterior probabilities because they depend on sample sizes, timing, shrinkage, and detection. 

### Posterior median survival probability (phi) by treatment group

- This is the most direct and interpretable comparison of survival between groups.
- Values come from the regression parameters (beta_phi), the random year/period effects (eps_phi), and do not depend on detection or imputed alive/dead states.

```{r}
# collapse eps over time (average across periods)
eps_phi_vec <- rowMeans(eps_phi_mat)

# compute logit(phi) for each treatment
phi_unexp  <- beta_phi_mat[,1] + eps_phi_vec
phi_exposed <- beta_phi_mat[,1] + beta_phi_mat[,2] + eps_phi_vec

# assemble into tidy data frame
plot_df1 <- tibble(
  draw = 1:nrow(beta_phi_mat),
  unexposed = plogis(phi_unexp),
  exposed   = plogis(phi_exposed)
)

plot_df1_long <- plot_df1 %>%
  pivot_longer(cols = -draw, names_to = "treatment", values_to = "phi")

# plot posterior distributions for all three treatments
library(viridis)

ggplot(plot_df1_long, aes(x = phi, fill = treatment)) +
  geom_density(alpha = 0.4) +
  scale_fill_viridis(discrete = TRUE, option = "C") +
  theme_bw() +
  xlab("Survival probability (φ)") +
  ylab("Posterior density") +
  ggtitle("Posterior survival probability φ by treatment group")
```

### Arrange survival, abundance, and recruitment plots

```{r}
# survival plot
p1 <- p1 +
#  guides(fill = guide_legend(nrow = 1)) + 
  theme(legend.position = "right")

# abundance plot
p2 <- plot_model(model_22008, what = "abundance") 
p2 <- p2 + ggtitle("B. Population size") +
    ylab("Number of adults")

# recruitment plot
p3 <- plot_model(model_22008, what = "recruitment")
p3 <- p3 + ggtitle("C. Recruitment") +
    ylab("Number of adults")

library(patchwork) 

p1 / p2 / p3 +
  plot_layout(heights = c(3, 1, 1))

ggsave(here("out", "plots", "22008_mrmr_plots_treatment.png"), width = 6.5, height = 7.2)
```

### Quick model check

- If the treatment increased survival, then in 2018 and 2019, we should have seen a greater proportion of treated than untreated frogs from the initial translocation.

```{r}
cohort_transloc <- translocations %>%
  filter(release_date == as.Date("2016-07-18"))

count(cohort_transloc, treatment) %>%
  mutate(frac = n / sum(n))
  
captures %>%
  inner_join(translocations, by = "pit_tag_id") %>% 
  select(survey_date, pit_tag_id, treatment) %>% 
  filter(survey_date > as.Date("2018-01-01"), 
         pit_tag_id %in% cohort_transloc$pit_tag_id) %>%
  group_by(survey_date) %>%
  count(treatment) %>%
  mutate(frac = n / sum(n))
```

- Our null expectation is that we should see approximately 71% exposed and 29% unexposed individuals based on the proportions of frogs in each category in this first translocation.
- For each survey, we see a somewhat larger fraction of exposed individuals than expected based on our null expectation.

### Posterior predictive cumulative survival curves

```{r}
# helper objects 
n_draws   <- nrow(beta_phi_mat)
n_periods <- ncol(eps_phi_mat)

# create primary_period -> date table if available
if (!exists("primary_period_dates")) {
  if (!is.null(model_22008$data$surveys)) {
    primary_period_dates <- model_22008$data$surveys %>%
      group_by(primary_period) %>%
      summarize(date = min(survey_date), .groups = "drop")
  } else {
    primary_period_dates <- tibble(primary_period = 1:n_periods)
  }
}

# treatments and corresponding beta column indices
treatments <- c("unexposed", "exposed")
beta_idx <- c(unexposed = 1, exposed = 2)

# function to compute cumulative survival per draw × period for a given treatment
compute_cum_surv <- function(beta_col_index) {
  # replicate intercept and treatment effect across periods (matrices n_draws x n_periods)
  intercept_mat <- matrix(beta_phi_mat[, 1], nrow = n_draws, ncol = n_periods)
  treat_mat     <- matrix(0, nrow = n_draws, ncol = n_periods)
  if (!is.null(beta_col_index) && beta_col_index > 1) {
    treat_mat <- matrix(beta_phi_mat[, beta_col_index], nrow = n_draws, ncol = n_periods)
  }
  # eps_phi_mat already n_draws x n_periods
  logit_phi_mat <- intercept_mat + treat_mat + eps_phi_mat

  # convert to probability
  phi_mat <- plogis(logit_phi_mat)           # matrix: draw × period

  # compute cumulative survival across periods for each draw
  # each row is a draw; cumprod along columns
  cumphi_mat <- t(apply(phi_mat, 1, cumprod)) # still draw × period

  # convert to long tibble: draw, period, cum_surv
  tibble(
    draw = rep(1:n_draws, times = n_periods),
    period = rep(1:n_periods, each = n_draws),
    cum_surv = as.vector(cumphi_mat)
  )
}

# compute for each treatment and bind
list_of_dfs <- lapply(treatments, function(trt) {
  df <- compute_cum_surv(beta_idx[[trt]])
  df$treatment <- trt
  df
})
cum_surv_df <- bind_rows(list_of_dfs)

# Summarize across draws: median and 95% CI per treatment × period
cum_surv_summ <- cum_surv_df %>%
  group_by(treatment, period) %>%
  summarize(
    lo  = quantile(cum_surv, probs = 0.025),
    med = median(cum_surv),
    hi  = quantile(cum_surv, probs = 0.975),
    .groups = "drop"
  ) %>%
  # attach dates if available
  left_join(primary_period_dates, by = c("period" = "primary_period")) %>%
  mutate(x = if ("date" %in% colnames(.)) date else period,
         treatment = tools::toTitleCase(treatment))

# Plot: cumulative survival through time (period or date on x-axis)
# --- release cohort table (unique release dates) ---
release_cohorts <- model_22008$data$translocations %>%
  distinct(release_date) %>%
  arrange(release_date) %>%
  mutate(release_label = paste("Released on", release_date))

# --- attach release cohort to cumulative survival summaries ---
cum_surv_summ_facet <- cum_surv_summ %>%
  crossing(release_cohorts) %>%   # Cartesian join: each treatment gets curves for each release cohort
  mutate(treatment = tools::toTitleCase(treatment))

# --- plot, faceting by release cohort ---
ggplot(cum_surv_summ_facet,
       aes(x = x, y = med, color = treatment, fill = treatment)) +
  geom_ribbon(aes(ymin = lo, ymax = hi),
              alpha = 0.25, color = NA) +
  geom_line(size = 1) +
  geom_point(size = 1.5) +
  scale_color_viridis_d(option = "C") +
  scale_fill_viridis_d(option = "C") +
  theme_bw() +
  xlab(ifelse("date" %in% colnames(cum_surv_summ_facet),
              "Date", "Primary period")) +
  ylab("Cumulative survival (model-predicted)") +
  ggtitle("Posterior predictive cumulative survival curves by treatment\nFaceted by release cohort") +
  facet_wrap(~ release_label, ncol = 1) +
  theme(legend.position = "bottom")
```









