---
title: "50183 - Create datasets for analysis using *mrmr*"
author: "Roland Knapp"
output: html_notebook
editor_options: 
  chunk_output_type: inline
---

This notebook creates the survey, capture, and translocation files for use in analyzing capture-mark-recapture data using the *mrmr* package. The code retrieves data from the PostgreSQL amphibians database, runs error checks, and creates the relevant files in *mrmr* format. The code is written in distinct chunks to allow review of results from one chunk before running the next. For example, review of the raw data files is necessary to identify potentially problematic records for review and possible removal from the data set.

Before using the code in this file, replace all instances of "xxxxx" with the 5-digit site_id of interest.

Users without credentials to access the database can use the saved CSV files created in the save-raw-data-files code chunk.

Note for credentialed users: Data from the most recent field season may not yet be appended to tables in the amphibians database. If these recently-collected data are necessary (e.g., for preliminary analyses), complete datasets can be created by copying code chunks in "fulcrum_data_append.Rmd" into this notebook and executing all code chunks.

## Load packages

```{r load-packages}
library(tidyverse)
library(RPostgreSQL)
library(lubridate)
library(assertthat)
# options(readr.show_col_types = FALSE)
```

## Create database connection
```{r db-connect-set-schemas}
pg = dbDriver("Postgres")
con_amphibians_v2 = dbConnect(pg, 
                user = rstudioapi::askForPassword("user name"), 
                password = rstudioapi::askForPassword("password"),
                host = Sys.getenv("host"), 
                port = Sys.getenv("port"), 
                dbname = Sys.getenv("dbname")  
                # options = "-c search_path=field_surveys" # add this to access the schema for all tables; versus setting schema as each table is accessed and assigned
                )
DBI::dbExecute(con_amphibians_v2, "SET search_path TO field_surveys, qpcr") # now the schemas are set, in order of query priority
DBI::dbListTables(con_amphibians_v2) # to test the connection and schemas


```

## Create current-year survey and capture datasets for appending
### Load current year data from local files.
#### Alternative Create current year data used in 2026
In 2026, the data are in more spread out locations and different tables that don't seem to adhere to the existing 'fulcrum_data_append.rmd' script, so the following script works around that.

Loading reintros, visits, surveys, captures, and Bd data from local sources and/or qpcr database allows us to build the data as needed prior to this year's data append.

```{r create_current_year_append_tables_from_local_source, eval=FALSE}
# REINTRO  - because 50183 had reintro in 2025 prior to last survey date.
reintro_visit_50183_current <- read_csv("P:/MLRG/code/ErrorChecking/2025/cleaned_data/reloc_visit_260124.csv")

reintro_survey_50183_current <- read_csv("P:/MLRG/code/ErrorChecking/2025/cleaned_data/reloc_survey_260124.csv")

reintro_release_50183_current <- read_csv("P:/MLRG/code/ErrorChecking/2025/cleaned_data/reloc_release_260124.csv")

reintro_individual_50183_current <- read_csv("P:/MLRG/code/ErrorChecking/2025/cleaned_data/reloc_indiv_260124.csv") %>% 
  mutate(pit_tag_ref = as.character(pit_tag_ref))

reintro_50183_current <-
  reintro_visit_50183_current %>% 
  filter(site_id == 50183) %>% 
  inner_join(reintro_survey_50183_current, by = "visit_id") %>% 
  inner_join(reintro_release_50183_current, by = "survey_id") %>% 
  inner_join(reintro_individual_50183_current, by = "release_id") %>% 
  select(site_id, release_date=visit_date, pit_tag_id=pit_tag_ref, bd_exposure)
  
# SURVEYS
### visit
visit_50183_current <- read_csv("P:/MLRG/code/ErrorChecking/2025/cleaned_data/final/append/visit_append_260323.csv") %>% 
  rename(visit_comment = comment, 
         visit_id = id)

# survey
survey_50183_current <- read_csv("P:/MLRG/code/ErrorChecking/2025/cleaned_data/final/append/survey_append_260323.csv") %>% 
  rename(survey_id = id, 
         survey_comment = comment ) %>% 
  select(survey_id, visit_id, survey_type, description, survey_quality, wind, sky, air_temp, duration, survey_comment)

# capture
capture_50183_current <- read_csv("P:/MLRG/code/ErrorChecking/2025/cleaned_data/final/append/capture_append_260409.csv") %>% 
  mutate(pit_tag_ref = as.character(pit_tag_ref)) %>% 
  rename(capture_survey_id = id,
         capture_comment = comment, 
         pit_tag_id = pit_tag_ref) %>% 
  select(capture_survey_id, survey_id, pit_tag_id, tag_new, species, capture_life_stage, capture_animal_state, length, weight, capture_comment, swab_id)
```
### Load current year Bd loads
```{sql retrieve-capture-data, connection=con_amphibians_v2, output.var = "bd_loads_only_all" }
select 
  bd_load.sample_id AS swab_id,
  bd_load.bd_load
from bd_load
```


### format the current year data for binding to historic data
```{r format-current-year-for-binding-to-historic}
# reintro_current_50183 ### should be OK as written above
# visit_50183_current
# survey_50183_current
# capture_50183_current
# bd_loads_only_all

# survey
survey_50183_2025 <- visit_50183_current %>% 
  filter(site_id %in% c( 50183 ) ) %>% 
  inner_join( survey_50183_current , by = "visit_id" ) %>% 
  filter(survey_type == "cmr") %>% 
  select(site_id, 
    visit_date, 
    visit_status, 
    survey_type, 
    description, 
    survey_quality,
    wind,
    sky,
    air_temp,
    duration,
    visit_comment,
    survey_comment)
  
# capture
capture_50183_2025 <- visit_50183_current %>% 
  filter(site_id %in% c( 50183) ) %>% 
  inner_join( survey_50183_current , by = "visit_id" )  %>% 
  filter( survey_type=="cmr" ) %>%
  # filter( survey_type %in% c("cmr", "night") ) %>% # included for exploratory purposes, night captures net two additional individuals
  inner_join( capture_50183_current, by = "survey_id" ) %>% 
  select(site_id, 
    visit_date,
    visit_status, 
    survey_type, 
    pit_tag_id, 
    tag_new, 
    species, 
    capture_life_stage, 
    capture_animal_state, 
    length,
    weight,
    visit_comment,
    survey_comment,
    capture_survey_comment=capture_comment,
    visit_id,
    survey_id,
    capture_survey_id,
    swab_id
  ) %>% 
  left_join(bd_loads_only_all, by = "swab_id" )
```

### Clean up environment
```{r clean-environment-carefully}
# rm(amphibian_capture_50183, survey_50183, visit_50183, relocate_50183, relocate_amphibians_50183)
```

### Retrieve reintroduction data from database

- Edit "where" statement to include relevant site_ids and types.  

```{sql retrieve_reintroduction_data, connection=con_amphibians_v2, output.var = "reintroduction_50183"}
select 
  visit.site_id, 
  release.release_type AS type,
  visit.visit_date AS release_date,
  release_frog.pit_tag_ref AS pit_tag_id, 
  release_frog.tag_new, 
  release_frog.sex, 
  release_frog.length, 
  release_frog.weight, 
  release_frog.swab_id, 
  release_frog.bd_exposure
from release_frog
INNER JOIN release ON release_frog.release_id = release.release_id
INNER JOIN visit ON release.visit_id = visit.visit_id
WHERE visit.site_id = 50183 
      
```

#### Bind current year reintro IDs
```{r bind-current-year-reintro-frogs}
reintroduction_50183 <- reintroduction_50183 %>% 
  bind_rows(reintro_50183_current)
```


### Retrieve SURVEY data from database

-   Edit "where" statement to include relevant site_ids and types.
-   If some relevant surveys are classified as "ves", include in "where" statement.

```{sql retrieve-survey-data, connection=con_amphibians_v2, output.var = "survey_50183"}
select 
  site_id, 
  visit_date, 
  visit_status, 
  survey_type, 
  description, 
  survey_quality,
  wind,
  sky,
  air_temp,
  duration,
  visit_comment,
  survey_comment
from visit
inner join survey on visit.visit_id = survey.visit_id
where site_id in (50183) AND 
      survey_type in ('cmr', 'visual') AND
      visit_date > '2020-01-01'
```
```{r}
view(survey_50183)
```

#### Add current year surveys if relevant
```{r current-year-surveys-bind-2025}
survey_50183 <- survey_50183 %>% 
  bind_rows( survey_50183_2025 %>% 
               select(site_id, 
                      visit_date, 
                      visit_status,
                      survey_type,
                      description, 
                      survey_quality,
                      wind,
                      sky,
                      air_temp,
                      duration,
                      visit_comment,
                      survey_comment
                      )
             )

```


### Retrieve frog CAPTURE data from database
-   Edit "where" statement to include relevant site_ids and types.

```{sql retrieve-capture-data, connection=con_amphibians_v2, output.var = "capture_50183"}

select 
  visit.site_id, 
  visit.visit_date, 
  visit.visit_status, 
  survey.survey_type, 
  capture_survey.pit_tag_ref AS pit_tag_id, 
  capture_survey.tag_new, 
  capture_survey.species, 
  capture_survey.capture_life_stage, 
  capture_survey.capture_animal_state, 
  capture_survey.length,
  capture_survey.weight,
  visit.visit_comment,
  survey.survey_comment,
  capture_survey.capture_survey_comment,
  visit.visit_id,
  survey.survey_id,
  capture_survey.capture_survey_id,
  capture_survey.swab_id,
  bd_load.bd_load
from visit
inner join survey on visit.visit_id = survey.visit_id
inner join capture_survey on survey.survey_id = capture_survey.survey_id
left join bd_load on bd_load.sample_id = capture_survey.swab_id
where site_id in (50183) and
      survey_type in ('cmr') and
      visit_date > '2020-01-01'
      
```

#### Add current year captures if relevant
```{r current-year-surveys-bind-2025}
capture_50183 <- capture_50183 %>% 
  bind_rows(capture_50183_2025)
```
 
 
### Disconnect database
```{r db-disconnect3}
dbDisconnect(con)
rm(con)
```

-   Review retrieved datasets before using in subsequent chunks.

### Save raw data files

```{r save-raw-data-files, message=FALSE}
reintroduction_50183 %>% write_csv(here::here("data", "raw", "50183_reintroduction_raw.csv")) # if relevant
survey_50183 %>% write_csv(here::here("data", "raw", "50183_survey_raw.csv"))
capture_50183 %>% write_csv(here::here("data", "raw", "50183_capture_raw.csv"))
```

- Read saved files with (for example): 
> translocation_50183 <- read_csv(here::here("data", "raw", "50183_translocation_raw.csv")) %\>% mutate(pit_tag_ref = as.character(pit_tag_ref))

## Create reintroduction dataset

- Skip if reintroductions are not relevant.  

### Retain only relevant columns

```{r reintroduction-relevant-columns}
#reintroduction_50183 <- reintroduction_50183 %>% 
  #rename(site_id = release_siteid1) %>% 
  #select(site_id, release_date, pit_tag_ref, tag_new, sex, length, weight, swab_id, bd_exposure)
```

### Reintroduction data checks

- Run checks before and after data cleaning (next code chunk) to ensure that all issues have been resolved.  
- Run lines of code one at a time to allow easy evaluation of messages and outputs.  

```{r reintroduction-data-checks, message=FALSE}
# Check for records where pit_tag_ref is NULL
assert_that(noNA(reintroduction_50183$pit_tag_ref))

# Check for records containing pit tags shorter than 15 characters
assert_that(!any(nchar(reintroduction_50183$pit_tag_ref) != 15))

# Check for records containing pit tags that end with ".1" (known to be potentially incorrect) 
assert_that(!any(str_detect(reintroduction_50183$pit_tag_ref, "\\.1$")))

# Check for records of subadult frogs
reintroduction_50183 %>% drop_na(length) %>% filter(length < 40)
# IT looks like a large number of Humphreys frogs in 2023 and 2024 were released at less than 40 mm

# Check for duplicate pit tags on each date
reintroduction_50183 %>% count(release_date, pit_tag_ref) %>% filter(n > 1)

# Are release dates correct?
reintroduction_50183 %>% distinct(release_date) %>% glimpse()
```

### Clean reintroduction data

- Edit data-cleaning code as necessary.  

```{r clean-translocation-data}
# no cleaning needed as of April 2026
```

### Create final reinroduction dataset formatted for mrmr

For 50183, one release date gets modified. A release was performed in the evening following a CMR survey, but to make them appear as separate primary periods we change the date of the release event by +1 day.

```{r final-reintroduct-data}
# adjust 2024 release date to make distinct primary period
reintroduction_50183 <- reintroduction_50183 %>% 
  mutate( release_date = case_when( release_date==ymd("2024-07-31") ~ ymd("2024-08-01") , 
                                    .default = release_date )
                                    ) %>% 
  select(site_id, release_date, pit_tag_id, bd_exposure)
```

### Save cleaned reintroduction data

```{r save-cleaned-reitnroduction-data, message=FALSE}
reintroduction_50183 %>% 
  write_csv(here::here("data", "clean", "50183_reintroduction.csv"))
```

- When visually viewing dataset in microsoft excel or libre office, format pit_tag_id as character.

## Create SURVEY dataset

### Append current-year survey records

```{r append_currentyear_survey_records}
survey_50183 <- survey_50183 %>% 
  arrange(visit_date, desc(site_id)) %>% 
  select(site_id, visit_date, visit_status, survey_type, description, survey_quality, wind, sky, air_temp, duration, visit_comment, survey_comment)
```

### Include only relevant surveys

Marmot Lake surveys include only Marmot Lake (50183).

#### Primary site surveyed during each secondary period?

-   If yes, code chunk will return zero records. Exclude other records as necessary.
-   In filter statement, edit site_id as necessary.

```{r check-survey-history}
survey_50183 %>% 
  add_count(visit_date) %>% 
  filter(n == 1 & site_id != "50183")
```

#### Include only surveys at primary site

-   Run this code chunk as is when only CMR surveys need to be included. If VES surveys also need to be included, modify code accordingly.

```{r include-relevant-surveys}
survey_50183 <- survey_50183 %>% 
  filter(site_id == 50183, 
         survey_type == 'cmr' ) %>%
  rename(survey_date = visit_date) %>% 
  arrange(survey_date)
```

### Add reintroduction dates as surveys
```{r add-reintroduction-surveys}
# change survey date name
# survey_50183 <- survey_50183 # %>% 
  # rename(survey_date=visit_date)

survey_50183 <- reintroduction_50183 %>% 
  distinct(site_id, release_date ) %>% 
  rename( survey_date = release_date ) %>%
  mutate( survey_type = 'zoo_reintroduction' ) %>% 
    bind_rows( survey_50183 ) %>% 
  arrange(survey_date) %>% 
  mutate(survey_date_2 = case_when(survey_date==ymd("2024-07-31") & survey_type=="zoo_reintroduction" ~ ymd("2024-08-01"), 
                                   TRUE ~ survey_date),
         survey_date = survey_date_2) %>% 
  select(-survey_date_2) %>% # revises the 2024 reintroduction to be +1 day to not overlap with the CMR; it occurred in the evening after the day's survey was complete.
  unique()
```

### Survey data checks

-   Run checks before and after data cleaning (next code chunk) to ensure that all issues have been resolved.\
-   Run lines of code one at a time to allow easy evaluation of messages and outputs.

```{r survey-data-checks}
# Check for duplicate survey dates
survey_50183 %>% count(site_id, survey_date) %>% filter(n > 1)

# Check for sites that were unsuitable for survey
survey_50183 %>% filter(visit_status != "suitable")
```

### Clean survey data

-   Edit data-cleaning code as necessary.

```{r clean-survey-data}
# Drop duplicate surveys
survey_50183 <- survey_50183 %>% 
  distinct(survey_date, .keep_all = TRUE)
```

### Create primary and secondary period columns in mrmr format

-   Edit case_when steps to reflect actual translocation/reintroduction and survey history.

#### For robust design dataset

```{r create-period-columns1}
# view all survey dates
survey_50183 %>%
  select(site_id, survey_date, survey_type) %>% 
  arrange(survey_date)


survey_50183 <- survey_50183 %>%
  select(site_id, survey_date, survey_type) %>% 
  # filter(survey_date > "2023-01-01") %>% 
  mutate(primary_period = case_when(survey_date==ymd("2022-09-16") ~ 1, # 2022 reintroduction
                                    between(survey_date, ymd("2023-09-05"), ymd("2023-09-06")) ~ 2,
                                    survey_date==ymd("2023-09-09") ~ 3, # 2023 reintroduction
                                    between(survey_date, ymd("2023-09-19"), ymd("2023-09-20")) ~ 4,
                                    between(survey_date, ymd("2024-07-29"), ymd("2024-07-31")) ~ 5,
                                    survey_date==ymd("2024-08-01") ~ 6, # 2024 reintroduction, which was actually evening of 7/31,, after surveys completed, but recoded above to create distinct primary periods
                                    between(survey_date, ymd("2024-09-06"), ymd("2024-09-08")) ~ 7, 
                                    between(survey_date, ymd("2025-07-03"), ymd("2025-07-05")) ~ 8,
                                    survey_date==ymd("2025-07-06") ~ 9, 
                                    survey_date==ymd("2025-08-21") ~10 ))


survey_50183 <- survey_50183 %>% 
  arrange(primary_period, survey_date) %>% 
  group_by(primary_period) %>% 
  mutate(secondary_period = seq_len(n()),
         secondary_period = replace(secondary_period, survey_type != "cmr", 0)) 
```

### Save cleaned survey file

```{r save-cleaned-survey-data}
survey_50183 %>% write_csv(here::here("data", "clean", "50183_survey.csv"))
```

## Create capture dataset

### Append current-year capture records, retain only relevant columns

```{r append_currentyear_capture_records}
# remove 2010 data
capture_50183 <- capture_50183 %>% 
  filter(visit_date > "2022-01-01")

# append 2023, 2024, and current year data
# capture_50183 <- capture_append_50183 %>% 
#  select(site_id, visit_date, visit_status, survey_type, pit_tag_ref, tag_new, species, capture_life_stage, capture_animal_state, length, comment_visit, comment_survey, comment_capture) %>% 
#  bind_rows(capture_50183)
```

### Retain only relevant columns

```{r capture-relevant-columns}
capture_50183 <- capture_50183 %>% 
  select(-visit_id, -survey_id, -capture_survey_id)
```

### Capture data checks

-   Run checks before and after data cleaning (next code chunk) to ensure that all issues have been resolved.\
-   Run lines of code one at a time to allow easy evaluation of messages and outputs.

```{r capture-data-checks, message=FALSE}
# Check for records where pit_tag_ref is NULL
assert_that(noNA(capture_50183$pit_tag_id))

# Check for records containing pit tags shorter than 15 characters (or 9 for early sites)
assert_that(!any(nchar(capture_50183$pit_tag_id) != 15))

# Check for records containing pit tags that end with ".1" (known to be potentially incorrect) 
assert_that(!any(str_detect(capture_50183$pit_tag_id, "\\.1$")))

# Check for records where species != ramu or rasi
capture_50183 %>% drop_na(species) %>% filter(species != "rasi")

# Check for records were survey_type = swab 
assert_that(!any(capture_50183$survey_type == "swab"))

# Check for records where capture_animal_state is NULL
assert_that(noNA(capture_50183$capture_animal_state))

# Check for records where capture_animal_state == "dead"
assert_that(!any(capture_50183$capture_animal_state == "dead"))

# Check for records of subadult frogs
capture_50183 %>% drop_na(length) %>% filter(length < 40)

# Check for sites that were unsuitable for survey
assert_that(!any(capture_50183$visit_status != "suitable"))

# Check for duplicate capture records on each date
capture_50183 %>% count(visit_date, pit_tag_id) %>% filter(n > 1)

# Check for captures that lack associated surveys
survey_50183 %>% 
  select(survey_date, primary_period) %>% 
  rename(visit_date = survey_date) %>% 
  right_join(capture_50183, by = c("visit_date")) %>% 
  filter(is.na(primary_period))

# Simpler alternative
setdiff(capture_50183$visit_date, survey_50183$survey_date) %>% as.Date(origin = "1970-01-01")

# Check for surveys that lack associated captures
survey_50183 %>% 
  filter(survey_type == "cmr") %>% 
  select(survey_date, primary_period) %>% 
  rename(visit_date = survey_date) %>% 
  left_join(capture_50183, by = c("visit_date")) %>% 
  filter(is.na(species))
# for 50183, there were two CMR surveys both in 2023 that were known to have zero observed frogs, for two primary periods of two days each - in other words, four survey rows have no associated frogs

# Simpler alternative
setdiff(survey_50183$survey_date, capture_50183$visit_date) %>% as.Date(origin = "1970-01-01") # this includes reintroduction event visit dates, which have no captures
```

### Clean capture data
-   Edit data-cleaning code as necessary. As of April 2026, no cleaning or revisions were made to individual frog capture data.

```{r clean-capture-data}
# Nothing to clean, everything checked out above
```

### Create final capture dataset formatted for mrmr

```{r final-capture-data}
capture_50183 <- capture_50183 %>%
  rename(survey_date = visit_date) %>%
  arrange(survey_date, pit_tag_id) %>%
  select(site_id, survey_date, pit_tag_id, bd_load)
```

### Save cleaned capture data

```{r save-cleaned-capture-data}
capture_50183 %>% write_csv(here::here("data", "clean", "50183_capture.csv"))
```


## Sanity check on data files:
```{r view-files-for-sanity}
read_csv( here::here("data", "clean", "50183_capture.csv") )  %>% 
  mutate(pit_tag_id = as.character(pit_tag_id) ) %>% 
  view()

read_csv( here::here("data", "clean", "50183_survey.csv") )  %>% 
  view()

read_csv( here::here("data", "clean", "50183_reintroduction.csv") )  %>% 
  mutate(pit_tag_id = as.character(pit_tag_id) ) %>% 
  view()



```


