---
title: "50783 - Analyze capture-mark-recapture data using *mrmr*"
author: "Roland Knapp"
output: html_notebook
editor_options: 
  chunk_output_type: inline
---

This script analyzes capture-mark-recapture data using the *mrmr* package. The code creates the required survey and capture tibbles from saved CSV files (created by xxxxx_createdata_mrmr.qmd), checks the datasets, fits the model, evaluates potential non-convergence of the MCMC algorithm, and visualizes model results. The code is written to be run on a local desktop computer with 8 physical cores and 32 MB of RAM. 

## Install latest versions of cmdstanr, cmdstan, and mrmr packages

Run only if updated versions were released since previous install

```{r install-stan-mrmr-packages}
# install cmdstanr and cmdstan
install.packages("cmdstanr", repos = c("https://mc-stan.org/r-packages/", getOption("repos")))

cmdstanr::install_cmdstan()

# install mrmr
remotes::install_github("SNARL1/mrmr")
```

## Setup

```{r}
# Load installed packages
if (!require(librarian)){
  install.packages("librarian")
  library(librarian)
}
librarian::shelf(here, readr, dplyr, tibble, cmdstanr, mrmr, patchwork)
```

## Create required data frames

```{r create-data-objects}
captures <- read_csv(here("data","clean", "50783_capture.csv"))
surveys <- read_csv(here("data","clean", "50783_survey.csv"))
```

## Check datasets

```{r check-data}
data_50783 <- clean_data(captures = captures, surveys = surveys)
```

## Fit model

* Code assumes that there are 8 physical cores available. Update parallel_chains and threads_per_chain as necessary based on actual number of cores. (parallel_chains * threads_per_chain) must equal the number of cores. 
* Add "adapt_delta = 0.99" or similar if divergent transitions occur. 

```{r fit-model}
model_50783 <- fit_model(
                    data = data_50783, 
                    chains = 4, 
                    parallel_chains = 4, 
                    cpp_options = list(stan_threads = TRUE),
                    threads_per_chain = 2, 
                    adapt_delta = 0.99)

```

* Using the Roland's local computer, the model-fitting process required ~50 minutes with adapt_delta = 0.8 (default) and 135 minutes with adapt_delta = 0.99. 
* Running model with default adapt_delta (0.8) resulted in one divergent transition (1/4000). Adjusting adapt_delta to 0.99 resulted in 0/4000 divergent transitions. 

### Save/retrieve model object

This step prevents having to rerun the model in the event of a crash during model visualizations in subsequent code blocks. Model files can be large, so this saves the file in a compressed format.

```{r write-model}
write_rds(model_50783, here("out", "model", "50783_model.rds"), "xz", compression = 9L)
```

```{r read-model}
model_50783 <- read_rds(here("out", "model", "50783_model.rds"))
```

### Diagnose potential non-convergence of the MCMC algorithm

Inspect traceplots and parameter summary table. In table, are any R~hat~ values $\ge 1.01$?

```{r model-diagnostics}
pars_to_plot <- c('alpha_lambda', 
                  'sigma_lambda', 
                  'beta_phi', 
                  'sigma_phi', 
                  'beta_detect')
bayesplot::mcmc_trace(model_50783$m_fit$draws(pars_to_plot))

model_50783$m_fit$summary(pars_to_plot)
```

* With adapt_delta = 0.8, trace plots for some parameters show a bit more between-chain variation than is typical and R~hat~ for alpha_lamda is $\ge 1.01.
* When rerun with adapt_delta = 0.99, R~hat~ values were all $\lt 1.01. Trace plots still show more variability than usual. 

## Visualize model results

### Default plots

Note: Custom plots (produced in next code chunk: "plots_custom") are often easier to interpret than the default plots produced by this code chunk.

```{r plots-default}
# plot_model(model_50783, what = "abundance")
# plot_model(model_50783, what = "recruitment")
# plot_model(model_50783, what = "survival")
```

### Customize plots

```{r plots-custom, fig.height=5, fig.width=7}
library(ggplot2)
p1 <- plot_model(model_50783, what = "abundance") 
p1 <- p1 + ggtitle("A. Population size") +
    ylab("Number of adults") + 
    expand_limits(y = 0)
  
p2 <- plot_model(model_50783, what = "recruitment")
p2 <- p2 + ggtitle("B. Recruitment") +
    ylab("Number of adults")
  
p1 / p2
```

### Save custom plots
```{r save-plot}
ggsave(here("out", "plots", "50783_mrmr_plots.png"), height = 5, width = 6.5)
```