---
title: "50785 - Create datasets for analysis using *mrmr*"
author: "Roland Knapp and Thomas C. Smith"
output: html_notebook
editor_options: 
  chunk_output_type: inline
---

This notebook creates the survey, capture, and translocation files for use in analyzing capture-mark-recapture data using the *mrmr* package.
The code retrieves data from the PostgreSQL amphibians database, runs error checks, and creates the relevant files in *mrmr* format.
The code is written in distinct chunks to allow review of results from one chunk before running the next.
For example, review of the raw data files is necessary to identify potentially problematic records for review and possible removal from the data set.

Users without credentials to access the database will not be able to create or use the raw data file that is central to this notebook.
See cmr-analysis/README.md for details.

Note for credentialed users: Data from the most recent field season may not yet be appended to tables in the amphibians database.
If these recently-collected data are necessary (e.g., for preliminary analyses), complete datasets can be created by copying code chunks in "fulcrum_data_append.Rmd" into this notebook and executing all code chunks.

## Load packages

```{r load-packages}
library(tidyverse)
library(RPostgreSQL)
library(lubridate)
library(assertthat)
library(DBI)
library(RPostgres)
library(here)
# options(readr.show_col_types = FALSE)
```

For times when current year records had not been appended, a script was written to access that data from Fulcrum flat files and create objects to append to database data.
```{r create_append_files, eval=FALSE}
visit_50785 <- read_csv(here::here("data", "raw", "current_year", "survey_50785.csv")) %>% 
  select(record_id, site_id, visit_date, visit_status, visit_comment) %>% 
  rename(comment_visit = visit_comment)

survey_50785 <- read_csv(here::here("data", "raw", "current_year", "survey_description_50785.csv")) %>% 
  select(record_id, survey_type, survey_description, wind, sky, air_temp, duration, survey_quality, general_survey_comment) %>% 
  filter(survey_type == "cmr") %>% 
  rename(description = survey_description, comment_survey = general_survey_comment)

amphibian_capture_50785 <- read_csv(here::here("data", "raw", "current_year", "survey_amphibian_capture_50785.csv")) %>% 
  filter(method == "cmr") %>% 
  rename(comment_capture = capture_comment) %>% 
  mutate(pit_tag_ref = as.character(pit_tag_ref)) %>% 
  replace_na(list(species = "ramu", capture_life_stage = "adult")) %>% 
  select(record_id, pit_tag_ref, tag_new, species, capture_life_stage, capture_animal_state, length, comment_capture)

relocate_50785 <- read_csv(here::here("data", "raw", "current_year", "relocate_50785.csv")) %>% 
  filter(relocate_type == "reintroduction") %>% 
  rename(comment_relocate = relocate_comments,
         type = relocate_type) %>% 
  select(record_id, release_location, type, release_date, comment_relocate)

relocate_amphibians_50785 <- read_csv(here::here("data", "raw", "current_year", "relocate_amphibians_50785.csv")) %>% 
  mutate(pit_tag_ref = as.character(pit_tag_ref)) %>% 
  rename(relocate_amphibians_comments = comments) %>% 
  select(record_id, pit_tag_ref, tag_new, collection_stage, sex, length, weight, swab_id, relocate_amphibians_comments)

survey_append_50785 <- visit_50785 %>% 
  inner_join(survey_50785, by = c("record_id")) %>% 
  select(-record_id)

capture_append_50785 <- visit_50785 %>% 
  inner_join(survey_50785, by = c("record_id")) %>% 
  inner_join(amphibian_capture_50785, by = c("record_id")) %>% 
  select(-record_id)

translocation_append_50785 <- relocate_50785 %>% 
  inner_join(relocate_amphibians_50785, by = c("record_id")) %>% 
  rename(site_id = release_location, 
         capture_life_stage = collection_stage) %>% 
  select(-record_id)
```
```{r}
rm(amphibian_capture_50785, relocate_50785, relocate_amphibians_50785, survey_50785, visit_50785)
```
As of April 2026, 50785 and related sites have not been surveyed in several years, so the 'current year' data process can be ignored, and move on to the database access process specified below.


### Connect to PostgreSQL database. 

### Connect to PostgreSQL database (see below for alternative)

See next chunk for alternative. Users without credentials can use CSV file created in code chunk "read_raw_translocation_data" below as a starting point. 
```{r open-db-connection-set-schemas}
pg = dbDriver("Postgres")
con_amphibians_v2 = dbConnect(pg, 
                user = rstudioapi::askForPassword("user name"), 
                password = rstudioapi::askForPassword("password"),
                host = Sys.getenv("host"), 
                port = Sys.getenv("port"), 
                dbname = Sys.getenv("dbname")  
                # options = "-c search_path=field_surveys" # add this to access the schema for all tables; versus setting schema as each table is accessed and assigned
                )
DBI::dbExecute(con_amphibians_v2, "SET search_path TO field_surveys, qpcr")

```

### Connect to PostgreSQL database (use only when connecting via SSH)
```{r}
con = dbConnect(dbDriver("PostgreSQL"),
                host = rstudioapi::askForPassword("host"), 
                port = rstudioapi::askForPassword("port"), 
                dbname = Sys.getenv("dbname"),
                user = rstudioapi::askForPassword("user name"),
                password = rstudioapi::askForPassword("password")
               )
```

### Retrieve frog reintroduction data from database
```{sql retrieve reintroduction data, connection=con_amphibians_v2, output.var = "reintroduction_50785"}
select 
  visit.site_id, 
  release.release_type AS type,
  visit.visit_date AS release_date,
  release_frog.pit_tag_ref AS pit_tag_id, 
  release_frog.tag_new, 
  release_frog.sex, 
  release_frog.length, 
  release_frog.weight, 
  release_frog.swab_id, 
  release_frog.bd_exposure
from release_frog
INNER JOIN visit_release ON release_frog.visit_release_id = visit_release.visit_release_id
INNER JOIN visit ON visit_release.visit_id = visit.visit_id
LEFT JOIN release ON visit_release.release_id = release.release_id
WHERE visit.site_id = 50785 
```

```{r disconnect-when-all-data-acquired, message=FALSE, echo=FALSE}
dbDisconnect(con)
rm(con)
```

### Save/retrieve raw translocation data
First, a chunk that saves the table created above.
```{r save_raw_reintroduction_data, message=FALSE}
reintroduction_50785 %>% write_csv(here::here("data", "raw", "50785_reintroduction_raw.csv"))
```

If not recreating the table, reload said table here.
```{r read_raw_reintroduction_data, message=FALSE}
reintroduction_50785 <- read_csv(here::here("data", "raw", "50785_reintroduction_raw.csv")) 
```
Next, we review the REINTRODUCTION data subset for potential problems. 

# Raw Data Review and Revision

### Append current-year reintroduction records (if current year not appended)
As of April 2026, no current year data are created for 50785
```{r append_currentyear_reintroduction_records, eval=FALSE}
#translocation_50785 <- translocation_append_50785 %>% 
  #filter(type == "reintroduction") %>% 
#  mutate(across(c("length", "weight"), as.integer),
#         swab_id = as.character(swab_id)) %>% 
  #select(site_id, release_date, type, pit_tag_ref, tag_new, sex, length, weight, swab_id) %>% 
  #bind_rows(translocation_50785)
```

### Reintroduction data checks
Run checks before and after data cleaning (next code chunk) to ensure that all issues have been resolved. Run lines of code one at a time to allow easy evaluation of messages and outputs. 
```{r translocation_data_checks, message=FALSE}
# Check for records where pit_tag_ref is NULL
assert_that(noNA(reintroduction_50785$pit_tag_id))

# Check for records containing pit tags shorter than 15 characters
assert_that(!any(nchar(reintroduction_50785$pit_tag_id) != 15))

# Check for records containing pit tags that end with ".1" (known to be potentially incorrect) 
assert_that(!any(str_detect(reintroduction_50785$pit_tag_id, "\\.1$")))

# Check for records of subadult frogs
reintroduction_50785 %>% drop_na(length) %>% filter(length < 40)

# Check for duplicate pit tags on each date
reintroduction_50785 %>% count(release_date, pit_tag_id) %>% filter(n > 1)

# Are release dates correct?
reintroduction_50785 %>% distinct(release_date) %>% glimpse()

# do all frogs have a Bd_exposure history?
assert_that(noNA(reintroduction_50785$bd_exposure))
```
As of April 2026, the bd_exposure records for the 2021 releases are all NA. Sleuthing required to complete that field. Per Tom and Alexa's sleuthing in April, **no exposures** at all occurred for frogs in 2021. Thus, these NAs can be easily recoded to indicate that. That should be also be communicated in any metadata for the 50785 project data.

### Clean reintroduction data
```{r clean_reintroduction_data}
reintroduction_50785 <- reintroduction_50785 %>% 
  filter(pit_tag_id != "6039") %>% # Apparently 6039 was a pit tag ID that had historically been misentered but is fixed in the database as of April 2026, so this is retained to document history of data cleaning although it seems to be fixed at the database level.
  filter(!is.na(pit_tag_id))


reintroduction_50785 <- reintroduction_50785 %>% 
  mutate(bd_exposure = case_when( is.na(bd_exposure)==T ~ "none_occurred" , 
                                  between(release_date,
                                          ymd("2020-01-01"), 
                                          ymd("2020-12-31"))  == T ~ "none_occurred" ,
                                          .default = bd_exposure)  )
```

### Create final reintroduction datasets formatted for mrmr
1) Save the standard minimal version with only the variables site_id, type, release_date, pit_tag_id.
```{r final_reintroduction_data_minimal}
reintroduction_50785_mrmr_minimal <- reintroduction_50785 %>% 
  arrange(release_date, pit_tag_id) %>% 
  select(site_id, type, release_date, pit_tag_id)
```

2) Save the second version that contains the Bd exposure history (or any other release-frog covariates).
```{r final_reintroduction_data_bd_exposure}
reintroduction_50785_mrmr_bd_exposure <- reintroduction_50785 %>% 
  arrange(release_date, pit_tag_id) %>% 
  select(site_id, type, release_date, pit_tag_id, bd_exposure)
```

### Save/retrieve cleaned reintroduction data if not creating it above
1) Save or retrieve the standard minimal version with only the variables site_id, type, release_date, pit_tag_id.
```{r save_cleaned_reintroduction_data_minimal, message=FALSE}
reintroduction_50785_mrmr_minimal %>% write_csv(here::here("data", "clean", "50785_reintroduction.csv"))
```
```{r read_cleaned_translocation_data_minimal, message=FALSE}
reintroduction_50785_minimal <- read_csv(here::here("data", "clean", "50785_reintroduction.csv")) %>% 
  mutate(pit_tag_id = as.character(pit_tag_id))
```

2) Save the second version that contains the Bd exposure history (or any other release-frog covariates).
```{r save_cleaned_reintroduction_data, message=FALSE}
reintroduction_50785_mrmr_bd_exposure %>% write_csv(here::here("data", "clean", "50785_reintroduction_Bd.csv"))
```
```{r read_cleaned_translocation_data, message=FALSE}
reintroduction_50785_bd_exposure <- read_csv(here::here("data", "clean", "50785_reintroduction_Bd.csv")) %>% 
  mutate(pit_tag_id = as.character(pit_tag_id))
```


## Create SURVEY dataset
### Retrieve survey data from database

If surveys at which no frogs were captured are classified as "ves", include in "where" statement (e.g., where site_id = 70619 and survey_type in ('cmr', 'visual'))

```{sql retrieve-survey-data, connection=con_amphibians_v2, output.var = "survey_50785"}
select 
  site_id, 
  visit_date, 
  visit_status, 
  survey_type, 
  description, 
  survey_quality,
  wind,
  sky,
  air_temp,
  duration,
  visit_comment,
  survey_comment
from visit
inner join survey on visit.visit_id = survey.visit_id
where site_id in ( 50784, 50785, 50786, 50787, 50788,51751, 42595, 51752) and
      survey_type in ( 'cmr', 'visual' )
    
```

```{r db-disconnect2 eval=false}
# disconnect from DB if relevant
# dbDisconnect(con)
# rm(con)
```

### Append current-year survey records if applicable to current year
As of April 2026, 50785 and associated lakes have no current year surveys to append.
```{r append_currentyear_survey_records, eval=FALSE}
survey_50785 <- bind_rows(survey_50785, survey_append_50785) %>% 
  arrange(visit_date, desc(site_id)) %>% 
  select(site_id, visit_date, visit_status, survey_type, description, survey_quality, wind, sky, air_temp, duration, comment_visit, comment_survey)
```

### Save/retrieve raw survey data
```{r save-raw-survey-data}
survey_50785 %>% write_csv(here::here("data", "raw", "50785_survey_raw.csv"))
```

```{r read-raw-survey-data}
#survey_50785 <- read_csv(here::here("data", "raw", "50785_survey_raw.csv"))
```

### Include only relevant site visits

When multiple numbered sites exist in survey area, include only one survey for entire site on each survey date. 50785 is the primary site at this location and was surveyed during each secondary period (as indicated by lack of records produced by following code chunk).

```{r check-survey-history}
# Check if primary site was surveyed during each visit
survey_50785 %>% 
  add_count(visit_date) %>% 
  filter(n == 1 & site_id != "50785")
```
```{r include-relevant-surveys}
# When only cmr surveys exist in survey history
survey_50785_mrmr <- survey_50785 %>% 
  filter(site_id == 50785) %>%
  rename(survey_date = visit_date) %>% 
  arrange(survey_date)
```
As of query in April 2026, the above data table includes VES surveys from 2011.

### Add REINTRODUCTION dates as surveys
To meet mrmr format requirement
```{r add_reintroduction_surveys}
survey_50785_mrmr <- reintroduction_50785_mrmr %>% 
  distinct(site_id, release_date, type) %>% 
  rename(survey_type = type, survey_date = release_date) %>% 
  bind_rows(survey_50785_mrmr) %>% 
  arrange(survey_date) 
```
Review records for potential problematic surveys, for example, based on comment fields and fields describing current conditions.
- As mentioned above, the 2011 VES isn't going to be a CMR related survey and will be removed.


### Survey data checks

Run checks before and after data cleaning (next code chunk) to ensure that all issues have been resolved.

```{r survey-data-checks}
# Check for duplicate survey dates
survey_50785_mrmr %>% count(site_id, survey_date) %>% filter(n > 1)

# Check for sites that were unsuitable for survey
survey_50785_mrmr %>% filter(visit_status != "suitable")
```
Here, those survey dates represented by n>1 are due to the inclusion of the VES survey data, knowing that some surveys at 50785 and nearby have instances with zero observations or captures of individual frogs. Those will have to be addressed in the next step, cleaning.

### Clean survey data

```{r clean-survey-data}
# Drop duplicate surveys
survey_50785_mrmr <- survey_50785_mrmr %>% 
  distinct(survey_date, .keep_all = TRUE)
```

```{r remove-non-CMR-related-visits }
survey_50785_mrmr <- survey_50785_mrmr %>% 
  filter(survey_date > "2019-01-01")
```


### Create primary and secondary period columns in mrmr format

```{r create-period-columns}
# For robust design dataset
survey_50785_mrmr <- survey_50785_mrmr %>%
  select(site_id, survey_date, survey_type) %>% 
  mutate(primary_period = case_when(survey_date == "2019-07-16" ~ 1,
                                    between(survey_date, ymd("2019-08-08"), ymd("2019-08-09")) ~ 2,
                                    between(survey_date, ymd("2020-06-20"), ymd("2020-06-21")) ~ 3,
                                    survey_date == "2020-07-16" ~ 4,
                                    between(survey_date, ymd("2020-08-09"), ymd("2020-08-11")) ~ 5,
                                    between(survey_date, ymd("2021-06-15"), ymd("2021-06-17")) ~ 6,
                                    survey_date == "2021-06-18" ~ 7,
                                    between(survey_date, ymd("2021-08-10"), ymd("2021-08-11")) ~ 8,
                                    between(survey_date, ymd("2022-06-13"), ymd("2022-06-14")) ~ 9)) 
# 50785 has one visit-survey that is represented by VES only, 2021-06-15; it is adjacent to and precedes two CMR surveys, and thus seems like it should be part of the primary period and not excluded, so it is retained here. Zoo-reintroduction visits are set to secondary period equals zero.
survey_50785_mrmr <- survey_50785_mrmr %>% 
  arrange(primary_period, survey_date) %>% 
  group_by(primary_period) %>% 
  mutate(secondary_period = seq_len(n()),
         secondary_period = replace(secondary_period, survey_type == "zoo_reintroduction", 0)) 

# For single day visit dataset
# survey_50785_mrmr <- survey_50785_mrmr %>%
  #select(site_id, survey_date, survey_type) %>% 
  #mutate(primary_period = row_number()) %>% 
  #group_by(primary_period) %>% 
  #mutate(secondary_period = seq_len(n()),
         #secondary_period = replace(secondary_period, survey_type != "cmr", 0))
```

### Save cleaned survey file

```{r save-cleaned-survey-data}
survey_50785_mrmr %>% write_csv(here::here("data", "clean", "50785_survey.csv"))
```
```{r read-cleaned-survey-data}
# survey_50785_mrmr <- read_csv(here::here("data", "clean", "50785_survey.csv"))
```

## Create CAPTURE dataset

### Connect to PostgreSQL database if not still connected as above.
```{r db-connect3}
con = dbConnect(dbDriver("PostgreSQL"),
                host = Sys.getenv("host"), 
                port = Sys.getenv("port"), 
                dbname = Sys.getenv("dbname"),
                user = rstudioapi::askForPassword("user name"),
                password = rstudioapi::askForPassword("password")
               )
```
### Connect to PostgreSQL database (use only when connecting via SSH)
```{r db-connect3-alt}
con = dbConnect(dbDriver("PostgreSQL"),
                host = rstudioapi::askForPassword("host"), 
                port = rstudioapi::askForPassword("port"), 
                dbname = Sys.getenv("dbname"),
                user = rstudioapi::askForPassword("user name"),
                password = rstudioapi::askForPassword("password")
               )
```

### Retrieve frog capture data from database

```{sql retrieve-post-reintro-capture-data, connection=con_amphibians_v2, output.var = "capture_50785"}
select 
  visit.site_id, 
  visit.visit_date, 
  visit.visit_status, 
  survey.survey_type, 
  capture_survey.pit_tag_ref, 
  capture_survey.tag_new, 
  capture_survey.species, 
  capture_survey.capture_life_stage, 
  capture_survey.capture_animal_state, 
  capture_survey.length,
  capture_survey.weight,
  visit.visit_comment,
  survey.survey_comment,
  capture_survey.capture_survey_comment,
  visit.visit_id,
  survey.survey_id,
  capture_survey.capture_survey_id,
  capture_survey.swab_id,
  bd_load.bd_load
from visit
inner join survey on visit.visit_id = survey.visit_id
inner join capture_survey on survey.survey_id = capture_survey.survey_id
left join bd_load on bd_load.sample_id = capture_survey.swab_id
where site_id in (50784, 50785, 50786, 50787, 50788, 51751, 51752, 42595) and
      survey_type = 'cmr'
```

To join the Bd_load table to the captures table, use a left join so as not to limits the number of captures because not all captures generate a swab, but all captures are valid. 


```{r db-disconnect3}
dbDisconnect(con)
rm(con)
```

### Append current-year capture records (if capture records exist in current-year Fulcrum)
As of April 2026, no current year capture records exist in Fulcrum; skip.
```{r append_currentyear_capture_records}
capture_50785 <- capture_append_50785 %>% 
  select(site_id, visit_date, visit_status, survey_type, pit_tag_ref, tag_new, species, capture_life_stage, capture_animal_state, length, comment_visit, comment_survey, comment_capture) %>% 
  bind_rows(capture_50785)
```

### Save/retrieve raw capture data

```{r save-raw-capture-data}
capture_50785 %>% write_csv(here::here("data", "raw", "50785_capture_raw.csv"))
```

```{r read-raw-capture-data}
# capture_50785 <- read_csv(here::here("data", "raw", "50785_capture_raw.csv")) %>% 
#  mutate(pit_tag_ref = as.character(pit_tag_ref))
```

Review records for potential problematic captures, for example, based on comment fields and fields describing frog characteristics (including tag_new).

### Capture data checks

Run checks before and after data cleaning.

```{r capture-data-checks, message=FALSE}
# Check for records where pit_tag_ref is NULL
assert_that(noNA(capture_50785$pit_tag_ref))

# Check for records containing pit tags shorter than 15 characters (or 9 for early sites)
assert_that(!any(nchar(capture_50785$pit_tag_ref) != 15))

# Check for records containing pit tags that end with ".1" (known to be potentially incorrect) 
assert_that(!any(str_detect(capture_50785$pit_tag_ref, "\\.1$")))

# Check for records where species != ramu
capture_50785 %>% drop_na(species) %>% filter(species != "ramu")

# Check for records were survey_type = swab
assert_that(!any(capture_50785$survey_type == "swab"))

# Check for records where capture_animal_state is NULL
assert_that(noNA(capture_50785$capture_animal_state))

# Check for records where capture_animal_state == "dead"
assert_that(!any(capture_50785$capture_animal_state == "dead"))
# Three individuals are dead. They are removed in subsequent code chunk.

# Check for records of subadult frogs
capture_50785 %>% drop_na(length) %>% filter(length < 40)

# Check for sites that were unsuitable for survey
assert_that(!any(capture_50785$visit_status != "suitable"))

# Check for duplicate capture records on each date
capture_50785 %>% count(visit_date, pit_tag_ref) %>% filter(n > 1)

# Check for captures that lack associated surveys
survey_50785_mrmr %>% 
  select(survey_date, primary_period) %>% 
  rename(visit_date = survey_date) %>% 
  right_join(capture_50785, by = c("visit_date")) %>% 
  filter(is.na(primary_period))

# Simpler alternative
setdiff(capture_50785$visit_date, survey_50785_mrmr$survey_date) %>% as.Date(origin = "1970-01-01")

# Check for surveys that lack associated captures
survey_50785_mrmr %>% 
  filter(survey_type == "cmr") %>% 
  select(survey_date, primary_period) %>% 
  rename(visit_date = survey_date) %>% 
  left_join(capture_50785, by = c("visit_date")) %>% 
  filter(is.na(species))

# Simpler alternative
setdiff(survey_50785_mrmr$survey_date, capture_50785$visit_date) %>% as.Date(origin = "1970-01-01")

```

### Clean capture data
Per the review we did above, three dead animals were found and could be removed. However, for bookkeeping purposes, the PIT tag IDs of those dead frogs are first explicitly recorded here:
```{r}
# Check for records where capture_animal_state == "dead"
capture_50785 %>% 
  filter(capture_animal_state == "dead") %>% 
  select(pit_tag_ref, visit_date)
# Three individuals are dead in the data from the database. 
```
-  982126056334212 on 2020-06-20				
-  982126056337058 on 2021-06-15				
-  982091062745642 on 2021-08-10

```{r clean-capture-data-livedead-bdloads}
capture_50785_livedead_bdloads <- capture_50785
capture_50785_live_bdloads <- capture_50785 %>% 
  filter(capture_animal_state != "dead")
```

### Create final capture dataset formatted for mrmr

1) create minimal dataset without Bd loads and without dead frogs
```{r final-capture-data}
capture_50785_mrmr <- capture_50785_live_bdloads %>%
  rename(survey_date = visit_date, pit_tag_id = pit_tag_ref) %>% # this renaming could be done in the SQL script that creates the capture records
  arrange(survey_date, pit_tag_id) %>% 
  select(site_id, survey_date, pit_tag_id)
```

2) create dataset with Bd loads and with dead frogs.
```{r final-capture-data-with-Bd}
capture_50785_mrmr_Bd <- capture_50785_livedead_bdloads %>%
  rename(survey_date = visit_date, pit_tag_id = pit_tag_ref) %>% # this renaming could be done in the SQL script that creates the capture records
  arrange(survey_date, pit_tag_id) %>% 
  select(site_id, survey_date, pit_tag_id, capture_animal_state, bd_load)
```

### Save cleaned capture data

```{r save-cleaned-capture-data}
capture_50785_mrmr %>% write_csv(here::here("data", "clean", "50785_capture.csv"))
capture_50785_mrmr_Bd %>% write_csv(here::here("data", "clean", "50785_capture_bd.csv"))
```

```{r read-cleaned-capture-data}
# capture_50785_mrmr <- read_csv(here::here("data", "clean", "50785_capture.csv")) %>% 
#    mutate(pit_tag_id = as.character(pit_tag_id))
```



<!--  Next steps:

1. create Bd load at each capture
- go back to data creation step above.
- add loads from Bd_load view to capture records.
- save separate data file from the standard, most simple capture records as indicated by the files for Ranger/Tyndall
- COMPLETE 4/17 5:39pm

2. create zoo exposure history
- go back to data creation step above.
- add exposure status to reintroduction
- save separate data file from the standard, most simple capture records as indicated by the files for Ranger/Tyndall
- COMPLETE 4/17 11:20am

3. create zoo exposure load trajectories
- this is somewhat of a separate project and probably a separate script and effort

--> 

