---
title: "50783 - Create datasets for analysis using *mrmr*"
author: "Roland Knapp"
format: html
editor_options: 
  chunk_output_type: inline
---

This notebook creates the survey and capture files for use in analyzing capture-mark-recapture data using the *mrmr* package. No frog translocations or reintroductions were conducted at this site. The code retrieves data from the PostgreSQL amphibians_v2 database and current-year Fulcrum data, runs error checks, and creates the relevant files in *mrmr* format. 

Users without credentials to access the database will not be able to create the raw data files, but can use the saved clean data files for analyses. See cmr-analysis/README.md for details. 

## Setup

```{r}
# Load installed packages
if (!require(librarian)){
  install.packages("librarian")
  library(librarian)
}
librarian::shelf(DBI, here, readr, dplyr, tidyr, stringr, lubridate, assertthat)
```

- Before running the next code chunk, establish connection to database server (via VPN or SSH)

## Connect to database

```{r}
# Connect to database
con <- dbConnect(RPostgres::Postgres(),  
                 host = "localhost", 
                 dbname = "amphibians_v2", 
                 port = 25432, 
                 user = "rknapp",
                 password = getPass::getPass("PostgreSQL password for rknapp: "))

# Set default schema
dbExecute(con, "SET search_path TO field_surveys")
```

## Retrieve data subsets from PostgreSQL database and Fulcrum

### Survey data

Only CMR surveys are typically conducted at this site, so no need to filter by cmr and ves. 

```{r}
survey_50783 <- tbl(con, "visit") |>
  filter(site_id %in% c(50783)) |>
  inner_join(tbl(con, "survey"), by = "visit_id") |>
  filter(survey_type %in% c("cmr")) |>
  select(
    site_id,
    visit_date,
    visit_status,
    survey_type,
    description,
    survey_quality,
    wind,
    sky,
    air_temp,
    duration,
    visit_comment,
    survey_comment
  ) |>
  collect() |> 
  arrange(visit_date, site_id)
```

Review the survey data subset for potential problems. 
- Survey on 2022-09-18 has survey_type == cmr, but notes indicate that this survey was for ribbtr sampling only. Drop. 

### Drop non-CMR survey

```{r}
survey_50783 <- survey_50783 |> 
  filter_out(visit_date == "2022-09-18")
```

### Capture data

```{r}
capture_50783 <- tbl(con, "visit") |>
  filter(site_id %in% c(50783)) |>
  inner_join(tbl(con, "survey"), by = "visit_id") |>
  filter(survey_type == "cmr") |>
  inner_join(tbl(con, "capture_survey"), by = "survey_id") |>
  select(
    site_id,
    visit_date,
    visit_status,
    survey_type,
    pit_tag_ref,
    tag_new,
    species,
    capture_life_stage,
    capture_animal_state,
    length,
    capture_survey_comment
  ) |>
  collect() |> 
  arrange(visit_date) |> 
  rename(capture_comment = capture_survey_comment)
```

Review the capture data subset for potential problems. 

### Remove frogs collected during non-CMR RIBBiTR survey

```{r}
capture_50783 <- capture_50783 |> 
  filter_out(visit_date == "2022-09-18")
```

### Disconnect from database 

```{r db_disconnect}
dbDisconnect(con)
rm(con)
```

- Close SSH connection 

### Create current-year survey and capture datasets for appending

```{r}
visit_50783_2026 <- read_csv(here("data", "raw", "current_year", "survey_14aug2026.csv")) |> 
  filter(site_id == 50783) |> 
  select(fulcrum_id, site_id, visit_date, visit_status, visit_comment) |> 
  rename(visit_id = fulcrum_id)
  
survey_50783_2026 <- read_csv(here("data", "raw", "current_year", "survey_description_14aug2026.csv")) |> 
  select(fulcrum_record_id, fulcrum_id, survey_type, survey_description, wind, sky, air_temp, duration, survey_quality, general_survey_comment) |> 
  filter(survey_type == "cmr") |> 
  rename(description = survey_description, 
         survey_comment = general_survey_comment,
         survey_id = fulcrum_id,
         visit_id = fulcrum_record_id) |> 
inner_join(visit_50783_2026 |> 
             select(visit_id), 
           by = "visit_id")

capture_50783_2026 <- read_csv(here("data", "raw", "current_year", "survey_description_amphibian_capture_14aug2026.csv")) |> 
  select(fulcrum_parent_id, method, pit_tag_ref, tag_new, species, capture_life_stage, capture_animal_state, length, capture_comment) |> 
  filter(method == "cmr") |> 
  rename(survey_id = fulcrum_parent_id) |> 
  inner_join(survey_50783_2026 |> 
               select(survey_id), 
             by = "survey_id")

survey_append_50783 <- visit_50783_2026 |> 
  inner_join(survey_50783_2026, by = "visit_id")

capture_append_50783 <- visit_50783_2026 |> 
  inner_join(survey_50783_2026, by = "visit_id") |> 
  inner_join(capture_50783_2026, by = "survey_id") |> 
  select(-visit_id, -survey_id, -description, -survey_quality, -method, -wind, -sky, -air_temp, -duration, -visit_comment, -survey_comment) |> 
  replace_na(list(species = "ramu", 
                  capture_life_stage = "adult")) |> 
  mutate(pit_tag_ref = as.character(pit_tag_ref))

survey_append_50783 <- survey_append_50783 |> 
  select(-visit_id, -survey_id)

```

### Append current year data to time series

```{r}
survey_50783 <- bind_rows(survey_50783, survey_append_50783)

capture_50783 <- bind_rows(capture_50783, capture_append_50783)
```

## Save raw data files

```{r}
survey_50783 |> 
  write_csv(here::here("data", "raw", "50783_survey_raw.csv"))

capture_50783 |> 
  write_csv(here::here("data", "raw", "50783_capture_raw.csv"))
```

## Read in raw data files (if necessary)

```{r}
survey_50783 <- read_csv(here("data", "raw", "50783_survey_raw.csv"))

capture_50783 <- read_csv(here("data", "raw", "50783_capture_raw.csv")) |> 
  mutate(pit_tag_ref = as.character(pit_tag_ref))
```

## Clean survey dataset

<!-- ### Review cmr survey history for all sites in project area -->

<!-- ```{r}  -->
<!-- cmr_survey_check <- survey_50783 |>  -->
<!--   filter(survey_type == "cmr", -->
<!--          visit_date > min(reintroduction_50783$visit_date)) |>  -->
<!--   distinct(visit_date) |> -->
<!--   arrange(visit_date) |> -->
<!--   mutate(primary_period = cumsum(visit_date - lag(visit_date, default = first(visit_date)) > 5) + 1) |> # Identify primary periods based on a gap between CMR surveys of > 5 days -->
<!--   inner_join(survey_50783, by = "visit_date") |>  -->
<!--   filter(survey_type == "cmr") |>  -->
<!--   select(site_id, primary_period) |>  -->
<!--   group_by(primary_period) |>  -->
<!--   count(site_id) |>  -->
<!--   mutate(num_sec_periods = max(n), # of times each site was visited during primary period -->
<!--          diff = num_sec_periods - n) |> # difference between number of secondary periods and the number of visits to each site -->
<!--   relocate(site_id) |>  -->
<!--   arrange(site_id, primary_period) -->
<!-- ``` -->

<!-- ```{r} -->
<!-- rm(cmr_survey_check) -->
<!-- ``` -->

<!-- - Primary site 50783 surveyed during each secondary period. No adjacent sites to include. -->

<!-- ### Identify VES with no associated CMR period for possible inclusion as a CMR survey - customize for each site -->

<!-- ```{r} -->
<!-- # For each primary period, create all dates in period +/- 2 days -->
<!-- cmr_dates <- survey_50783 |>  -->
<!--   filter(survey_type == "cmr", -->
<!--          visit_date > min(reintroduction_50783$visit_date) + days(5)) |> # Remove CMR surveys from before or soon after frogs were first reintroduced -->
<!--   distinct(visit_date) |> -->
<!--   arrange(visit_date) |> -->
<!--   mutate(primary_period = cumsum(visit_date - lag(visit_date, default = first(visit_date)) > 5) + 1) |> # Identify primary periods based on a gap between CMR surveys of > 1 day -->
<!--   group_by(primary_period) |> -->
<!--   summarise(  # buffer primary periods by 2 days at start and end -->
<!--     start = min(visit_date) - days(2), -->
<!--     end   = max(visit_date) + days(2), -->
<!--     .groups = "drop" -->
<!--   ) |> -->
<!--   rowwise() |>  -->
<!--   mutate(visit_date = list(seq.Date(start, end, by = "day"))) |>  # Create all dates in each expanded period -->
<!--   unnest(visit_date) |> -->
<!--   select(primary_period, visit_date) -->

<!-- # Select VES dates -->
<!-- ves_dates <- survey_50783 |> -->
<!--   filter( -->
<!--     survey_type == "visual", -->
<!--     visit_date > min(reintroduction_50783$visit_date + days(5)), # Remove VES surveys from before or soon after frogs were first reintroduced -->
<!--     visit_date <= max(cmr_dates$visit_date) -->
<!--   ) |>  -->
<!--   select(visit_date) |>  -->
<!--   arrange(visit_date) -->

<!-- # Identify all visual surveys that fall outside of CMR primary periods (expanded) -->
<!-- unmatched_ves <- ves_dates |>  -->
<!--   anti_join(cmr_dates, by = "visit_date") |>  -->
<!--   distinct(visit_date) -->
<!-- ``` -->

<!-- - No unmatched VES. Include only CMR suyrveys. -->

<!-- ### Remove unneeded objects -->

<!-- ```{r} -->
<!-- rm(cmr_dates, ves_dates, unmatched_ves) -->
<!-- ``` -->

<!-- ### Remove irrelevant surveys from survey records -->

<!-- ```{r} -->
<!-- survey_50783_mrmr <- survey_50783 |>  -->
<!--   filter(survey_type == "cmr",  -->
<!--          visit_date > min(reintroduction_50783$visit_date)) |>  -->
<!--   rename(survey_date = visit_date) -->
<!-- ``` -->

- Review records for potential problematic surveys, for example, based on comment fields and fields describing current conditions.
- Some non-ideal survey conditions, opted to retain all surveys. 

### Select unique survey dates

```{r}
survey_50783_mrmr <- survey_50783 |> 
  rename(survey_date = visit_date) |> 
  distinct(site_id, survey_date, survey_type, visit_status)
```

### Survey data checks

```{r}
# Check for duplicate survey dates
see_if(!any(survey_50783_mrmr |> count(survey_date) |> pull(n) > 1))

# Check for sites that were unsuitable for survey
see_if(all(survey_50783_mrmr$visit_status == "suitable"))

# Check for non-cmr and non-reintroduction survey types
see_if(all(survey_50783_mrmr$survey_type == "cmr"))
```

### Assign primary and secondary periods

```{r}
survey_50783_mrmr <- survey_50783_mrmr |> 
  arrange(survey_date) |> 
  # Create a flag for breaks between surveys: gap > 4 days
  mutate(gap_break = (survey_date - lag(survey_date, default = first(survey_date)) > 4),
    primary_period = cumsum(gap_break) + 1) |>   
  # Assign secondary periods within each primary period
  mutate(secondary_period = row_number(), .by = primary_period) |>
  select(-gap_break, -survey_type, -visit_status)
```

### Save cleaned survey file

```{r}
survey_50783_mrmr |> write_csv(here("data", "clean", "50783_survey.csv"))
```

## Clean capture dataset

### Capture data checks

```{r}
# Verify that all dates in capture are present in survey 
see_if(nrow(capture_50783 |> anti_join(survey_50783_mrmr, by = join_by(visit_date == survey_date))) == 0)

# Verify that visit_status is "suitable" for all records
see_if(all(capture_50783$visit_status == "suitable"))

# Verify that survey_type is "cmr" for all records
see_if(all(capture_50783$survey_type == "cmr"))

# Check for records where pit_tag_ref is NULL
see_if(noNA(capture_50783$pit_tag_ref))

# Check for records containing pit tags shorter than 15 characters (or 9 for early sites)
see_if(!any(nchar(capture_50783$pit_tag_ref) != 15))

# Check for records containing pit tags that end with ".1" (known to be potentially incorrect) 
see_if(!any(str_detect(capture_50783$pit_tag_ref, "\\.1$")))

# Check for records where species != rasi
see_if(all(capture_50783$species == "ramu"))

# Check for records where capture_animal_state is NULL
see_if(noNA(capture_50783$capture_animal_state))

# Check for records where capture_animal_state == "dead"
see_if(!any(capture_50783$capture_animal_state == "dead"))

# Check for records of subadult frogs 
see_if(all(capture_50783$length >= 40, na.rm = TRUE)) 

# Check for duplicate pit tags on each date
see_if(nrow(capture_50783 |> count(visit_date, pit_tag_ref) |> filter(n > 1)) == 0)
```

### Clean capture data

```{r}
# Remove frogs with pit_tag_ref == NA
capture_50783 <- capture_50783 |> 
  filter_out(is.na(pit_tag_ref))

# Remove pit_tag_refs entered twice on same day
count <- capture_50783 |> 
  count(visit_date, pit_tag_ref)

capture_50783 <- capture_50783 |> 
  inner_join(count, by = c("visit_date", "pit_tag_ref")) |> 
  mutate(capture_comment = replace_na(capture_comment, "")) |> 
  filter_out(n > 1 & capture_comment == "") |> 
  select(-n)
rm(count)

# Remove record with incorrect pit_tag_ref (based on comments)
capture_50783 <- capture_50783 %>% 
  filter(pit_tag_ref != '900067000117595')

# Fix incorrect pit_tag_ref (based on comments and data review)
capture_50783 <- capture_50783 |> 
  mutate(pit_tag_ref = replace_values(pit_tag_ref, "900067000114249" ~ "900043000144249"))

```

### Create final capture dataset formatted for mrmr

```{r}
capture_50783_mrmr <- capture_50783 |>
  rename(survey_date = visit_date, 
         pit_tag_id = pit_tag_ref) |>
  select(site_id, survey_date, pit_tag_id) |> 
  arrange(survey_date, pit_tag_id)
```

### Save cleaned capture data (one file with Bd load and one without)

```{r}
capture_50783_mrmr |> write_csv(here("data", "clean", "50783_capture.csv"))
```





