
---
title: "sample_run_54188"
output: html_notebook
date: "2025-02-13"
---

## Introduction

This notebook analyzes capture-mark-recapture data using the *mrmr* and *mrmrmr* packages. *mrmrmr* provides access to an adapted version of the model deployed by *mrmr*.

Added functionality in *mrmrmr* includes:

1.  A random effects parameter for detection probabilities, which accounts for unobserved sources of variation in detection over each survey session.

2.  Calculation of average survival probability of all frogs between primary periods in the generated quantities block.

3.  Calculation of time-varying detection probabilities for each session in the generated quantities block.

4.  Calculation of the log likelihood of each individual's capture history in the generated quantities block, which is used to quantify the model's predictive accuracy and is necessary for model comparison methods.

## Install the mrmr and mrmrmr packages

```{r}

# install cmdstanr and cmdstan
remotes::install_github("stan-dev/cmdstanr")
cmdstanr::install_cmdstan(overwrite = TRUE)

# install mrmr and mrmrmr
remotes::install_github("SNARL1/mrmr")
remotes::install_github("john-imperato/mrmrmr")

```

## Load required packages

```{r}

library(dplyr)
library(readr)
library(tibble)
library(cmdstanr)
library(mrmr)
library(mrmrmr)
library(patchwork)
library(bayesplot)

```

## Create required data sets

```{r}

captures <- read_csv(here::here("data", "54188_capture.csv"))
surveys <- read_csv(here::here("data", "54188_survey.csv"))

```

## Clean data with mrmr

```{r}

# clean data with the mrmr function
data_54188_TVD <- mrmr::clean_data(captures, surveys)

```

## Fit mrmrmr model with session-specific detection probabilities

```{r}

# fit model and assign to object
model_54188_TVD <- mrmrmr::fit_model_TVD(data = data_54188_TVD,
                                            chains = 4,
                                            parallel_chains = 4,
                                            threads_per_chain = 2,
                                            adapt_delta = 0.99)

# save model as .rds file
write_rds(model_54188_TVD, here::here("sample_run", "54188_model_TVD.rds"))

```

## Save the model summary to an object

```{r}

summary_54188 <- model_54188_TVD$m_fit$summary()

```

## Diagnose potential non-convergence of the MCMC algorithm

-   Inspect traceplots and parameter summary table. In table, are any R~hat~ values $\ge 1.01$?

```{r model-diagnostics}
pars_to_plot <- c('alpha_lambda', 
                  'sigma_lambda', 
                  'beta_phi', 
                  'sigma_phi', 
                  'beta_detect')
bayesplot::mcmc_trace(model_54188_TVD$m_fit$draws(pars_to_plot))

model_54188_TVD$m_fit$summary(pars_to_plot)

```

## Print parameters

```{r}

# abundance
abundance_54188 <- subset(summary_54188, grepl("N\\[[0-9]+\\]", variable))

# recruitment  
recruitment_54188 <- subset(summary_54188, grepl("B\\[[0-9]+\\]", variable))

# detection
detection_54188 <- subset(summary_54188, grepl("p\\[[0-9]+\\]", variable))

# average survival
avg.survival_54188 <- subset(summary_54188, grepl("overall_phi\\[[0-9]+\\]", variable))

print(abundance_54188)
print(recruitment_54188)
print(detection_54188)
print(avg.survival_54188)

```
