---
title: "70397 - Analyze capture-mark-recapture data using *mrmr*"
author: "Roland Knapp"
output: html_notebook
editor_options: 
  chunk_output_type: inline
---
For the site 70397, aka East Evelyn Pond, this notebook analyzes capture-mark-recapture data using the *mrmr* package. The code creates the required survey and capture data frames from saved CSV files (created by 70397_createdata_mrmr.Rmd), checks the datasets, fits the model, evaluates potential non-convergence of the MCMC algorithm, and visualizes model results. 

Before using the code in this file, replace all instances of "70397" with the 5-digit site_id of interest (70397)

## Install latest versions of cmdstanr, cmdstan, and mrmr packages

- Run only if updated versions were released since previous install.  
```{r install-stan-mrmr-packages}
# install cmdstanr and cmdstan
install.packages("cmdstanr", repos = c("https://mc-stan.org/r-packages/", getOption("repos")))
cmdstanr::install_cmdstan()

# install mrmr
remotes::install_github("SNARL1/mrmr")
```

## Load required packages

```{r load-packages}
library(dplyr)
library(readr)
library(tibble)
library(cmdstanr)
library(mrmr)
library(patchwork)
```

## Create required datasets

```{r create-data-objects}
captures <- read_csv(here::here("data", "clean", "70397_capture.csv"))
surveys <- read_csv(here::here("data", "clean", "70397_survey.csv")) %>% 
  rename(survey_date = visit_date)
# translocations <- read_csv(here::here("data", "clean", "70397_translocation.csv")) # no translocations conducted at this site as of 2025
```

## Check datasets

```{r check-data}
data_70397 <- clean_data(captures = captures, surveys = surveys)
```

## Fit model

- Code assumes that there are 8 physical cores available. Update parallel_chains and threads_per_chain as necessary based on actual number of cores.  
- Add "adapt_delta = 0.99" or similar (after "treats_per_chain") if divergent transitions occur. 

```{r fit-model}
model_70397 <- fit_model(
                    data = data_70397, 
                    chains = 4, 
                    parallel_chains = 4, 
                    cpp_options = list(stan_threads = TRUE),
                    threads_per_chain = 2, 
                    adapt_delta = 0.01) 
```

- Running model with default adapt_delta (0.8) resulted in zero divergent transition (0/4000).  

### Save model object

- This step prevents having to rerun the model in the event of a crash during model visualizations in subsequent code chunks.  
- Model files can be large, so this saves the file in a compressed format.  

```{r write-model}
write_rds(model_70397, here::here("out", "model", "70397_model.rds"), "xz", compression = 9L)
```

- To read model file, use "model_70397 <- read_rds(here::here("out", "model", "70397_model.rds"))"

### Diagnose potential non-convergence of the MCMC algorithm

- Inspect traceplots and parameter summary table. In table, are any R~hat~ values $\ge 1.01$?

```{r model-diagnostics}
pars_to_plot <- c('alpha_lambda', 
                  'sigma_lambda', 
                  'beta_phi', 
                  'sigma_phi', 
                  'beta_detect')
bayesplot::mcmc_trace(model_70397$m_fit$draws(pars_to_plot))

model_70397$m_fit$summary(pars_to_plot)
```

## Visualize model results

### Default plots

- Custom plots (produced in next code chunk: "plots_custom") are often easier to interpret than the default plots produced by this code chunk.  

```{r plots-default}
# plot_model(model_70397, what = "abundance")
# plot_model(model_70397, what = "recruitment")
# plot_model(model_70397, what = "survival")
```

### Custom plots

- Remove p1 if translocations/reintroductions are not relevant.  

```{r plots-custom, fig.height=8, fig.width=7}
library(ggplot2)
p1 <- plot_model(model_70397, what = "survival") 
p1 <- p1 + ggtitle("A. Survival of translocated frogs") +
    expand_limits(y = 0) +
    theme(legend.position = "none")
  
p2 <- plot_model(model_70397, what = "abundance") 
p2 <- p2 + ggtitle("B. Population size") +
    ylab("Number of adults") + 
    expand_limits(y = 0)
  
p3 <- plot_model(model_70397, what = "recruitment")
p3 <- p3 + ggtitle("C. Recruitment") +
    ylab("Number of adults")
  
p1 / p2 / p3
```

### Save custom plots

```{r save-plot}
ggsave(here::here("out", "plots", "70397_mrmr_plots.png"), width = 6.5, height = 7.2)
```

## Create survival tables

- Calculated for translocated/reintroduced frogs only, so skip if translocations/reintroductions are not relevant. 
- Only the cohort-level survival table is displayed because the individual-level survival table can be large.  

```{r survival-tables}
survival_cohort <- survival_table(model_70397, by_cohort = TRUE)
survival_cohort

survival_ind <- survival_table(model_70397, by_individual = TRUE) 
```

### Save survival tables

```{r save-survival-tables}
survival_cohort %>% write_csv(here::here("out", "tables", "70397_survival_cohort.csv"))
survival_ind %>% write_csv(here::here("out", "tables", "70397_survival_ind.csv"))
```

