---
title: "70556 - Create datasets for analysis using *mrmr*"
author: "Roland Knapp"
output: html_notebook
editor_options: 
  chunk_output_type: inline
---

This notebook creates the survey, capture, and translocation files for use in analyzing capture-mark-recapture data using the *mrmr* package. The code retrieves data from the PostgreSQL amphibians database, runs error checks, and creates the relevant files in *mrmr* format. 

Note that data from the most recent field season may not yet be appended to tables in the amphibians database. If these recently-collected data are necessary (e.g., for preliminary analyses), complete datasets can be created by copying code chunks in "fulcrum_data_append.Rmd" into this notebook and executing all code chunks. 

## Load packages
```{r load_packages}
library(tidyverse)
library(RPostgreSQL)
library(lubridate)
library(assertthat)
library(assertr)
options(readr.show_col_types = FALSE)
```

## Create translocation dataset
Not all sites have received frog translocations. Run or skip the following code chunks as necessary. 

### Connect to PostgreSQL database. 

Users without credentials can use CSV file created in code chunk "read_raw_translocation_data" below as a starting point. 
```{r}
con = dbConnect(dbDriver("PostgreSQL"),
                host = Sys.getenv("host"), 
                port = Sys.getenv("port"), 
                dbname = Sys.getenv("dbname"),
                user = rstudioapi::askForPassword("user name"),
                password = rstudioapi::askForPassword("password")
               )
```

### Retrieve translocation data from database
```{sql retrieve_translocation_data, connection=con, output.var = "translocation_70556"}
select 
  release_siteid1, 
  release_date, 
  type, 
  pit_tag_ref, 
  tag_new,
  sex,
  length,
  weight,
  swab_id,
  relocate.comment as comment_relocate,
  relocate_frog.comment as comment_relocate_frog,
  relocate.id as relocate_id,
  relocate_frog.id as relocate_frog_id
from relocate
inner join relocate_frog on relocate.id = relocate_id
where release_siteid1 = 70556 and
      type = 'translocation'
```
```{r}
dbDisconnect(con)
rm(con)
```
Review the translocation data subset for potential problems. 

### Save/retrieve raw translocation data
```{r save_raw_translocation_data, message=FALSE}
translocation_70556 %>% write_csv(here::here("data", "raw", "70556_translocation_raw.csv"))
```
```{r read_raw_translocation_data, message=FALSE}
translocation_70556 <- read_csv(here::here("data", "raw", "70556_translocation_raw.csv")) %>% 
  mutate(pit_tag_ref = as.character(pit_tag_ref))
```

### Retain only relevant columns
```{r translocation_relevant_columns}
translocation_70556 <- translocation_70556 %>% 
  rename(site_id = release_siteid1) %>% 
  select(site_id, release_date, type, pit_tag_ref, tag_new, sex, length, weight, swab_id)
```

### Translocation data checks
Run checks before and after data cleaning (next code chunk) to ensure that all issues have been resolved. Run lines of code one at a time to allow easy evaluation of messages and outputs. 
```{r translocation_data_checks, message=FALSE}
# Check for records where pit_tag_ref is NULL
assert_that(noNA(translocation_70556$pit_tag_ref))

# Check for records containing pit tags shorter than 15 characters
assert_that(!any(nchar(translocation_70556$pit_tag_ref) != 15))

# Check for records containing pit tags that end with ".1" (known to be potentially incorrect) 
assert_that(!any(str_detect(translocation_70556$pit_tag_ref, "\\.1$")))

# Check for records of subadult frogs
translocation_70556 %>% drop_na(length) %>% filter(length < 40)

# Check for duplicate pit tags on each date
translocation_70556 %>% count(release_date, pit_tag_ref) %>% filter(n > 1)

# Are release dates correct?
translocation_70556 %>% distinct(release_date) %>% glimpse()
```

### Clean translocation data
```{r clean_translocation_data}
# No problems to clean
```

### Create final translocation dataset formatted for mrmr
```{r final_translocation_data}
translocation_70556_mrmr <- translocation_70556 %>% 
  rename(pit_tag_id = pit_tag_ref) %>% 
  arrange(release_date, pit_tag_id) %>% 
  select(site_id, type, release_date, pit_tag_id)
```

### Save/retrieve cleaned translocation data
```{r save_cleaned_translocation_data, message=FALSE}
translocation_70556_mrmr %>% write_csv(here::here("data", "clean", "70556_translocation.csv"))
```
```{r read_cleaned_translocation_data, message=FALSE}
translocation_70556 <- read_csv(here::here("data", "clean", "70556_translocation.csv")) %>% 
  mutate(pit_tag_id = as.character(pit_tag_id))
```

## Create survey dataset

### Connect to PostgreSQL database. 

Users without credentials can use CSV file created in code chunk "read_raw_survey_data" below as a starting point. 
```{r}
con = dbConnect(dbDriver("PostgreSQL"),
                host = Sys.getenv("host"), 
                port = Sys.getenv("port"), 
                dbname = Sys.getenv("dbname"),
                user = rstudioapi::askForPassword("user name"),
                password = rstudioapi::askForPassword("password")
               )
```

### Retrieve survey data from database
```{sql retrieve_survey_data, connection=con, output.var = "survey_70556"}
select 
  site_id, 
  visit_date, 
  visit_status, 
  survey_type, 
  description, 
  survey_quality,
  wind,
  sky,
  duration,
  visit.comment as comment_visit,
  survey.comment as comment_survey
from visit
inner join survey on visit.id = visit_id
where site_id = 70556 and
      survey_type = 'cmr'
```
```{r, message=FALSE}
dbDisconnect(con)
rm(con)
```

### Save/retrieve raw survey data
```{r save_raw_survey_data, message=FALSE}
survey_70556 %>% write_csv(here::here("data", "raw", "70556_survey_raw.csv"))
```
```{r read_raw_survey_data, message=FALSE}
survey_70556 <- read_csv(here::here("data", "raw", "70556_survey_raw.csv"))
```

### Include only one survey for entire site on each survey date
70556 is the primary site and was surveyed during each secondary period. 
```{r}
survey_70556 <- survey_70556 %>% 
  filter(site_id == 70556) %>%
  rename(survey_date = visit_date)
```

### Add translocation dates as surveys
To meet mrmr format requirement
```{r add_translocation_surveys}
survey_70556_mrmr <- translocation_70556 %>% 
  distinct(site_id, release_date, type) %>% 
  rename(survey_type = type, survey_date = release_date) %>% bind_rows(survey_70556) %>% 
  arrange(survey_date) 
```
Review records for potential problematic surveys, for example, based based on comment fields and fields describing current conditions.

### Survey data checks
Run checks before and after data cleaning (next code chunk) to ensure that all issues have been resolved.
```{r survey_data_checks}
# Check for duplicate survey dates
survey_70556_mrmr %>% count(site_id, survey_date) %>% filter(n > 1)

# Check for sites that were unsuitable for survey
survey_70556_mrmr %>% filter(visit_status != "suitable")
```

### Clean survey data
```{r clean_survey_data}
# No problems to clean
```

### Create primary and secondary period columns in mrmr format
```{r create_period_columns}
survey_70556_mrmr <- survey_70556_mrmr %>%
  select(site_id, survey_date, survey_type) %>% 
  mutate(primary_period = case_when(survey_date == "2018-08-16" ~ 1, 
                                    between(survey_date, ymd("2019-08-24"), ymd("2019-08-26")) ~ 2,
                                    between(survey_date, ymd("2020-07-07"), ymd("2020-07-08")) ~ 3,
                                    survey_date == "2020-07-09" ~ 4, 
                                    between(survey_date, ymd("2020-08-23"), ymd("2020-08-25")) ~ 5,
                                    between(survey_date, ymd("2021-07-22"), ymd("2021-07-24")) ~ 6,
                                    between(survey_date, ymd("2021-08-26"), ymd("2021-08-28")) ~ 7)) 
                                    
survey_70556_mrmr <- survey_70556_mrmr %>% 
  arrange(primary_period, survey_date) %>% 
  group_by(primary_period) %>% 
  mutate(secondary_period = seq_len(n()),
         secondary_period = replace(secondary_period, survey_type == "translocation", 0))
                                    
```

### Save/retrieve cleaned survey file
```{r save_cleaned_survey_data}
survey_70556_mrmr %>% write_csv(here::here("data", "clean", "70556_survey.csv"))
```
```{r read_cleaned_survey_data}
survey_70556 <- read_csv(here::here("data", "clean", "70556_survey.csv"))
```

## Create capture dataset

### Connect to PostgreSQL database. 

Users without credentials can use CSV file created in code chunk "read_raw_capture_data" below as a starting point. 
```{r}
con = dbConnect(dbDriver("PostgreSQL"),
                host = Sys.getenv("host"), 
                port = Sys.getenv("port"), 
                dbname = Sys.getenv("dbname"),
                user = rstudioapi::askForPassword("user name"),
                password = rstudioapi::askForPassword("password")
               )
```

### Retrieve frog capture data from database
```{sql retrieve_capture_data, connection=con, output.var = "capture_70556"}
select 
  site_id, 
  visit_date, 
  visit_status, 
  survey_type, 
  pit_tag_ref, 
  tag_new, 
  species, 
  capture_life_stage, 
  capture_animal_state, 
  length, 
  visit.comment as comment_visit,
  survey.comment as comment_survey,
  capture_survey.comment as comment_capture,
  visit.id as visit_id,
  survey.id as survey_id,
  capture_survey.id as capture_id
from visit
inner join survey on visit.id = visit_id
inner join capture_survey on survey.id = survey_id
where site_id = 70556 and
      survey_type = 'cmr'
```
```{r, message=FALSE}
dbDisconnect(con)
rm(con)
```

### Save/retrieve raw capture data
```{r save_raw_capture_data, message=FALSE}
capture_70556 %>% write_csv(here::here("data", "raw", "70556_capture_raw.csv"))
```
```{r read_raw_capture_data, message=FALSE}
capture_70556 <- read_csv(here::here("data", "raw", "70556_capture_raw.csv")) %>% 
  mutate(pit_tag_ref = as.character(pit_tag_ref))
```
Review records for potential problematic captures, for example, based on comment fields and fields describing frog characteristics.

### Retain only relevant columns
```{r capture_relevant_columns}
capture_70556 <- capture_70556 %>% 
  select(-visit_id, -survey_id, -capture_id)
```

### Capture data checks
Run checks before and after data cleaning)
```{r capture_data_checks, message=FALSE}
# Check for records where pit_tag_ref is NULL
assert_that(noNA(capture_70556$pit_tag_ref))

# Check for records containing pit tags shorter than 15 characters
assert_that(!any(nchar(capture_70556$pit_tag_ref) != 15))

# Check for records containing pit tags that end with ".1" (known to be potentially incorrect) 
assert_that(!any(str_detect(capture_70556$pit_tag_ref, "\\.1$")))

# Check for records where species != ramu
capture_70556 %>% drop_na(species) %>% filter(species != "ramu")

# Check for records were survey_type = swab
assert_that(!any(capture_70556$survey_type == "swab"))

# Check for records where capture_animal_state is NULL
assert_that(noNA(capture_70556$capture_animal_state))

# Check for records where capture_animal_state == "dead"
assert_that(!any(capture_70556$capture_animal_state == "dead"))

# Check for records of subadult frogs
capture_70556 %>% drop_na(length) %>% filter(length < 40)

# Check for sites that were unsuitable for survey
assert_that(!any(capture_70556$visit_status != "suitable"))

# Check for duplicate capture records on each date
capture_70556 %>% count(visit_date, pit_tag_ref) %>% filter(n > 1)

# Check for captures that lack associated surveys
survey_70556 %>% 
  select(survey_date) %>% 
  rename(visit_date = survey_date) %>% 
  right_join(capture_70556, by = c("visit_date")) %>% 
  filter(is.na(visit_date))

# Check for surveys that lack associated captures
survey_70556 %>% 
  select(survey_date) %>% 
  rename(visit_date = survey_date) %>% 
  right_join(capture_70556, by = c("visit_date")) %>% 
  filter(is.na(site_id))
```

### Clean capture data
```{r clean_capture_data}
# No problems to clean
```

### Create final capture dataset formatted for mrmr
```{r final_capture_data}
capture_70556_mrmr <- capture_70556 %>%
  rename(survey_date = visit_date, pit_tag_id = pit_tag_ref) %>%
  arrange(survey_date, pit_tag_id) %>%
  select(site_id, survey_date, pit_tag_id)
```

### Save/retrieve cleaned capture data
```{r save_cleaned_capture_data}
capture_70556_mrmr %>% write_csv(here::here("data", "clean", "70556_capture.csv"))
```
```{r read_cleaned_capture_data}
capture_70556 <- read_csv(here::here("data", "clean", "70556_capture.csv"))
```







