---
title: "74976 - Analyze capture-mark-recapture data using *mrmr*"
author: "Roland Knapp"
output: html_notebook
editor_options: 
  chunk_output_type: inline
---
This notebook analyzes capture-mark-recapture data using the *mrmr* package. The code creates the required survey, capture, and translocation data frames from saved CSV files (created by xxxxx_createdata_mrmr.Rmd), checks the datasets, fits the model, evaluates potential non-convergence of the MCMC algorithm, and visualizes model results. 

## Install latest versions of cmdstanr, cmdstan, and mrmr packages
Run only if updated versions were released since previous install
```{r install_stan_packages, message=FALSE, warning=FALSE}
# install cmdstanr and cmdstan
install.packages("cmdstanr", repos = c("https://mc-stan.org/r-packages/", getOption("repos")))
cmdstanr::install_cmdstan()

# install mrmr
remotes::install_github("SNARL1/mrmr")
```

## Load required packages
```{r load_packages}
library(dplyr)
library(readr)
library(tibble)
library(cmdstanr)
library(mrmr)
library(patchwork)
```

## Create required data frames
```{r create_data_objects}
captures <- read_csv(here::here("data", "clean", "74976_capture.csv"))
surveys <- read_csv(here::here("data", "clean", "74976_survey.csv"))
translocations <- read_csv(here::here("data", "clean", "74976_translocation.csv"))
```

## Check datasets
```{r check_data}
data_74976 <- clean_data(captures = captures, surveys = surveys, translocations = translocations)
```

## Fit model
```{r fit_model}
model_74976 <- fit_model(
                    data = data_74976, 
                    chains = 4, 
                    parallel_chains = 4, 
                    cpp_options = list(stan_threads = TRUE),
                    threads_per_chain = 2, 
                    adapt_delta = 0.99) # comment out this line unless needed to resolve divergent transitions
```
* Running model with default adapt_delta (0.8) resulted in one divergent transition (1/4000).
* Increase of adapt_delta to 0.99 resulted in zero divergent transitions. 

### Save/retrieve model object
This step prevents having to rerun the model in the event of a crash during model visualizations in subsequent code blocks. Model files can be large, so this saves the file in a compressed format.
```{r write_model}
write_rds(model_74976, here::here("out", "model", "74976_model.rds"), "xz", compression = 9L)
```
```{r read_model}
model_74976 <- read_rds(here::here("out", "model", "74976_model.rds"))
```

### Diagnose potential non-convergence of the MCMC algorithm
Inspect traceplots and parameter summary table. In table, are any R~hat~ values $\ge 1.01$?
```{r model_diagnostics}
pars_to_plot <- c('alpha_lambda', 
                  'sigma_lambda', 
                  'beta_phi', 
                  'sigma_phi', 
                  'beta_detect')
bayesplot::mcmc_trace(model_74976$m_fit$draws(pars_to_plot))

model_74976$m_fit$summary(pars_to_plot)
```

## Visualize model results
### Default plots
Note: Custom plots (produced in next code chunk: "plots_custom") are often easier to interpret than the default plots produced by this code chunk.
```{r model_visualize}
plot_model(model_74976, what = "abundance")
plot_model(model_74976, what = "recruitment")
plot_model(model_74976, what = "survival")
```

### Customize plots
```{r plots_custom, fig.height=8, fig.width=7}
library(ggplot2)
p1 <- plot_model(model_74976, what = "survival") 
p1 <- p1 + ggtitle("A. Survival of translocated frogs") +
    expand_limits(y = 0) +
    theme(legend.position = "none")
  
p2 <- plot_model(model_74976, what = "abundance") 
p2 <- p2 + ggtitle("B. Population size") +
    ylab("Number of adults") + 
    expand_limits(y = 0)
  
p3 <- plot_model(model_74976, what = "recruitment")
p3 <- p3 + ggtitle("C. Recruitment") +
    ylab("Number of adults")
  
p1 / p2 / p3
```

### Save custom plots
```{r save_plot}
ggsave(here::here("out", "figures", "74976_mrmr_plots.png"), width = 6.5, height = 7.2)
```

## Create survival tables
Only the cohort-level survival table is displayed (individual-level survival can be a large table).
```{r survival_tables}
survival_cohort <- survival_table(model_74976, by_cohort = TRUE)
survival_cohort

survival_ind <- survival_table(model_74976, by_individual = TRUE) 
```

### Save survival tables
```{r save_survival_tables}
survival_cohort %>% write_csv(here::here("out", "tables", "74976_survival_cohort.csv"))
survival_ind %>% write_csv(here::here("out", "tables", "74976_survival_ind.csv"))
```
