---
title: "Methods to explore the Bayesian posterior"
author: "Roland A. Knapp"
format: html
editor: source
---

## Load packages

```{r load_packages}
if (!require(librarian)){
  install.packages("librarian")
  library(librarian)
}
librarian::shelf(cmdstanr, mrmr, readr, here, posterior, tidyverse, reshape2, abind)
```

## Explore posterior

```{r}
library(posterior)
library(dplyr)

model_22008 <- read_rds(here("out", "model", "22008_model_2groups.rds"))

class(model_22008)
str(model_22008)

# 3D array: iterations × chains × parameters
draws_array <- model_22008$m_fit$draws()

# Check dimensions
dim(draws_array)

# Extract only parameters for survival coefficients (\beta_phi[1], beta_phi[2], etc.)
surv_draws <- model_22008$m_fit$draws(variables = c("beta_phi"))

# Convert to tidy data frame
post_df <- as_draws_df(surv_draws)

colnames(post_df)

# Summarize specific posterior parameters

posterior::summarise_draws(post_df %>% select("beta_phi[1]"))

# Extract all parameter names from the CmdStanMCMC fit (if there are many parameters, use next line of code for summary)
all_params <- posterior::variables(model_22008$m_fit$draws())

# Extract prefix (everything before "[") for grouping
param_df <- tibble(parameter = all_params) %>%
  mutate(prefix = sub("\\[.*\\]", "", parameter))  # removes indices like [1], [2]

# Summarize: count and show first few examples per prefix
param_summary <- param_df %>%
  group_by(prefix) %>%
  summarise(
    count = n(),
    examples = paste(head(parameter, 5), collapse = ", ")
  ) %>%
  arrange(desc(count))

param_summary

```

## Definitions of typical parameters from mrmr/cmdstanr fit object

```{r}
library(tibble)
library(knitr)

# Create the table
param_table <- tibble(
  Prefix = c("s", "logit_detect", "eps_lambda", "eps_phi", "lambda", "B",
             "N", "beta_phi", "Nsuper", "alpha_lambda", "beta_detect",
             "lp__", "sigma_lambda", "sigma_phi"),
  Description = c(
    "Latent alive/dead state",
    "Linear predictor for detection probability (logit scale)",
    "Random effect for recruitment rate (lambda)",
    "Random effect for survival probability (phi)",
    "Recruitment or introduction rate",
    "Regression coefficients for recruitment covariates",
    "Estimated population size",
    "Regression coefficients for survival (logit scale)",
    "Superpopulation size (augmented individuals)",
    "Baseline recruitment rate (hierarchical mean)",
    "Regression coefficients for detection probability (logit scale)",
    "Log posterior density (Stan internal)",
    "SD of recruitment random effects",
    "SD of survival random effects"
  )
)

# Display nicely
knitr::kable(param_table, caption = "Summary of Posterior Parameter Prefixes")

```

## Survival probabilities by treatment group

### Extract posterior draws for survival coefficients

```{r}
draws_array <- model_22008$m_fit$draws()  # 3D array: iterations × chains × parameters

post_df <- as_draws_df(draws_array)

beta_phi_post <- post_df %>% 
  select(starts_with("beta_phi"))

colnames(beta_phi_post)
```

### Identify which coefficient corresponds to which treatment

```{r}
# colnames(model.matrix(~ treatment, data = data_22008$survival_covariate_df))

```

-   Column names are "Intercept" and "treatmentexposed". That indicates the following:
    -   beta_phi\[1\] = intercept = survival for the baseline group (unexposed)
    -   beta_phi\[2\] = effect of treatment = exposed relative to the baseline
-   Model formula incidates the same, without having to read from data_22008.

### Compute survival probabilities

```{r}
# For convenience
beta1 <- beta_phi_post$`beta_phi[1]`   # Intercept (unexposed reference level)
beta2 <- beta_phi_post$`beta_phi[2]`   # Effect of exposed

# Survival probabilities
phi_unexposed <- plogis(beta1)
phi_exposed <- plogis(beta1 + beta2)

```

-   beta_phi parameters on logit scale (= log-odds scale), plogis converts log-odds into survival probabilities
-   beta_phi\[1\] = logit(survival of unexposed, the baseline)
-   beta_phi\[2\] = log-odds difference for "exposed" relative to unexposed: beta_phi\[1\] + beta_phi\[2\]

### Put results in tidy dataframe

```{r}
surv_probs <- tibble(
  exposed = phi_exposed,
  unexposed = phi_unexposed
) %>%
  pivot_longer(everything(), names_to = "treatment", values_to = "phi")

head(surv_probs)

```

### Plot posterior survival distributions

```{r}
ggplot(surv_probs, aes(x = phi, fill = treatment)) +
  geom_density(alpha = 0.5) +
  labs(
    x = "Survival probability",
    y = "Posterior density",
    title = "Posterior Distribution of Survival by Treatment"
  ) +
  theme_minimal()

```

### Posterior difference (exposed - exposed)

```{r}
diff_exposed_unexposed <- phi_exposed - phi_unexposed

mean(diff_exposed_unexposed)
quantile(diff_exposed_unexposed, c(0.025, 0.5, 0.975))

```

-   On average, across the posterior, exposed frogs have \~5.7% lower survival probability than exposed frogs.
-   Interpretation of Bayesian credible interval: Given the model and data, there is a 95% probability that the true difference in survival (exposed – unexposed) lies between –18.6% and +4.4%.
-   The most likely difference is a small negative effect (exposed survive slightly worse). However, the credible interval overlaps zero, so the data do not strongly support either a real decrease in survival from exposure, or a real increase.
